---
title: "Obtaining, manipulating, and fitting climate data"
subtitle: "Climate data in infectious disease modelling · malaria in Thailand case study — Session 2"
format:
  training-html: default
presenter: "Michael & Roslyn"
contributors:
  - "Dr Vassili Kitsios"
  - "Prof. Roslyn Hickson"
  - "Dr Michael Lydeamore"
  - "Claude (Anthropic) — assisted with the initial translation of the existing Python workflow to R Quarto; all content to be checked by the course team"
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE
)

# Path to the raw data. This points at the existing Python workflow's data
# folder so nothing needs to be copied. Change this one line if the data moves.
data_dir <- "../data"
```

```{css, echo=FALSE}
.speaker-note {
    background-color: #fff8e1;
    border-left: 4px solid #f9a825;
    padding: 10px 14px;
    margin: 14px 0;
    border-radius: 3px;
    font-size: 0.92em;
    color: #5d4037;
}
.speaker-note::before {
    content: "🗒 Facilitator note: ";
    font-weight: bold;
}
.todo-note {
    background-color: #fdecea;
    border-left: 4px solid #d32f2f;
    padding: 10px 14px;
    margin: 14px 0;
    border-radius: 3px;
    font-size: 0.92em;
    color: #611a15;
}
.todo-note::before {
    content: "🚧 TODO (course team): ";
    font-weight: bold;
}
```

::: speaker-note
**Status of this document.** This is an initial, direct R translation of the
expert's Python analysis (`python_workflow/malaria_workflow.ipynb`). The analysis
steps have been kept faithful to the original — the same data sources, the same
processing choices, and the same models — with interactive discussion questions,
speaker notes, and solutions layered on top for teaching. Sections that the
course team plan to expand (download instructions, and an expanded statistical
climate-to-incidence fit with projections) are marked with **TODO** boxes.

**Environment.** The R practicals assume the following packages are installed:
`jsonlite`, `dplyr`, `tidyr`, `lubridate`, `ggplot2`, `patchwork`, `sf`,
`terra`, and `ncdf4`. Please confirm these are installed before the session. The
geospatial steps (reanalysis and CMIP6) read netCDF and shapefiles directly and
can take a little while on first render — `freeze: auto` means they are cached
after the first successful render.

**Worked example.** Throughout, the running example is malaria surveillance in
Thailand, linked to ERA5 reanalysis climate and CMIP6 future projections.
:::

# Summary {.tabset .tabset-pills .tabset-fade}

This session works through the practical workflow of using climate data in
infectious disease modelling, with malaria in Thailand as the case study: from
obtaining raw data through to a fitted statistical model that links climate to
malaria incidence, with future projections.

The session opens by framing the **research question** — always the first step of
any modelling project — and then works through six practical parts:

**The research question.** What we are trying to answer, and how it decides what
data and modelling we need.

**Part 1 — Obtaining the data.** Where the data come from (malaria
surveillance, climate reanalysis, future projections, administrative boundaries)
and how to obtain them.

**Part 2 — Malaria case data.** Reading the surveillance line list and summarising
cases by month, province, and district.

**Part 3 — Environmental variables (reanalysis).** Linking gridded ERA5 climate
data to administrative units using a shapefile, and building per-province monthly
temperature and rainfall series.

**Part 4 — Future climate projections.** Summarising CMIP6 projections under two
emissions scenarios (SSP1-2.6 and SSP5-8.5) for the same provinces.

**Part 5 — Fitting climate to incidence.** A statistical (ARIMA/ARX panel) model
relating standardised climate to changes in malaria cases across provinces.

**Part 6 — Projecting future cases.** Pushing the fitted statistical model forward
under the CMIP6 scenarios to project future cases, with the important caveats of
long-range extrapolation.

**Optional — a mechanistic SEIRS alternative.** A more advanced, biologically
structured transmission model, tucked behind an expandable box at the end for anyone
who gets through the six parts with time to spare.

::: {.callout-tip}
We recommend saving your written answers at the end of each part by printing the
page (there is a **Print this page** button at the end of each part).
:::

---

# The research question

::: speaker-note
Start here, not with data or code. The single most common mistake in applied
modelling is reaching for a dataset (or a method) before the question is clear.
Spend the first few minutes establishing *what we are trying to answer* and *what
decision it would inform*, and let that drive everything downstream. Draw out that
the question is what tells us (a) whether climate data is even relevant, (b) which
data we need, and (c) whether it makes sense to project forward. Participants do
this same first step for their own work in Session 3.
:::

Every modelling project begins with a clear question — not with a dataset and not
with a method. The question determines what data we need, which modelling approach
is appropriate, and how far we can push the results.

For this case study, our research question is:

> **How do temperature and rainfall relate to malaria incidence across the
> provinces of Thailand, and how might malaria risk change under future emissions
> scenarios?**

Reading that question tells us what we must assemble, before we touch any data:

- a measure of the **outcome** — malaria cases, over time and by location;
- the **drivers** we think matter — temperature and rainfall;
- a **spatial unit** to tie them together — provinces (the unit cases are reported
  in);
- a **historical period** where we observe outcome and drivers together, to *learn*
  the relationship; and
- **future climate** for the same region, to *apply* that relationship forward.

That list is exactly the set of data sources we obtain in Part 1. Framing the
question first also exposes the *ideal versus achievable* tension — the most
biologically faithful approach may need data we cannot get — which we take up as we
obtain the data.

**Your turn: think of a climate-sensitive health (or other) outcome in your own
context. Write it as a one-sentence research question. What decision would a good
answer inform?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**Your turn: for your question, list the pieces you would need to answer it — the
outcome, the drivers, the spatial/temporal unit, and the historical data needed to
link them.**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Notes for these prompts — click to reveal (try them first)

There is no single right answer; the point is the discipline of writing the
question *before* choosing data or methods. A good research question is specific
about the outcome, the location/time, and the decision it informs (e.g. "where and
when will dengue risk rise enough to warrant expanded vector control by 2050?"
rather than "is climate bad for health?"). The list of needed pieces should always
include the outcome measure, the candidate drivers, the unit that links them, and a
period where both are observed together — and, if the question looks forward, a
credible way to represent the future.
:::

---

# **Part 1:** Obtaining the data

::: speaker-note
This part is deliberately light on code — it is about *where the data come from*
and *why each source was chosen*. The original workflow notes are reproduced
below. Spend a few minutes on the mosquito-data caveat (there is no recent
mosquito data, which is why climate is used as a proxy for vector dynamics later),
as it motivates the modelling choices in Parts 4 and 5.
:::

The workflow draws on four data sources. Three are used directly in the analysis;
the fourth (mosquito occurrence) motivates the approach but is not currently used.

**1. Malaria case data (surveillance line list).** Individual records of malaria
blood draws in Thailand, with date, province, district, and result (*P. vivax*
`V`, *P. falciparum* `F`, or other). Obtained from the Thai DDC malaria API:
[https://malaria-api.ddc-malaria.org/study/2year/calendar](https://malaria-api.ddc-malaria.org/study/2year/calendar){target="_blank"}.
Saved as `data/malaria/calendar.json`.

**2. Administrative boundaries (shapefile).** Province (ADM1) boundaries for
Thailand. Shapefiles contain polygon descriptions of boundaries (countries,
provinces, districts) and are used to apply operations — such as a spatial average
— to gridded data over only the region of interest. Downloaded from
[Open Development Mekong](https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries){target="_blank"}.
Saved under `data/shape_files/tha_admbnda_adm1_rtsd_20190221/`.

**3. Climate reanalysis (ERA5).** Gridded, monthly-averaged 2 m temperature (`2t`)
and total precipitation (`tp`) — a physically consistent reconstruction of past
weather used here to *learn* the climate–malaria relationship. The required data
is already curated for you; additional variables can be downloaded freely from the
Copernicus Climate Change Service (C3S) Climate Data Store:
[ERA5 monthly single levels](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download){target="_blank"}.
Saved under `data/reanalysis/monthly-averaged/`.

**4. Future climate projections (CMIP6).** A multi-model ensemble of monthly
temperature (`tas`) and precipitation (`pr`) for Thailand under two emissions
scenarios, SSP1-2.6 and SSP5-8.5 — used to *apply* the learned relationship into
the future. The data is curated and trimmed to the region for this tutorial;
additional variables, scenarios, or the full globe can be downloaded freely from
the same Climate Data Store:
[CMIP6 projections](https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download){target="_blank"}.
Saved under `data/cmip6/monthly-averaged/`.

**On mosquito data (not currently used).** Mosquito occurrence records are
available from GBIF
([dataset db4211d6](https://www.gbif.org/dataset/db4211d6-b480-4bc5-9202-87f2072c4b12){target="_blank"}),
but there are no samples after 2022, so they do not overlap the 2025–2026 malaria
data. For now, rather than modelling mosquito populations explicitly, we relate
malaria to environmental variables directly (the approach taken in Part 5). A
natural extension is to obtain overlapping mosquito data, learn the influence of
temperature on the mosquito population, then the impact of that population on
malaria infections.

::: {.callout-note}
## Obtaining the data yourself
The Copernicus Climate Data Store (ERA5 and CMIP6) requires a free account and,
for bulk downloads, the CDS API (a `~/.cdsapirc` key file). Because these
downloads need accounts/API keys and can be large, they are not run as code here —
the curated subset needed for this tutorial is provided separately in the `data/`
folder at the repository root.
:::

## Ideal vs achievable workflows

::: speaker-note
This is a key discussion point to raise explicitly with participants: **what you
*want* to do is constrained by the data you can actually get.** When planning this
example we had hoped to match local mosquito vector abundance to climate, then feed
mosquito population estimates directly into a transmission model — the most
biologically defensible path. No overlapping abundance data existed, so we fell
back to a simpler, biologically less correct SEIRS model in which climate forces
transmission directly (the optional section at the end of this session), and a
direct statistical model of incidence (Part 5). The general principle worth landing:
prefer to model the *known*
climate-sensitive mechanism — the mosquito — explicitly, **wherever you have the
data to do it**; only drop to simpler models as data availability forces you to.
:::

The right workflow is rarely the most sophisticated one — it is the most
mechanistic model your data can actually support. For malaria, the biologically
known climate-sensitive link is the *mosquito*: temperature and rainfall drive
mosquito abundance, biting, and survival, which in turn drive transmission. That
suggests three tiers of approach:

| Approach | Pipeline | What it needs | Biological fidelity |
|---|---|---|---|
| **Ideal** (vector-explicit) | climate → mosquito abundance → transmission model → incidence | overlapping local mosquito **abundance** data *and* climate | highest |
| **Achievable here** (climate-forced SEIRS) | climate → transmission (β forced by climate) → incidence | case data + climate | medium — climate stands in for the missing vector dynamics |
| **Simplest** (direct statistical) | climate → incidence | case data + climate | lowest — no explicit mechanism |

For this example we wanted the **ideal** path, but no mosquito abundance data
overlapped the malaria period (see the GBIF note above), so we use the lower two
tiers: a direct statistical model (Part 5) and a climate-forced SEIRS model (the
optional section at the end of this session). Both skip the mosquito layer and let
climate stand in for it. This is
a deliberate, data-driven compromise — not the biologically preferred design — and
it is worth being honest about that trade-off when presenting results.

**1.1 Look at the four data sources above. For your own context, which of these
would be the hardest to obtain, and why? Is there an equivalent malaria (or other
health) surveillance dataset you could access?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**1.2 Reanalysis is used to *learn* the climate–disease relationship and CMIP6
projections are used to *apply* it into the future. Why can we not simply use the
future projections to learn the relationship in the first place?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 1 — click to reveal (try the questions first)

**1.1** No single answer. The malaria surveillance data is often the hardest to
obtain in practice, because health data is sensitive, may require data-sharing
agreements, and is frequently not openly available at fine spatial/temporal
resolution. Climate reanalysis (ERA5) and CMIP6 are, by contrast, openly available
(they just require an account and some bandwidth). The point is to get
participants thinking about the data landscape in their own setting.

**1.2** Future projections are, by construction, *projections* — they describe
plausible futures, not observed weather, and there is no malaria data from the
future to pair them with. We can only learn a relationship where we observe *both*
climate and cases together, which means the historical/reanalysis period. The
projections are then used to *apply* that learned relationship forward.
:::

---

# **Part 2:** Malaria case data

::: speaker-note
This part reads the surveillance line list and aggregates it. The key teaching
points are: (1) surveillance data are messy and unevenly reported across
provinces and time — this is why we later restrict to provinces with enough data;
and (2) aggregating to monthly counts per province gives us the response variable
that we will link to climate. The code is a direct translation of the Python
`pandas` workflow. Have participants run it and *look at* the province panels
(question 2.2) before moving on.
:::

We start from the surveillance line list: one row per blood draw, with a date,
province, district, and result. We read it, parse the dates, and aggregate to
monthly counts.

**Run the code below, then work through the questions underneath it.**

```{r part2-read, echo=TRUE, eval=TRUE}
library(jsonlite)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)

# Read the JSON line list and parse dates ---------------------------------
df <- fromJSON(file.path(data_dir, "malaria", "calendar.json"))
df <- df %>%
  mutate(
    date_daily = as.Date(as.character(blood_draw_date), format = "%Y%m%d"),
    month      = floor_date(date_daily, "month")
  )

cat("Number of unique province names:", n_distinct(df$province_en), "\n")
cat("Number of unique district names:", n_distinct(df$district_en), "\n")
cat("Number of unique blood_draw_date:", n_distinct(df$blood_draw_date), "\n")
```

```{r part2-counts, echo=TRUE, eval=TRUE, fig.width=10, fig.height=6}
# Total counts per day and per month --------------------------------------
counts_per_day <- df %>% count(date_daily, name = "count") %>% arrange(date_daily)
counts_per_month <- df %>% count(month, name = "count") %>% arrange(month)

p_day <- ggplot(counts_per_day, aes(date_daily, count)) +
  geom_line() +
  labs(x = "date", y = "total daily counts") +
  theme_bw()

p_month <- ggplot(counts_per_month, aes(month, count)) +
  geom_line() + geom_point(size = 1) +
  labs(x = "date (monthly total, aligned to first of month)",
       y = "total monthly counts") +
  theme_bw()

p_day / p_month
```

```{r part2-province, echo=TRUE, eval=TRUE, fig.width=11, fig.height=11}
# Counts per province per month -------------------------------------------
counts_prov_month <- df %>%
  count(month, province_en, name = "Count") %>%
  arrange(month, province_en)

# Keep provinces reported in more than 12 months (enough for a seasonal signal),
# matching the original workflow's selection.
prov_order <- unique(counts_prov_month$province_en)
selected_provinces <- prov_order[
  vapply(prov_order,
         function(p) sum(counts_prov_month$province_en == p) > 12,
         logical(1))
]
cat("Selected provinces:", length(selected_provinces), "\n")

counts_prov_month %>%
  filter(province_en %in% selected_provinces) %>%
  ggplot(aes(month, Count)) +
  geom_line() + geom_point(size = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  labs(x = "date", y = "monthly count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r part2-district, echo=TRUE, eval=TRUE}
# Counts per district per month (finer spatial scale) ---------------------
counts_dist_month <- df %>%
  count(month, district_en, name = "Count")
head(counts_dist_month)
```

**2.1 How many provinces are reported in the surveillance data, and how many are
kept after the "more than 12 months" filter? Why might we want to restrict to
provinces with a longer record before trying to relate cases to climate?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**2.2 Look at the per-province panels. Describe the seasonal pattern you see. Do
all provinces peak at the same time of year? What might explain differences
between provinces?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**2.3 The results column pools *P. vivax* and *P. falciparum* together. Why might
it matter, for a climate-driven analysis, that these two species are combined?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 2 — click to reveal (try the questions first)

**2.1** There are many provinces in the raw data, but only those reported in more
than 12 months are kept (a couple of dozen), and of those only the ones whose names
match the boundary shapefile are carried forward. Restricting to provinces with a
longer record means each province has enough months to show a seasonal signal and to
standardise sensibly; provinces with only a handful of reported months cannot support
a climate–seasonality analysis.

**2.2** Most provinces show a seasonal rise and fall, but not all peak in the same
month — malaria seasonality depends on local climate, elevation, land use, and
the timing of the rains, which differ across the country. Border provinces and
forested provinces often behave differently from central ones.

**2.3** *P. vivax* and *P. falciparum* differ in incubation period, in relapse
behaviour (vivax can relapse from liver-stage hypnozoites months later), and
likely in their climate sensitivity. Pooling them means the fitted
climate–malaria relationship is an average over two biologically different
processes, which can blur or bias the signal. A species-stratified analysis is a
natural next step.
:::

---

# **Part 3:** Environmental variables (reanalysis)

::: speaker-note
This part links gridded climate to the administrative units that malaria is
reported in. The conceptual step to draw out is the *spatial mismatch*: climate
comes on a regular latitude–longitude grid, but cases come per province. We
resolve it in two steps that reproduce the original Python exactly. First,
**strict cell-centre assignment**: a grid cell belongs to the province whose polygon
contains the cell's *centre* (a point-in-polygon test, matching the `regionmask`
package). Second, the province's cells are averaged with the notebook's chained
`.mean('lat').mean('lon')` — average over latitude *within each longitude column*,
then over longitude. That second step is a subtle choice: it differs from a flat
mean over all the province's cells for irregularly shaped provinces (see the TODO
below — we need Vassili's rationale so we can justify it to participants).
Participants often ask why we keep temperature in Kelvin during standardisation:
because a constant offset (Kelvin vs Celsius) does not change a standardised
(z-scored) series, so it only matters for the plots, where we convert to °C.
:::

Malaria is reported per province, but ERA5 climate comes on a regular grid. To
link them, we summarise the grid to provinces in two steps that match the original
analysis: **strict cell-centre assignment** — each grid cell is assigned to the
province whose polygon contains the cell's centre (a point-in-polygon test with
`sf`, matching the [`regionmask`](https://regionmask.readthedocs.io){target="_blank"}
package) — and then a **chained latitude-then-longitude average** of those cells,
reproducing the notebook's `.mean('lat').mean('lon')` (see the note below).

::: {.todo-note}
The provincial climate is averaged with the original workflow's chained
`.mean('lat').mean('lon')`: average over latitude within each longitude column, then
over longitude. For an irregularly shaped province this differs from a flat mean
over all its cells (e.g. Uthai Thani: 295.08 K vs 294.64 K). **Ask Vassili** why
this averaging was chosen — whether it is deliberate (some weighting rationale) or an
xarray convention — so we can explain and justify it to participants, and decide
whether to keep it or move to a flat or area-weighted (cos-latitude) spatial mean.
R reproduces the notebook's choice exactly so the two implementations agree.
:::

**Run the code below, then work through the questions.**

```{r part3-shape, echo=TRUE, eval=TRUE}
library(sf)
library(terra)

# Province (ADM1) boundaries ----------------------------------------------
shape_file <- st_read(
  file.path(data_dir, "shape_files",
            "tha_admbnda_adm1_rtsd_20190221",
            "tha_admbnda_adm1_rtsd_20190221.shp"),
  quiet = TRUE
)
region_names <- shape_file$ADM1_EN
cat("Number of provinces in shapefile:", length(region_names), "\n")

# The surveillance province names must match the shapefile's ADM1_EN names to be
# mapped. Keep only the selected provinces that are present in the boundaries
# (mirrors the original workflow, which reduces the province list to those found
# in the shapefile). Report any that are dropped so name mismatches are visible.
missing <- setdiff(selected_provinces, region_names)
if (length(missing))
  message("Provinces not matched in shapefile and dropped: ",
          paste(missing, collapse = ", "))
selected_provinces <- selected_provinces[selected_provinces %in% region_names]
cat("Provinces carried forward (matched to boundaries):",
    length(selected_provinces), "\n")
```

```{r part3-reanalysis, echo=TRUE, eval=TRUE}
# ERA5 monthly reanalysis: 2 m temperature (2t -> t2m) and precip (tp) -----
# One netCDF file per month; sorting the file paths puts them in time order.
tas_files <- sort(list.files(
  file.path(data_dir, "reanalysis", "monthly-averaged", "2t"),
  pattern = "\\.nc$", recursive = TRUE, full.names = TRUE))
pr_files <- sort(list.files(
  file.path(data_dir, "reanalysis", "monthly-averaged", "tp"),
  pattern = "\\.nc$", recursive = TRUE, full.names = TRUE))

tas_r <- rast(tas_files)   # temperature, one layer per month (Kelvin)
pr_r  <- rast(pr_files)    # total precipitation, one layer per month (metres)

# Restrict to the Thailand region used in the original workflow
region_ext <- ext(97, 108, 5, 22)
tas_r <- crop(tas_r, region_ext)
pr_r  <- crop(pr_r,  region_ext)

# Recover the month for each layer from the filename (YYYYMMDD-...)
month_from_file <- function(f)
  floor_date(as.Date(sub(".*_(\\d{8})-\\d{8}\\.nc$", "\\1", basename(f)),
                     format = "%Y%m%d"), "month")
clim_months <- month_from_file(tas_files)
cat("Reanalysis covers", length(clim_months), "months:",
    format(min(clim_months)), "to", format(max(clim_months)), "\n")
```

```{r part3-maps, echo=TRUE, eval=TRUE, fig.width=11, fig.height=5}
# Map of the time-mean temperature and rainfall over the region -----------
tas_mean_c <- mean(tas_r) - 273.15          # Kelvin -> Celsius for display
pr_mean    <- mean(pr_r)

map_df_t <- as.data.frame(tas_mean_c, xy = TRUE); names(map_df_t)[3] <- "value"
map_df_r <- as.data.frame(pr_mean,    xy = TRUE); names(map_df_r)[3] <- "value"

m_t <- ggplot(map_df_t, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey30",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_c(name = "°C") +
  coord_sf(expand = FALSE) +
  labs(title = "Mean temperature", x = "lon", y = "lat") +
  theme_bw()

m_r <- ggplot(map_df_r, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey30",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_c(name = "m", option = "mako") +
  coord_sf(expand = FALSE) +
  labs(title = "Mean rainfall", x = "lon", y = "lat") +
  theme_bw()

m_t + m_r
```

```{r part3-extract, echo=TRUE, eval=TRUE}
# Assign each grid cell to a province by STRICT CELL-CENTRE assignment --------
# A cell belongs to the province whose polygon contains the cell's *centre*, and
# the provincial value is the simple mean over those cells. This reproduces the
# standard Python geospatial workflow used in the original analysis (the
# `regionmask` package), rather than relying on the raster library's own polygon
# extraction, whose boundary handling can differ slightly.
grid_xy  <- terra::xyFromCell(tas_r, seq_len(terra::ncell(tas_r)))
grid_pts <- sf::st_as_sf(as.data.frame(grid_xy), coords = c("x", "y"),
                         crs = sf::st_crs(shape_file))
within   <- sf::st_within(grid_pts, shape_file)          # cell centre in polygon?
cell_region <- vapply(within,
                      function(h) if (length(h)) h[1] else NA_integer_,
                      integer(1))

# Provincial mean for every layer (time step), reproducing the original workflow's
# chained `ds.where(mask==k).mean('lat').mean('lon')`: average over latitude *within
# each longitude column* first, then average those column-means over longitude. This
# differs from a flat mean over all cells for provinces whose cells are spread
# unevenly across columns, so we match the notebook exactly rather than approximate.
# TODO (course team): ask Vassili why this chained average was used rather than a
# flat or area-weighted (cos-latitude) spatial mean, so we can justify it to students.
# The grid is shared by the reanalysis and (after resampling) the CMIP data, so the
# same cell assignment and longitudes are reused in Part 4.
region_means <- function(r) {
  vals <- terra::values(r)                                # ncell x nlayer
  lon  <- grid_xy[, 1]                                    # per-cell longitude
  out  <- matrix(NA_real_, length(region_names), ncol(vals),
                 dimnames = list(region_names, NULL))
  for (k in seq_along(region_names)) {
    sel <- which(cell_region == k)
    if (!length(sel)) next
    cols <- split(sel, lon[sel])                          # province cells grouped by longitude
    col_means <- vapply(cols,                             # mean over latitude within each column
      function(ix) colMeans(vals[ix, , drop = FALSE], na.rm = TRUE),
      numeric(ncol(vals)))                                # nlayer x n_columns
    out[k, ] <- rowMeans(col_means, na.rm = TRUE)         # then mean over longitude columns
  }
  out
}

tas_prov <- region_means(tas_r)
pr_prov  <- region_means(pr_r)

# Tidy into long tables, keeping only the selected provinces
to_long <- function(mat, value_name) {
  m <- as.matrix(mat)
  rownames(m) <- region_names
  colnames(m) <- as.character(clim_months)
  as.data.frame(m) %>%
    tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = value_name) %>%
    mutate(month = as.Date(month))
}

clim_long <- to_long(tas_prov, "temp_k") %>%
  left_join(to_long(pr_prov, "rain_m"), by = c("province_en", "month")) %>%
  mutate(temp_c = temp_k - 273.15) %>%
  filter(province_en %in% selected_provinces)

head(clim_long)
```

```{r part3-series, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
# Per-province monthly temperature and rainfall series --------------------
ggplot(clim_long, aes(month, temp_c)) +
  geom_line() + geom_point(size = 0.5) +
  facet_wrap(~ province_en, scales = "free_y") +
  labs(x = "date", y = "monthly mean temperature (°C)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

**3.1 We averaged climate over each province. What information do we lose by
doing this? For a large province with varied terrain, is a single provincial
average a good summary of the climate that mosquitoes and people experience?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**3.2 Look at the per-province temperature series. How large is the within-year
temperature range compared to the differences between provinces? Which do you
think will be more useful for explaining malaria seasonality?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**3.3 The reanalysis only covers a couple of years here. What are the risks of
learning a climate–malaria relationship from such a short record?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 3 — click to reveal (try the questions first)

**3.1** Averaging to a province discards all within-province spatial variation —
elevation gradients, forest edges, urban vs rural. For a large or topographically
varied province, a single average may not represent the conditions where
transmission actually happens (often specific ecological niches), so the link to
cases is weakened.

**3.2** Typically the within-year (seasonal) temperature swing within a province is
modest compared with, say, rainfall seasonality, and provinces differ from each
other too. Both the seasonal cycle and the between-province contrasts carry
information; the panel model in Part 5 uses both by standardising within province
and pooling across provinces.

**3.3** A short record risks fitting to year-specific anomalies (an unusual season,
a one-off outbreak) rather than a stable climate–malaria relationship, and gives
very little power to separate climate effects from other drivers. Projections built
on such a fit should be treated as illustrative.
:::

---

# **Part 4:** Future climate projections (CMIP6)

::: speaker-note
This part summarises the CMIP6 multi-model ensemble to the same provinces, under
two scenarios. The key ideas for participants: (1) CMIP6 is an *ensemble* of many
climate models — we summarise across models with a mean and an interquartile band,
rather than trusting any single model; and (2) the two scenarios (SSP1-2.6 low
emissions, SSP5-8.5 high emissions) bracket a range of plausible futures. This is
a direct translation of the original Python, including the precipitation unit
conversion. It is the heaviest step in the document; on a slow machine give it a
minute, and note that `freeze: auto` caches it afterwards.
:::

The learned relationship (Part 5) is only useful if we can apply it to the future.
CMIP6 provides projections of temperature (`tas`) and precipitation (`pr`) from
many climate models, under different emissions scenarios. We summarise them to the
same provinces, under two scenarios: **SSP1-2.6** (low emissions) and **SSP5-8.5**
(high emissions).

**Run the code below, then work through the questions.**

```{r part4-read, echo=TRUE, eval=TRUE}
library(ncdf4)

cmip_dir <- file.path(data_dir, "cmip6", "monthly-averaged")
cmip_files <- list(
  ssp126_tas = "cmip6_regrided_model_ensemble.Thai_region.ssp126.tas.nc",
  ssp585_tas = "cmip6_regrided_model_ensemble.Thai_region.ssp585.tas.nc",
  ssp126_pr  = "cmip6_regrided_model_ensemble.Thai_region.ssp126.pr.nc",
  ssp585_pr  = "cmip6_regrided_model_ensemble.Thai_region.ssp585.pr.nc"
)

# Read one CMIP variable as a list of per-model SpatRasters (time layers),
# returning the models, times, and the raster grid. Dimensions are read by
# name so the code does not depend on storage order.
read_cmip_var <- function(path, var, from_date = NULL) {
  nc  <- nc_open(path)
  on.exit(nc_close(nc))
  lon <- ncvar_get(nc, "lon"); lat <- ncvar_get(nc, "lat")
  tim <- ncvar_get(nc, "time")
  models <- ncvar_get(nc, "model")
  # time units -> dates
  tunits <- ncatt_get(nc, "time", "units")$value
  origin <- as.Date(sub(".*since ", "", tunits))
  dates  <- floor_date(origin + tim, "month")
  arr <- ncvar_get(nc, var)                       # dims in file order
  dn  <- sapply(nc$var[[var]]$dim, function(d) d$name)
  # reorder to lon, lat, time, model
  perm <- match(c("lon", "lat", "time", "model"), dn)
  arr <- aperm(arr, perm)
  # Trim to the reanalysis start onwards (as in the original workflow), so the
  # historical fit and the future projection share the same start month.
  if (!is.null(from_date)) {
    keep <- dates >= from_date
    arr  <- arr[, , keep, , drop = FALSE]
    dates <- dates[keep]
  }
  list(arr = arr, lon = lon, lat = lat, dates = dates, models = models)
}

cmip_start <- min(clim_months)   # reanalysis start month
tas126 <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp126_tas), "tas", cmip_start)
cat("CMIP6 grid:", length(tas126$lon), "lon x", length(tas126$lat), "lat;",
    length(tas126$models), "models;", length(tas126$dates), "months\n")
```

```{r part4-summarise, echo=TRUE, eval=TRUE}
# Build a raster template on the reanalysis grid so CMIP can be interpolated
# onto it (matching the original workflow), then averaged per province.
ref_grid <- tas_r[[1]]        # reanalysis grid (already cropped to region)

# Interpolate a per-model, per-time CMIP array onto the reanalysis grid and
# average over each province. Returns a tidy data frame of provincial means.
# Build a north-up, multi-layer SpatRaster from one model's [lon, lat, time]
# slice. Handles either coordinate ordering and treats the lon/lat values as
# cell centres (so the extent is padded by half a cell each side).
slice_to_rast <- function(slc, lon, lat, ref) {
  lon_o <- order(lon); lat_o <- order(lat)          # index ascending
  slc <- slc[lon_o, lat_o, , drop = FALSE]          # lon asc, lat asc
  a <- aperm(slc, c(2, 1, 3))                        # -> lat x lon x time
  a <- a[dim(a)[1]:1, , , drop = FALSE]              # flip: north (max lat) = row 1
  r <- rast(a)                                       # 3rd dim -> layers
  lo <- sort(lon); la <- sort(lat)
  dx <- mean(diff(lo)); dy <- mean(diff(la))
  ext(r) <- c(min(lo) - dx/2, max(lo) + dx/2,
              min(la) - dy/2, max(la) + dy/2)
  crs(r) <- crs(ref)                                 # match the reanalysis grid
  r
}

# Interpolate every model's CMIP slice onto the reanalysis grid and average per
# province, KEEPING the per-model detail (province x month x model). We keep the
# per-model array because the Part 6 projection needs the model spread, not just
# the ensemble mean.
cmip_provincial_bymodel <- function(cv, var, ref, provinces) {
  nm <- length(cv$models)
  # Optional speed/debug knob: set env var CMIP_MODEL_CAP to a small integer to
  # use only the first few models (much faster). Unset / 0 uses all models.
  cap <- suppressWarnings(as.integer(Sys.getenv("CMIP_MODEL_CAP", "")))
  if (!is.na(cap) && cap > 0) nm <- min(nm, cap)
  out <- vector("list", nm)
  for (m in seq_len(nm)) {
    slc <- cv$arr[, , , m]                     # lon x lat x time
    r <- slice_to_rast(slc, cv$lon, cv$lat, ref)
    names(r) <- as.character(cv$dates)
    if (var == "pr") r <- r * 86400 / 1000     # flux (kg m-2 s-1) -> m (as in original)
    else             r <- r - 273.15           # Kelvin -> Celsius
    r <- resample(r, ref, method = "bilinear") # interpolate onto reanalysis grid
    pm <- region_means(r)                       # strict cell-centre assignment (as Part 3)
    out[[m]] <- pm[provinces, , drop = FALSE]
  }
  arr3 <- simplify2array(out)                  # province x month x model
  list(arr = arr3, dates = cv$dates, provinces = provinces)
}

# Ensemble summary (mean + interquartile band) from a per-model array.
summarise_cmip <- function(bm, scenario, var) {
  arr3 <- bm$arr; nt <- length(bm$dates)
  qf <- function(p) as.vector(apply(arr3, c(1, 2), quantile, probs = p, na.rm = TRUE))
  data.frame(
    province_en = rep(bm$provinces, times = nt),
    month       = rep(bm$dates, each = length(bm$provinces)),
    scenario = scenario, variable = var,
    mean = as.vector(apply(arr3, c(1, 2), mean, na.rm = TRUE)),
    q25 = qf(0.25), q75 = qf(0.75)
  )
}
```

::: {.todo-note}
The CMIP6 interpolation-and-averaging above is the single heaviest step and the
part most likely to need a check on first render (it reads a 45-model ensemble and
resamples it onto the reanalysis grid, exactly as the original Python did). If it
is too slow for a live session, the course team may wish to pre-compute the
provincial summaries once and cache them. It is left as a direct port here so the
method is transparent.
:::

```{r part4-run, echo=TRUE, eval=TRUE}
tas585 <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp585_tas), "tas", cmip_start)
pr126  <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp126_pr),  "pr",  cmip_start)
pr585  <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp585_pr),  "pr",  cmip_start)

# Restrict all four files to the CLIMATE MODELS they have in common, in the same
# order (as the original workflow does). The files hold different numbers of
# models, and the projection pairs temperature and rainfall model-by-model, so
# they must refer to the same models in the same order.
common_models <- sort(Reduce(intersect,
  list(tas126$models, tas585$models, pr126$models, pr585$models)))
align_models <- function(cv) {
  idx <- match(common_models, cv$models)
  cv$arr    <- cv$arr[, , , idx, drop = FALSE]
  cv$models <- common_models
  cv
}
tas126 <- align_models(tas126); tas585 <- align_models(tas585)
pr126  <- align_models(pr126);  pr585  <- align_models(pr585)
cat("CMIP models in common across all four files:", length(common_models), "\n")

# Per-model provincial climate (kept for the Part 6 projection)
cmip_tas <- list(
  "SSP1-2.6" = cmip_provincial_bymodel(tas126, "tas", ref_grid, selected_provinces),
  "SSP5-8.5" = cmip_provincial_bymodel(tas585, "tas", ref_grid, selected_provinces))
cmip_pr <- list(
  "SSP1-2.6" = cmip_provincial_bymodel(pr126, "pr", ref_grid, selected_provinces),
  "SSP5-8.5" = cmip_provincial_bymodel(pr585, "pr", ref_grid, selected_provinces))

# Ensemble summary for plotting
cmip_summary <- bind_rows(
  summarise_cmip(cmip_tas[["SSP1-2.6"]], "SSP1-2.6", "tas"),
  summarise_cmip(cmip_tas[["SSP5-8.5"]], "SSP5-8.5", "tas"),
  summarise_cmip(cmip_pr[["SSP1-2.6"]],  "SSP1-2.6", "pr"),
  summarise_cmip(cmip_pr[["SSP5-8.5"]],  "SSP5-8.5", "pr")
)
```

```{r part4-plot, echo=TRUE, eval=TRUE, fig.width=11, fig.height=8}
# Annual means for a clearer long-range view, for one example province -----
example_province <- selected_provinces[1]

annual <- cmip_summary %>%
  mutate(year = year(month)) %>%
  filter(province_en == example_province) %>%
  group_by(province_en, scenario, variable, year) %>%
  summarise(mean = mean(mean), q25 = mean(q25), q75 = mean(q75), .groups = "drop")

lab <- c(tas = "temperature (°C)", pr = "total precipitation (m)")

ggplot(annual, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_y",
             labeller = labeller(variable = lab)) +
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = NULL,
       title = paste0("CMIP6 projections for ", example_province),
       subtitle = "Model mean (line) and interquartile range (band)") +
  theme_bw()
```

**4.1 The projections are shown as a model mean with an interquartile band. Why do
we show a *range* across models rather than picking the single "best" model?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**4.2 Compare the two scenarios (SSP1-2.6 and SSP5-8.5). How far apart are the
temperature projections by the end of the century? What about rainfall — is the
model agreement as strong?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**4.3 We are about to feed these projections into a malaria model that was fitted
on only a couple of years of reanalysis. What assumption are we making about the
climate–malaria relationship when we extrapolate it decades into the future?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 4 — click to reveal (try the questions first)

**4.1** Individual climate models disagree, especially on rainfall. The
ensemble mean is more reliable than any single model, and the spread (here the
interquartile band) is itself the honest expression of projection *uncertainty*.
Picking one "best" model hides that uncertainty.

**4.2** Temperature projections usually diverge clearly between the two scenarios
by late century, with SSP5-8.5 substantially warmer, and models tend to agree on
the direction of warming. Rainfall is far less certain: the inter-model band is
usually wide and the scenarios may overlap, reflecting genuine disagreement about
how precipitation will change.

**4.3** We are assuming the climate–malaria relationship learned over a couple of
recent years holds, unchanged, decades into the future — i.e. that no adaptation,
intervention, land-use change, drug resistance, or shift in the mosquito/parasite
system alters it. That is a strong assumption and the main caveat on any such
projection.
:::

---

# **Part 5:** Fitting climate to incidence

::: speaker-note
This is the core of the session: a statistical model relating climate to malaria
cases across all provinces at once (a "panel" model). The design choices to draw
out: (1) we model the *first difference* (month-to-month change) in cases rather
than the level, and we *standardise* per province, so that provinces of very
different sizes contribute comparably; (2) climate enters as temperature, rainfall,
and their squares, allowing a hump-shaped (optimum) response; and (3) the fitted
coefficients are *shared* across provinces — one climate–malaria relationship
applied everywhere. This is a direct translation of the original "panel method"
ordinary-least-squares fit. Frame it explicitly as the *simplest* model — one set
of coefficients for the whole country. Part 6 then pushes it forward under CMIP6 to
project future cases; the optional SEIRS section at the end of this session is the
more scientifically sound alternative (per-province transmission) for those who want
a mechanistic model.
:::

We now link climate to cases directly with a simple statistical model, pooling all
provinces. We use a **panel method** — the same approach used to learn the impact
of temperature and rainfall on the gross domestic product of the world's economies
([Burke, Hsiang & Miguel, *Nature*, 2015](https://www.nature.com/articles/nature15725){target="_blank"})
— but here applied to learn the impact of temperature and rainfall on the
month-to-month *change* in infection count. Because these simple panel methods
cannot account for trends, we first remove the trend by taking the first difference,
then standardise each province by its own mean and standard deviation over time.
Temperature and rainfall enter as themselves and squared, allowing a non-linear
(hump-shaped) response, with one shared relationship across provinces. Because we
model the first difference (the "integrated" step) and reconstruct the level by a
cumulative sum, this is an **ARIMA-style** model — more precisely an autoregressive
model with exogenous inputs (ARX), with temperature and rainfall as the exogenous
climate drivers (hence `arx_fit` below).

This is perhaps the most basic model one can build, and is designed to illustrate
the logic of integrating the climate data. It uses a single set of coefficients for
all provinces. In Part 6 we push this fitted model forward under the CMIP6 climate
to project future cases. The optional SEIRS section at the end of this session is
more scientifically sound (it has per-province transmission) and gives you a code
base for fitting a transmission model directly.

**Run the code below, then work through the questions.**

```{r part5-matrices, echo=TRUE, eval=TRUE}
# Align cases and climate on the same province x month grid ----------------
provinces <- selected_provinces
months    <- sort(unique(clim_long$month))

# Case matrix: provinces x months, zero-filled where no cases were reported
cases_mat <- counts_prov_month %>%
  filter(province_en %in% provinces) %>%
  complete(province_en = provinces, month = months, fill = list(Count = 0)) %>%
  filter(month %in% months) %>%
  arrange(province_en, month) %>%
  pivot_wider(names_from = month, values_from = Count) %>%
  tibble::column_to_rownames("province_en") %>%
  as.matrix()
cases_mat <- cases_mat[provinces, , drop = FALSE]

# Climate matrices on the same grid
mat_from_long <- function(var) {
  clim_long %>%
    select(province_en, month, all_of(var)) %>%
    pivot_wider(names_from = month, values_from = all_of(var)) %>%
    tibble::column_to_rownames("province_en") %>%
    as.matrix() %>% `[`(provinces, , drop = FALSE)
}
clim_temp <- mat_from_long("temp_k")   # Kelvin (offset irrelevant once standardised)
clim_rain <- mat_from_long("rain_m")

cat("cases:", nrow(cases_mat), "provinces x", ncol(cases_mat), "months;",
    "total cases =", sum(cases_mat), "\n")
```

```{r part5-standardise, echo=TRUE, eval=TRUE}
# Standardise per province ------------------------------------------------
# Use the population standard deviation (divide by n, not n-1) to match NumPy's
# np.std in the original workflow. R's built-in sd() uses n-1, which would shift
# the standardisation and change the fitted coefficients.
popsd <- function(x) sqrt(mean((x - mean(x))^2))

# Response: standardised first difference of cases, per province
cases_fd <- cases_mat[, -1] - cases_mat[, -ncol(cases_mat)]
fd_mean  <- rowMeans(cases_fd)
fd_sd    <- apply(cases_fd, 1, popsd)
z_cases_fd <- (cases_fd - fd_mean) / fd_sd

# Predictors: standardised temperature and rainfall, per province
z_temp <- (clim_temp - rowMeans(clim_temp)) / apply(clim_temp, 1, popsd)
z_rain <- (clim_rain - rowMeans(clim_rain)) / apply(clim_rain, 1, popsd)
```

```{r part5-fit, echo=TRUE, eval=TRUE}
# Ordinary least squares on the pooled (flattened) panel ------------------
# Predictors are temperature, temperature^2, rainfall, rainfall^2, using the
# months aligned with the first differences (drop the first month).
Y <- as.vector(z_cases_fd)
zt <- z_temp[, -1]; zr <- z_rain[, -1]
X  <- cbind(as.vector(zt), as.vector(zt)^2, as.vector(zr), as.vector(zr)^2)

# Centre (as in the original), solve, and predict
Y_mean <- mean(Y);            Y_prime <- Y - Y_mean
X_mean <- colMeans(X);        X_prime <- sweep(X, 2, X_mean)
beta   <- solve(t(X_prime) %*% X_prime, t(X_prime) %*% Y_prime)
Y_hat  <- as.vector(X_prime %*% beta) + Y_mean

colnames(X) <- c("temp", "temp^2", "rain", "rain^2")
print(round(setNames(as.vector(beta), colnames(X)), 3))

# Reshape prediction back to province x month, and rebuild case counts by
# reversing the standardisation and the first-differencing (cumulative sum).
z_hat <- matrix(Y_hat, nrow = nrow(z_cases_fd))
fitted_cases <- cases_mat * 0
for (i in seq_len(nrow(cases_mat))) {
  fitted_cases[i, 1]  <- cases_mat[i, 1]
  fitted_cases[i, -1] <- fitted_cases[i, 1] +
    cumsum(z_hat[i, ] * fd_sd[i] + fd_mean[i])
}
```

```{r verify-export, echo=FALSE, message=FALSE}
# Save this fit's actual outputs to the shared outputs/ folder, so they can be
# compared directly against the Python notebook's saved outputs (see
# python_workflow/compare.R). Instrumentation only — it exports the arrays already
# computed above, it does not recompute anything.
.odir <- file.path("..", "outputs")
dir.create(.odir, showWarnings = FALSE, recursive = TRUE)
jsonlite::write_json(list(
  selected   = provinces,
  months     = as.character(months),
  cases_mat  = cases_mat,
  clim_temp  = clim_temp,
  clim_rain  = clim_rain,
  panel_beta = as.vector(beta),
  arx_fit    = fitted_cases
), file.path(.odir, "r_panel.json"), matrix = "rowmajor", digits = NA)
```

```{r part5-diag, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
# Observed vs fitted monthly cases, per province --------------------------
fit_long <- bind_rows(
  as.data.frame(cases_mat) %>% tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>%
    mutate(series = "observed"),
  as.data.frame(fitted_cases) %>% tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>%
    mutate(series = "fitted")
) %>% mutate(month = as.Date(month))

ggplot(fit_long, aes(month, cases, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c(observed = "black", fitted = "red")) +
  labs(x = "date", y = "monthly cases",
       title = "Observed vs panel-fitted monthly cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

The fitted `beta` coefficients define the climate response. In **Part 6** we apply
them to the future CMIP6 climate to project cases forward.

**5.1 The model uses temperature, temperature², rainfall, and rainfall². What
shape of climate–malaria relationship do the squared terms allow that a
straight-line model could not? Why is that biologically sensible for a
mosquito-borne disease?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**5.2 We standardised cases and climate *per province* before pooling. What would
go wrong if we pooled the raw counts across provinces without standardising?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**5.3 Look at the observed-vs-fitted panels. Where does the model do well, and
where does it do poorly? What might it be missing (think about what drives malaria
besides climate)?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**5.4 To project future cases, we would apply these fitted coefficients to the
CMIP6 climate from Part 4. List one thing you would want to check about the future
climate values before trusting such a projection.**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 5 — click to reveal (try the questions first)

**5.1** The squared terms allow a *hump-shaped* (unimodal) response: cases rising
with temperature or rainfall up to an optimum and then falling. This is
biologically sensible because mosquito survival, biting, and parasite development
each have thermal optima, and both drought and flushing rains can reduce breeding —
so the response to climate is rarely monotonic.

**5.2** Provinces differ enormously in population and case numbers. Pooling raw
counts would let a few large-caseload provinces dominate the fit, and would confound
"more cases because bigger" with "more cases because of climate". Standardising per
province puts every province on a comparable footing so the shared coefficients
reflect the climate relationship, not province size.

**5.3** The model tends to capture the broad seasonal rise and fall but misses
sharp peaks, sudden outbreaks, and province-specific idiosyncrasies. Malaria is
also driven by non-climate factors — interventions (bed nets, spraying), human
movement and importation, drug resistance, health-system reporting changes — none
of which are in a climate-only model.

**5.4** Sensible checks include: are the future climate values within the range the
model was trained on (extrapolation beyond the training range is dangerous)? Are the
units and standardisation applied identically to the projections? Is the seasonal
structure of the projections realistic? Does the province averaging match?
:::

---

# **Part 6:** Projecting future cases (statistical model)

::: speaker-note
This is the pay-off: we take the model fitted on the historical period (Part 5) and
push it forward under the CMIP6 climate (Part 4) to project future cases. The
mechanics: standardise each future climate value using the *historical*
per-province mean and standard deviation (never re-standardise on the future data,
or you throw away the trend you are trying to project), apply the fitted `beta`, and
rebuild case counts. We run every climate model separately and show the ensemble
mean and interquartile band, exactly as we did for the raw climate in Part 4.

Be sure to make the key caveat explicit (see the warning box): reconstructing counts
by cumulatively summing predicted month-to-month changes over ~75 years behaves like
a random walk and will drift. This is a genuine limitation of projecting a
differenced/ARX model far ahead, and a large part of *why* one might prefer a bounded
mechanistic model (the optional SEIRS section at the end of this session) for long
horizons. Treat the projection as illustrative of method, not as a forecast.
:::

We now push the fitted model forward. For each climate model and scenario, we
standardise the future CMIP6 temperature and rainfall using the **historical**
per-province mean and standard deviation (the same scaling the model was fitted
with), apply the fitted coefficients to get the predicted month-to-month change in
cases, and rebuild counts by cumulative sum — seeding from the last observed month.

::: {.callout-warning}
## A projection, not a forecast
Rebuilding case counts by cumulatively summing predicted monthly *changes* over
decades behaves like a random walk and will drift, so the absolute numbers far into
the future should not be read literally. This is an inherent limitation of pushing a
first-difference (ARX) model a long way forward, and is one reason a bounded
mechanistic model (the optional SEIRS section at the end of this session) is often
preferred for long-range projection. Use this to see *how* climate projections feed
a fitted model, not as a prediction of future case counts.
:::

```{r part6-project, echo=TRUE, eval=TRUE}
# Historical per-province climate scaling, in the SAME units as the CMIP arrays
# (temperature in °C, rainfall in m). Standardising is unaffected by a constant
# offset, so these reproduce the scaling used to fit the model in Part 5.
temp_mean_p <- rowMeans(clim_temp) - 273.15   # clim_temp is Kelvin; CMIP tas is °C
temp_sd_p   <- apply(clim_temp, 1, popsd)
rain_mean_p <- rowMeans(clim_rain)
rain_sd_p   <- apply(clim_rain, 1, popsd)

last_obs <- cases_mat[, ncol(cases_mat)]      # seed the projection at the last observed month
hist_end <- max(months)

# Apply the fitted ARX model to one climate model's future climate, rebuild counts.
project_one <- function(tas_pm, pr_pm, dates) {
  zt <- (tas_pm - temp_mean_p) / temp_sd_p    # province x month (vectors recycle per row)
  zr <- (pr_pm  - rain_mean_p) / rain_sd_p
  yhat <- (zt   - X_mean[1]) * beta[1] +      # predicted standardised first-difference
          (zt^2 - X_mean[2]) * beta[2] +
          (zr   - X_mean[3]) * beta[3] +
          (zr^2 - X_mean[4]) * beta[4] + Y_mean
  dcases <- yhat * fd_sd + fd_mean            # back to case-change units, per province
  fut <- which(dates > hist_end)             # project future months only
  proj <- last_obs + t(apply(dcases[, fut, drop = FALSE], 1, cumsum))
  list(proj = proj, dates = dates[fut])
}

# scenario -> array (province x future-month x model), dropping unstable models
project_scenario <- function(scenario) {
  tas <- cmip_tas[[scenario]]; pr <- cmip_pr[[scenario]]
  nm  <- dim(tas$arr)[3]
  lst <- lapply(seq_len(nm), function(m)
    project_one(tas$arr[, , m], pr$arr[, , m], tas$dates)$proj)
  arr  <- simplify2array(lst)                 # province x month x model
  keep <- apply(arr, 3, function(x) all(is.finite(x)))
  arr[, , keep, drop = FALSE]
}

fut_dates <- project_one(cmip_tas[[1]]$arr[, , 1], cmip_pr[[1]]$arr[, , 1],
                         cmip_tas[[1]]$dates)$dates
proj_arr  <- list("SSP1-2.6" = project_scenario("SSP1-2.6"),
                  "SSP5-8.5" = project_scenario("SSP5-8.5"))
```

```{r part6-plot, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
# Annual totals per province, with ensemble mean and interquartile band --------
annual_proj <- function(arr, dates, scenario) {
  yr <- lubridate::year(dates); years <- sort(unique(yr))
  ap <- sapply(years, function(y)
    apply(arr[, yr == y, , drop = FALSE], c(1, 3), sum))   # -> (province*?)...
  ap <- array(ap, dim = c(length(provinces), dim(arr)[3], length(years)))  # prov x model x year
  ap <- aperm(ap, c(1, 3, 2))                              # prov x year x model
  data.frame(
    province_en = rep(provinces, times = length(years)),
    year = rep(years, each = length(provinces)),
    scenario = scenario,
    mean = as.vector(apply(ap, c(1, 2), mean)),
    q25  = as.vector(apply(ap, c(1, 2), quantile, 0.25)),
    q75  = as.vector(apply(ap, c(1, 2), quantile, 0.75))
  )
}

proj_annual <- bind_rows(
  annual_proj(proj_arr[["SSP1-2.6"]], fut_dates, "SSP1-2.6"),
  annual_proj(proj_arr[["SSP5-8.5"]], fut_dates, "SSP5-8.5")
)

ggplot(proj_annual, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = "projected annual cases",
       title = "Projected annual cases under CMIP6 (statistical/ARX model)",
       subtitle = "Ensemble mean (line) and interquartile range (band). See caveat above.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

**6.1 The projection standardises the future climate using the *historical* mean and
standard deviation, not the future data's own. Why is that essential? What would go
wrong if we standardised the future climate on itself?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**6.2 Look at the projected trajectories and re-read the caveat. Which features of
these plots would you trust, and which would you not? How would you describe the
result honestly to a decision-maker?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 6 — click to reveal (try the questions first)

**6.1** The model was fitted on climate standardised by its *historical* mean and
standard deviation, so the coefficients only make sense on that same scale. If we
re-standardised the future climate on itself, we would subtract off the very warming
(or drying/wetting) trend we are trying to project — every period would look
"average" relative to its own era, and the projected climate signal would vanish. We
must apply the historical scaling so that a genuinely hotter future shows up as a
large positive standardised value.

**6.2** Trust the *direction and relative* differences — e.g. that the higher-
emissions scenario diverges from the lower one, and the ranking of provinces — more
than the absolute case numbers, which drift because of the cumulative-sum
reconstruction. Honest framing to a decision-maker: "this shows how a climate-driven
statistical model *responds* to two emissions futures; the shape and the gap between
scenarios are informative, but the absolute counts decades out are illustrative, not
a forecast, and a mechanistic model would be needed for credible long-range numbers."
:::

---

# Optional: a mechanistic SEIRS alternative

::: speaker-note
This whole section is optional and more advanced — it lives behind an expandable
box so it does not interrupt the main flow, and so Session 3 can stay a clean group
discussion. Run it only if the group powers through Parts 1–6 with time to spare, or
point interested participants to it afterwards. It reuses the case and climate data
already built above (no re-loading), fits a compartmental transmission model, and
doubles as a **worked code base for fitting a transmission model directly**. Keep the
optimiser budget small for a live demo — the fit is deliberately not run to
convergence, and every number is illustrative.
:::

::: {.callout-note collapse="true"}
## Why an SEIRS model? — optional and more advanced (click to expand)

The model in Parts 5–6 is deliberately the simplest thing that works: one shared
statistical relationship between climate and the month-to-month *change* in cases. A
mechanistic **SEIRS** model is the more biologically structured alternative. Instead
of regressing cases on climate, it represents the disease process itself — people
move through Susceptible → Exposed → Infected → Recovered → Susceptible compartments
— and lets **climate drive the transmission rate**. That buys three things the
statistical model cannot: per-province transmission with *shared* natural-history
parameters (so provinces borrow strength); built-in saturation and waning immunity
(so it cannot produce the negative or runaway counts a differenced model can); and a
template you can adapt to fit a transmission model directly to your own data. The
cost is more assumptions and a harder fit.

Each province runs an SEIRS model sharing the natural-history parameters but with a
province- and time-varying transmission rate:

$$\frac{dS}{dt} = -\beta(t) S I + \omega R \quad
  \frac{dE}{dt} = \beta(t) S I - \sigma E \quad
  \frac{dI}{dt} = \sigma E - \gamma I \quad
  \frac{dR}{dt} = \gamma I - \omega R$$

with $\beta_p(t) = \beta_{0,p}\exp(b_{temp}\, z_{temp}(p,t) + b_{rain}\, z_{rain}(p,t))$,
and reported cases modelled as
$\text{cases}_p(t) \sim \text{Poisson}(\text{scale}_p \cdot \sigma E_p(t))$, fitted by
maximum likelihood.

**Caveats — read before trusting any number.** The climate covariates stand in for
missing mosquito data; there are no population denominators (so `scale` absorbs
population size and reporting rate); the natural-history parameters
($\sigma,\gamma,\omega$) are fitted freely rather than fixed to published
vivax/falciparum biology; the two parasite species are pooled; and the optimiser uses
a small iteration budget so it runs quickly — it is **not** run to convergence. Treat
every output as illustrative of method, not as an estimate.

**The data.** The SEIRS reuses the case matrix and per-province climate already built
in Parts 2–3 and 5. The only difference is that it standardises climate **globally**
(one shared scale) rather than per province, so that `b_temp`/`b_rain` are a single
comparable climate sensitivity.

```{r seirs-setup, echo=TRUE, eval=TRUE}
# reuse cases_mat, clim_temp (Kelvin), clim_rain (m), popsd() from Part 5
ztemp <- (clim_temp - mean(clim_temp)) / popsd(clim_temp)
zrain <- (clim_rain - mean(clim_rain)) / popsd(clim_rain)
cat(sprintf("SEIRS inputs: %d provinces x %d months, %.0f total cases\n",
            nrow(cases_mat), ncol(cases_mat), sum(cases_mat)))
```

The model is built up in five small pieces below, then fitted. Read them in order.

**Step 1 — the transmission dynamics (the ODE right-hand side).** This is the SEIRS
model itself: the instantaneous rate of change of each compartment. `state` is a
5 × n matrix — one column per province, rows S, E, I, R, and C (a running count of
new infections) — so everything is vectorised over provinces at once.

```{r seirs-rhs, echo=TRUE, eval=TRUE}
SUBSTEPS_PER_MONTH <- 20   # RK4 sub-steps per month

seirs_rhs <- function(state, beta, sigma, gamma, omega) {
  S <- state[1, ]; E <- state[2, ]; I <- state[3, ]; R <- state[4, ]
  rbind(-beta * S * I + omega * R,     #  dS  (lose to infection, gain from waning immunity)
         beta * S * I - sigma * E,     #  dE  (newly infected, then become infectious)
         sigma * E   - gamma * I,      #  dI  (become infectious, then recover)
         gamma * I   - omega * R,      #  dR  (recover, then lose immunity)
         sigma * E)                    #  dC  (cumulative NEW infections)
}
```

**Step 2 — simulating it forward (fixed-step RK4).** We integrate the ODE month by
month using a fixed-step Runge–Kutta (RK4) scheme with `SUBSTEPS_PER_MONTH` steps per
month, for all provinces simultaneously. A burn-in lets the model settle before we
record anything, and we reset the accumulator `C` at the start of each month so that
`incidence` holds the new infections *within* that month.

```{r seirs-sim, echo=TRUE, eval=TRUE}
simulate_incidence <- function(beta, sigma, gamma, omega,
                               i0 = 1e-4, e0 = 1e-4, burn_in_months = 12) {
  n_prov <- nrow(beta); n_months <- ncol(beta); dt <- 1 / SUBSTEPS_PER_MONTH
  state <- rbind(rep(1 - i0 - e0, n_prov), rep(e0, n_prov),
                 rep(i0, n_prov), rep(0, n_prov), rep(0, n_prov))
  step <- function(state, beta_m) {                    # one RK4 sub-step
    k1 <- seirs_rhs(state,              beta_m, sigma, gamma, omega)
    k2 <- seirs_rhs(state + 0.5*dt*k1,  beta_m, sigma, gamma, omega)
    k3 <- seirs_rhs(state + 0.5*dt*k2,  beta_m, sigma, gamma, omega)
    k4 <- seirs_rhs(state + dt*k3,      beta_m, sigma, gamma, omega)
    state + (dt/6) * (k1 + 2*k2 + 2*k3 + k4)
  }
  if (burn_in_months > 0) {                            # settle using month-1 transmission
    beta0_m <- beta[, 1]
    for (b in seq_len(burn_in_months)) {
      for (s in seq_len(SUBSTEPS_PER_MONTH)) state <- step(state, beta0_m)
      state[5, ] <- 0
    }
  }
  incidence <- matrix(0, n_prov, n_months)
  for (m in seq_len(n_months)) {
    state[5, ] <- 0; beta_m <- beta[, m]               # reset accumulator; use this month's beta
    for (s in seq_len(SUBSTEPS_PER_MONTH)) state <- step(state, beta_m)
    incidence[, m] <- state[5, ]
    state[1:4, ] <- pmin(pmax(state[1:4, ], 0), 1)     # keep compartments in [0, 1]
  }
  incidence
}
```

**Step 3 — unpacking the parameters.** The optimiser works in an unconstrained,
log-transformed space so that rates stay positive. This turns the flat parameter
vector back into named, natural-scale quantities: the shared natural-history rates,
the two climate sensitivities, and the per-province baseline transmission `beta0` and
reporting `scale`.

```{r seirs-unpack, echo=TRUE, eval=TRUE}
unpack_params <- function(params, n) {
  list(sigma = exp(params[1]), gamma = exp(params[2]), omega = exp(params[3]),
       b_temp = params[4], b_rain = params[5],
       beta0 = exp(params[6:(5 + n)]), scale = exp(params[(6 + n):(5 + 2*n)]))
}
```

**Step 4 — the likelihood.** Given a parameter vector, build the climate-forced
transmission rate `beta_t`, simulate the incidence, and score it against the observed
counts with a Poisson negative log-likelihood — the quantity the optimiser minimises.

```{r seirs-nll, echo=TRUE, eval=TRUE}
neg_log_lik <- function(params, cases, ztemp, zrain) {
  n <- nrow(cases); p <- unpack_params(params, n)
  beta_t <- p$beta0 * exp(p$b_temp * ztemp + p$b_rain * zrain)   # climate forces transmission
  beta_t <- pmin(pmax(beta_t, 1e-6), 50)
  incidence <- simulate_incidence(beta_t, p$sigma, p$gamma, p$omega)
  lam <- pmax(p$scale * incidence, 1e-8)                         # expected reported cases
  sum(lam - cases * log(lam))   # Poisson NLL (constant omitted)
}
```

**Step 5 — sensible starting values.** A reasonable initial guess for the optimiser:
incubation ~0.5 months, infectious ~1 month, immunity ~6 months, no climate effect to
begin with, and per-province baselines scaled to each province's mean caseload.

```{r seirs-x0, echo=TRUE, eval=TRUE}
default_x0 <- function(cases) {
  n <- nrow(cases)
  c(log(1/0.5), log(1/1.0), log(1/6.0), 0, 0,      # sigma, gamma, omega, b_temp, b_rain
    log(rep(2.5, n)),                              # beta0 (per province)
    log(pmax(rowMeans(cases), 1) * 50))            # scale (per province)
}
```

**Step 6 — fit.** Now minimise the negative log-likelihood with a bounded optimiser
(`L-BFGS-B`). The `lower`/`upper` vectors keep every parameter in a biologically
plausible range — incubation/infectious periods between ~0.05 and 3 months, immunity
between 0.5 and 36 months, the climate sensitivities within ±2 — and `maxit` is kept
deliberately small so it finishes quickly (raise it for a real fit).

```{r seirs-fit, echo=TRUE, eval=TRUE}
n <- nrow(cases_mat)
x0 <- default_x0(cases_mat)
lower <- c(log(1/3.0), log(1/3.0), log(1/36.0), -2, -2,
           rep(log(0.05), n), rep(log(1.0),  n))
upper <- c(log(1/0.05), log(1/0.05), log(1/0.5), 2, 2,
           rep(log(30),   n), rep(log(1e9),  n))

# maxit kept small so this runs quickly; raise it for a production fit. ndeps sets
# the finite-difference gradient step to match SciPy's default, so this tracks the
# notebook's fit as closely as the two L-BFGS-B implementations allow (see the note
# in python_workflow/compare.R about why the fitted parameters are close, not exact).
fit <- optim(x0, neg_log_lik, method = "L-BFGS-B",
             lower = lower, upper = upper,
             cases = cases_mat, ztemp = ztemp, zrain = zrain,
             control = list(maxit = 200, ndeps = rep(1e-8, length(x0))))
p <- unpack_params(fit$par, n)

cat(sprintf("convergence code=%d  nll=%.1f\n", fit$convergence, fit$value))
cat(sprintf("incubation period 1/sigma = %.2f months\n", 1/p$sigma))
cat(sprintf("infectious period 1/gamma = %.2f months\n", 1/p$gamma))
cat(sprintf("immunity duration 1/omega = %.2f months\n", 1/p$omega))
cat(sprintf("climate sensitivity b_temp = %+.3f  b_rain = %+.3f\n", p$b_temp, p$b_rain))
```

**Step 7 — diagnostics.** Finally, rebuild the fitted incidence from the estimated
parameters and compare it, province by province, against the observed monthly counts.

```{r seirs-diag, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
beta_fit <- p$beta0 * exp(p$b_temp * ztemp + p$b_rain * zrain)
lam_fit  <- p$scale * simulate_incidence(beta_fit, p$sigma, p$gamma, p$omega)

ss_res <- sum((cases_mat - lam_fit)^2)
ss_tot <- sum((cases_mat - mean(cases_mat))^2)
cat(sprintf("pseudo-R2 (vs fitted lambda): %.3f\n", 1 - ss_res / ss_tot))

diag_long <- bind_rows(
  as.data.frame(cases_mat) %>% tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>% mutate(series = "observed"),
  as.data.frame(lam_fit) %>% `rownames<-`(rownames(cases_mat)) %>%
    `colnames<-`(colnames(cases_mat)) %>%
    tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>% mutate(series = "fitted")
) %>% mutate(month = as.Date(month))

ggplot(diag_long, aes(month, cases, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c(observed = "black", fitted = "red")) +
  labs(x = "date", y = "monthly cases",
       title = "Observed vs SEIRS-fitted monthly cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r seirs-export, echo=FALSE, message=FALSE}
# Save the fitted SEIRS outputs for the direct comparison (python_workflow/compare.R).
.odir <- file.path("..", "outputs"); dir.create(.odir, showWarnings = FALSE, recursive = TRUE)
jsonlite::write_json(list(
  selected = provinces, months = as.character(months),
  nll = fit$value,
  sigma = p$sigma, gamma = p$gamma, omega = p$omega,
  b_temp = p$b_temp, b_rain = p$b_rain,
  beta0 = p$beta0, scale = p$scale,
  lam_fit = lam_fit
), file.path(.odir, "r_seirs.json"), matrix = "rowmajor", digits = NA, auto_unbox = TRUE)
```

**A note on identifiability.** Fitting this exact model in both R and Python is
instructive. The two reach the *same* likelihood and predict essentially the same
cases — per-province annual totals agree to about 0.3% — yet their fitted `beta0`
and `scale` differ by a couple of percent. That is a textbook symptom of
**non-identifiability**: `beta0` (baseline transmission) and `scale` (population ×
reporting rate) trade off to set the case *level*, so the optimiser can shuffle
magnitude between them, and the natural-history rates ($\sigma,\gamma,\omega$) are
only weakly constrained by the data. It is concrete evidence for the warning above —
trust the model's *predictions* and the climate sensitivities `b_temp`/`b_rain`, not
the individual parameter values. (This is why the verification in
`python_workflow/compare.R` checks the fitted incidence and climate coefficients, and
treats the level parameters as illustrative.)

**Discussion (SEIRS): the statistical model (Parts 5–6) and this mechanistic model
both link climate to malaria. What can the mechanistic model tell you that the
statistical one cannot, and vice versa? And why are the fitted natural-history
parameters ($\sigma,\gamma,\omega$) a weakness here — how would you fix it?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>
