---
title: "Obtaining, manipulating, and fitting climate data"
subtitle: "SPARKLE Session 2 on climate data in infectious disease modelling"
format:
  training-html: default
presenter: "Michael & Roslyn"
contributors:
  - "Dr Vassili Kitsios"
  - "Prof. Roslyn Hickson"
  - "Dr Michael Lydeamore"
  - "Claude (Anthropic) did the translation of the existing Python workflow to R Quarto; all content to be checked by the course team"
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE,
  results   = "hold"   # show the whole code chunk first, then all its output
)
```

```{css, echo=FALSE}
.speaker-note {
    background-color: #fff8e1;
    border-left: 4px solid #f9a825;
    padding: 10px 14px;
    margin: 14px 0;
    border-radius: 3px;
    font-size: 0.92em;
    color: #5d4037;
}
.speaker-note::before {
    content: "🗒 Facilitator note: ";
    font-weight: bold;
}
.todo-note {
    background-color: #fdecea;
    border-left: 4px solid #d32f2f;
    padding: 10px 14px;
    margin: 14px 0;
    border-radius: 3px;
    font-size: 0.92em;
    color: #611a15;
}
.todo-note::before {
    content: "🚧 TODO (course team): ";
    font-weight: bold;
}
/* Wide printed output (e.g. glimpse, data-frame previews) scrolls sideways
   instead of wrapping, similar to a pandas data-frame preview. */
.cell-output-stdout pre {
    white-space: pre;      /* do not wrap long lines */
    overflow-x: auto;      /* scroll right instead */
}
/* Wide rendered tables (e.g. kable data-frame previews) scroll sideways too. */
.cell-output-display {
    overflow-x: auto;
}
```

::: speaker-note
**Environment.** The R practicals assume the following packages are installed:
`jsonlite`, `dplyr`, `tidyr`, `lubridate`, `ggplot2`, `patchwork`, `sf`,
`terra`, and `ncdf4`. Please confirm these are installed before the session. The
geospatial steps (reanalysis and CMIP6) read netCDF and shapefiles directly and
can take a little while on first render. Using `freeze: auto` caches them
after the first successful render.

**Worked example.** Throughout, the running example is malaria incidence in
Thailand, linked to ERA5 reanalysis climate and CMIP6 future projections.
:::

# Summary {.tabset .tabset-pills .tabset-fade}

This session works through the practical workflow of using climate data in
infectious disease modelling, with malaria in Thailand as the case study: from
obtaining raw data through to a fitted statistical model that links climate to
malaria incidence, with future projections.

The session opens by framing the **research question** and then works through 
six practical parts, with an extension section on fitting an SEIRS transmission 
model:

**The research question.** What we are trying to answer, and how it informs what
data and modelling we need.

**Part 1: Obtaining the data.** Where the data come from (malaria surveillance, 
climate reanalysis, future projections, administrative boundaries) and how to 
obtain them.

**Part 2: Malaria case data.** Reading the surveillance line list and 
summarising cases by month, province, and district.

**Part 3: Environmental variables (reanalysis).** Linking gridded ERA5 climate
data to administrative units using a shapefile, and building per-province 
monthly temperature and rainfall series.

**Part 4: Future climate projections.** Summarising CMIP6 projections under two
emissions scenarios (SSP1-2.6 and SSP5-8.5) for the same provinces.

**Part 5: Fitting climate to incidence.** A statistical (ARIMA/ARX panel) model
relating standardised climate to changes in malaria cases across provinces.

**Part 6: Projecting future cases.** Pushing the fitted statistical model 
forward under the CMIP6 scenarios to project future cases, with the important 
caveats of long-range extrapolation.

**Optional extension: An SEIRS model.** A (basic) transmission model approach.

::: {.callout-tip}
We recommend saving your written answers at the end of each part by printing the
page (there is a **Print this page** button at the end of each part).
:::

This is a port of the python analysis 
(`python_workflow/malaria_workflow.ipynb`), which we have left available for 
those interested.

---

# The research question

::: speaker-note
Spend the first few minutes establishing *what we are trying to answer* and 
*what decision it would inform*. Draw out that the question is what tells us 
(a) whether climate data is even relevant, (b) which data we need, and (c) 
whether it makes sense to project forward. Participants do this same first step 
for their own work in Session 3.
:::

It is important in modelling for the question to determine the methodology, and 
it also informs what data we need. The data available may in turn force us to 
update and refine our question and methodology.

For this case study, our research question is:

> **How do temperature and rainfall relate to malaria incidence across Thailand, 
> and how might that incidence change under future emissions scenarios?**

Reading that question tells us:

- a measure of the **outcome**: malaria cases, over time and by location.
- the **drivers** we hypothesise to matter: temperature and rainfall.
- a **spatial region**: Thailand. The spatial resolution is often driven by 
  data availability. In this case, we will use provinces (the unit cases are 
  reported in).
- a **historical period** where we observe outcome and drivers together, to \
  *learn* the relationship. This is also often driven by data availability.
- **future climate** for the same region, to *project* that relationship 
  forward.

This list is the data we found available, not the initial list assembled. 

**What data do you think is missing from this list, and why?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Suggested answer: click to reveal (try to answer yourself first)

There is no single right answer.

We were initially planning on relating mosquito vector abundance to the climatic
variables, and projecting those forward and linking to malaria incidence via a 
transmission model tracking humans and vectors. But we couldn't find abundance 
data in Thailand at the same time period as the malaria incidence data.
:::

---

# **Part 1:** Obtaining the data

::: speaker-note
This part is about *where the data come from* and *why each source was chosen*. 
:::

The workflow draws on four data sources. Three are used directly in the 
analysis; the fourth (mosquito occurrence) motivates the approach but is not 
currently used. The table below is the quick reference; each source, and why it 
was chosen, is expanded underneath.

| # | Data | What it is / role | Why this source | Source |
|---|--------|-------------------|-----------------|-------------|
| 1 | Malaria cases | national surveillance line list (the outcome) | authoritative, national, open, right resolution | [Thai DDC](https://malaria-api.ddc-malaria.org/study/2year/calendar){target="_blank"} |
| 2 | Spatial region boundaries | province (ADM1) polygons; aggregate grid to province | official ADM1; names align with the case data | [Open Dev Mekong](https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries){target="_blank"} |
| 3 | ERA5 reanalysis | gap-free past climate, used to *learn* | state of the art, ~0.25°, complete coverage, open | [Copernicus CDS](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download){target="_blank"} |
| 4 | CMIP6 projections | multi-model future climate, used to *project* | latest coordinated ensemble; scenarios bracket futures | [Copernicus CDS](https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download){target="_blank"} |

::: {.callout-tip}
Reanalysis (ERA5) is used to *learn* the relationship between climate and malaria
over the historical period; the CMIP6 projections are used to *project* that
relationship forward under future emissions scenarios.
:::

**1. Malaria case data (surveillance line list).** Individual records of malaria
blood draws in Thailand, with date, province, district, and result (*P. vivax* `V`,
*P. falciparum* `F`, or other). This is the outcome we are modelling, and it comes
from the national surveillance system (Thailand's DDC), so it is authoritative and
consistent across the whole country. It is openly queryable via the API (no
data-sharing agreement for this aggregated endpoint) and carries date, province,
district and species per record, at the spatial and temporal resolution we need for
monthly climate. Alternatives like WHO country-year totals are too coarse, and
hospital records are not openly accessible.

*Source and files:* [https://malaria-api.ddc-malaria.org/study/2year/calendar](https://malaria-api.ddc-malaria.org/study/2year/calendar){target="_blank"} , saved as `data/malaria/calendar.json`.

::: {.callout-note appearance="minimal"}
**DDC** is Thailand's Department of Disease Control (its vector-borne disease division runs the malaria surveillance).
:::

**2. Administrative boundaries (shape files).** Province boundaries for Thailand. We
need province polygons because cases are reported per province but climate is
gridded; the polygons let us aggregate the grid to the reporting unit. This set was
chosen because it is the official Thai ADM1 (province) boundaries, so its units and
names align with how the surveillance data is reported (name alignment matters: it
is why some provinces drop out when names do not match), and it is freely available
in a standard GIS format.

*Source and files:* [https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries](https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries){target="_blank"} , saved under `data/shape_files/tha_admbnda_adm1_rtsd_20190221/`.

::: {.callout-note appearance="minimal"}
A **shape file** stores polygon boundaries (countries, provinces, districts); **ADM1** is the first sub-national administrative level, here the provinces.
:::

**3. Climate reanalysis (ERA5).** Gridded, monthly-averaged 2 m temperature (`2t`)
and total precipitation (`tp`): a physically consistent, gap-free reconstruction of
past weather, used here to *learn* how climate relates to malaria. We use reanalysis
rather than weather stations because stations are sparse and gappy, whereas
reanalysis gives complete fields everywhere, which is essential for giving every
province a climate value. We use ERA5 specifically because it is the current
state-of-the-art global reanalysis (~0.25° resolution, well validated, openly
available), carrying the variables we want (temperature and precipitation drive
mosquito abundance and survival) at monthly resolution. It is the de facto standard
for climate and health work.

*Source and files:* [https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download){target="_blank"} , saved under `data/reanalysis/monthly-averaged/`. The required subset is already curated for you; other variables can be downloaded freely from the Copernicus Climate Change Service (C3S) Climate Data Store.

::: {.callout-note appearance="minimal"}
**ERA5** = ECMWF Re-Analysis, 5th generation (ECMWF is the European Centre for Medium-Range Weather Forecasts; the lineage runs ERA-15, ERA-40, ERA-Interim, ERA5).
:::

**4. Future climate projections (CMIP6).** A multi-model ensemble of monthly
temperature (`tas`) and precipitation (`pr`) for Thailand under two emissions
scenarios (SSP1-2.6 and SSP5-8.5), used to *project* the learned relationship into
the future. We use CMIP6 specifically because it is the latest coordinated
multi-model generation (the models underpinning IPCC AR6), delivered in a standard,
comparable format, so we can use a multi-model ensemble to represent model
uncertainty and the SSP scenarios to bracket futures. SSP1-2.6 and SSP5-8.5 were
chosen as a low- and high-emissions pair to show sensitivity to the pathway. It is
also regriddable onto the ERA5 grid, so the historical and future steps stay
consistent.

*Source and files:* [https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download](https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download){target="_blank"} , saved under `data/cmip6/monthly-averaged/`. The data is curated and trimmed to the region for this case study; other variables, scenarios, or the full globe are available from the same Climate Data Store.

::: {.callout-note appearance="minimal"}
**CMIP6** = Coupled Model Intercomparison Project, 6th phase ("coupled" refers to coupled atmosphere and ocean models). In an **SSP** (Shared Socioeconomic Pathway) label such as SSP1-2.6, the first digit is the socioeconomic story (1 = sustainability, 5 = fossil-fuel-intensive) and the second is the resulting radiative forcing (2.6 vs 8.5 W/m² by 2100, i.e. low vs high warming).
:::

**Mosquito occurrence (GBIF), considered but not used.** Mosquito occurrence records
for Thailand. We explored GBIF because it is the largest aggregator of open occurrence
records, so it is the obvious first place to look for vector data. It is not used here
because it gives *occurrence*, not *abundance*, and has no samples after 2022, so it
does not overlap the 2025 to 2026 malaria data. For now we relate malaria incidence in
humans to environmental variables directly (the approach taken in Part 5); regardless,
this is a good, straightforward way to test whether environmental variables such as
these are affecting the health outcomes of interest, and which ones. A natural
extension is to obtain overlapping mosquito data, learn the influence of temperature on
the mosquito population, then the impact of that population on malaria infections via a
transmission model.

*Source and files:* [https://www.gbif.org/dataset/db4211d6-b480-4bc5-9202-87f2072c4b12](https://www.gbif.org/dataset/db4211d6-b480-4bc5-9202-87f2072c4b12){target="_blank"} (not downloaded; not part of the analysis).

::: {.callout-note appearance="minimal"}
**GBIF** is the Global Biodiversity Information Facility, the largest aggregator 
of open species-occurrence records.
:::

::: {.callout-note}
## Obtaining the data yourself
The Copernicus Climate Data Store (ERA5 and CMIP6) requires a free account and,
for bulk downloads, the CDS API (a `~/.cdsapirc` key file). Because these
downloads need accounts/API keys and can be large, they are not run as code 
here. Instead, the curated subset needed for this tutorial is provided 
separately in the `data/` folder at the repository root.
:::

## Ideal vs achievable workflows

::: speaker-note
The key discussion point here: what you *want* to do is often constrained by the
data you can actually get. 
:::

The right workflow is rarely the most sophisticated one, but one that fits 
both your question and that your data can actually support. For malaria, the 
biologically known climate-sensitive link is the *mosquito* population: 
temperature and breeding site available, typically via rainfall, drive mosquito 
abundance, which in turn drives transmission. 

| Approach | Pipeline | Data needed | Biological fidelity |
|---|---|---|---|
| **Ideal** (vector-explicit) | climate → mosquito abundance → transmission model → incidence | overlapping spatio-temporal local mosquito **abundance** data *and* climate | highest |
| **Simplest** (direct statistical) | climate → incidence | overlapping spatio-temporal case data + climate | lowest: no explicit mechanism |
| **Achievable here** (climate-forced SEIRS) | climate → transmission model (transmission forced by climate) → incidence | overlapping spatio-temporal case data + climate | medium: climate link through implicit vector dynamics |

For this example we wanted the **ideal** path, but we couldn't find mosquito 
abundance data that spatio-temporally overlapped the malaria period (see the 
GBIF note above), so we use the lower two tiers: a direct statistical model 
(Part 5) and a climate-forced SEIRS model (the optional section at the end of 
this session). Both skip the mosquito layer. This is a data-driven compromise.

**1.1 For each of the four data sources above, what alternative dataset or 
approach could you use in its place? Think about substitutes for ERA5, for 
CMIP6, and for the provincial case and boundary data, and what you would trade 
off by switching. If you identfied other data earlier, what sources are there?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**1.2 Reanalysis is used to *learn* the climate–disease relationship and CMIP6
projections are used to *project* it into the future. Why can we not simply use 
the future projections to learn the relationship in the first place?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 1: click to reveal (try the questions first)

**1.1** There are alternatives for every source, each with trade-offs. For the
reanalysis, other products such as MERRA-2 or JRA-55, or gridded observational
datasets (for example CHIRPS for rainfall, CRU for temperature), could replace 
ERA5, trading spatial and temporal resolution and which variables are available. 
For the projections, CMIP5 or a downscaled or emulated ensemble could replace 
CMIP6, changing resolution, the scenario definitions, and the model spread. For 
the boundaries, a different administrative product or a gridded population layer 
could be used, changing which spatial unit cases are aggregated to. For the case 
data, a different health outcome or a modelled incidence surface could stand in 
where line-list surveillance is not available. The point is to notice that each 
swap changes resolution, coverage, or assumptions rather than to find one right 
answer.

**1.2** Future projections are, by construction, *projections*: they describe
plausible futures, not observed weather, and there is no malaria data from the
future to pair them with. We can only learn a relationship where we observe 
*both* climate and cases together, which means the historical/reanalysis period. 
The projections are then used to *project* that learned relationship forward.
:::

---

# **Part 2:** Malaria case data

::: speaker-note
This part reads the surveillance line list and aggregates it. The key teaching
points are: 

(1) surveillance data are messy and unevenly reported across provinces and time; 
this is why we later restrict to provinces with enough data; and 

(2) aggregating to monthly counts per province gives us the response variable
that we will link to climate. 

The code is a direct translation of the Python `pandas` workflow. Have 
participants run it and *consider* the province panels (question 2.2) before 
moving on.
:::

We start from the surveillance line list: one row per blood draw, with a date,
province, district, and result. We read it, parse the dates, and aggregate to
monthly counts.

First read in the data from the json file.
```{r part2-read, echo=TRUE, eval=TRUE}
library(jsonlite)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(here)

# Locate the data
data_dir <- here::here("data")

# Read the JSON line list and parse dates ---------------------------------
df <- fromJSON(file.path(data_dir, "malaria", "calendar.json"))
df <- df %>%
  mutate(
    date_daily = as.Date(as.character(blood_draw_date), format = "%Y%m%d"),
    month      = floor_date(date_daily, "month")
  )

# Preview what we have loaded ---------------------------------------------
knitr::kable(head(df, 5))      # the first 5 rows as a clean table in quarto

cat("Number of unique province names:", n_distinct(df$province_en), "\n")
cat("Number of unique district names:", n_distinct(df$district_en), "\n")
cat("Number of unique blood_draw_date:", n_distinct(df$blood_draw_date), "\n")
```

Counts for Thailand per month and per day
```{r part2-counts, echo=TRUE, eval=TRUE, fig.width=10, fig.height=6}
# Total counts per day and per month --------------------------------------
counts_per_day <- df %>% count(date_daily, name = "count") %>% arrange(date_daily)
counts_per_month <- df %>% count(month, name = "count") %>% arrange(month)

p_day <- ggplot(counts_per_day, aes(date_daily, count)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(x = "date", y = "total daily counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_month <- ggplot(counts_per_month, aes(month, count)) +
  geom_line() + geom_point(size = 1) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(x = "date (monthly total, aligned to first of month)",
       y = "total monthly counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_day / p_month)
```

Counts per province per month: since we will be building models estimating the infection counts per province per month, including the impact of the climate.
```{r part2-province, echo=TRUE, eval=TRUE, fig.width=11, fig.height=11}
# Counts per province per month -------------------------------------------
counts_prov_month <- df %>%
  count(month, province_en, name = "Count") %>%
  arrange(month, province_en)

# Keep provinces with more than 12 months of reported cases 
prov_order <- unique(counts_prov_month$province_en)
selected_provinces <- prov_order[
  vapply(prov_order,
         function(p) sum(counts_prov_month$province_en == p) > 12,
         logical(1))
]
cat("Selected provinces:", length(selected_provinces), "\n")

# Facet titles: how many of the possible months each province reports ---------
n_possible <- n_distinct(counts_prov_month$month)     # total months in the data (18)
prov_counts <- counts_prov_month %>%
  filter(province_en %in% selected_provinces) %>%
  group_by(province_en) %>%
  summarise(n_months = n_distinct(month), .groups = "drop")
prov_lab <- setNames(
  paste0(prov_counts$province_en, " (", prov_counts$n_months, "/", n_possible, ")"),
  prov_counts$province_en
)

print(counts_prov_month %>%
  filter(province_en %in% selected_provinces) %>%
  ggplot(aes(month, Count)) +
  geom_line() + geom_point(size = 0.6) +
  facet_wrap(~ province_en, scales = "free_y",
             labeller = labeller(province_en = prov_lab)) +
  labs(x = "date", y = "monthly count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

The natural next question is whether we could work at a finer spatial scale.
```{r part2-district, echo=TRUE, eval=TRUE, fig.width=8, fig.height=7}
# Counts per district per month (finer spatial scale) ---------------------
counts_dist_month <- df %>%
  count(month, district_en, name = "Count")

# Would a finer (district) scale work? 
n_months_total <- n_distinct(df$month)                  # 18 possible months
dist_months <- counts_dist_month %>%
  group_by(district_en) %>%
  summarise(n_months = n_distinct(month), .groups = "drop")

cat("Districts in total:", nrow(dist_months), "\n")
cat("Districts with >12 months of reported cases:",
    sum(dist_months$n_months > 12), "\n")
cat("Provinces with >12 months (from above):", length(selected_provinces), "\n")

# Share of all reported cases falling in the qualifying districts vs provinces
ok_dist <- dist_months$district_en[dist_months$n_months > 12]
cat(sprintf("The %d qualifying districts hold %.1f%% of all reported cases\n",
            length(ok_dist), 100 * mean(df$district_en %in% ok_dist)))
cat(sprintf("The %d qualifying provinces hold %.1f%% of all reported cases\n",
            length(selected_provinces), 100 * mean(df$province_en %in% selected_provinces)))

# Months of reported cases per province and per district
prov_months <- counts_prov_month %>%
  group_by(province_en) %>%
  summarise(n_months = n_distinct(month), .groups = "drop")

months_bar <- function(d, unit_label) {
  ggplot(d, aes(n_months)) +
    geom_bar(fill = "steelblue") +
    geom_vline(xintercept = 12.5, linetype = "dashed", colour = "red") +
    scale_x_continuous(breaks = 1:n_months_total, limits = c(0.5, n_months_total + 0.5)) +
    labs(y = paste0("number of ", unit_label)) +
    theme_bw()
}

p_prov_freq <- months_bar(prov_months, "provinces") +
  labs(x = NULL, title = "Provinces by number of months with reported cases")
p_dist_freq <- months_bar(dist_months, "districts") +
  labs(x = paste0("months with reported cases (out of ", n_months_total, ")"),
       title = "Districts by number of months with reported cases")

print(p_prov_freq / p_dist_freq)
```

**2.1 How many provinces are reported in the surveillance data, and how many are
kept after the "more than 12 months" filter? Why might we want to restrict to
provinces with a longer record before trying to relate cases to climate?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**2.2 Why analyse at the province scale? What do we gain and lose by treating 
the province as the spatial unit, and what would push you to a finer or coarser 
scale?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**2.3 Look at the per-province panels. Describe the seasonal pattern you see. Do
all provinces peak at the same time of year? What might explain differences
between provinces?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**2.4 The results column pools *P. vivax* and *P. falciparum* together. Why might
it matter, for a climate-driven analysis, that these two species are combined?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 2: click to reveal (try the questions first)

**2.1** There are many provinces in the raw data, but only those reported in more
than 12 months are kept (18 provinces kept out of 66), and of those only the ones whose names
match the boundary shapefile are carried forward. Restricting to provinces with a
longer record means each province has enough months to show a seasonal signal and to
standardise sensibly; provinces with only a handful of reported months cannot support
a climate–seasonality analysis.

**2.2** The spatial unit is a trade-off between signal and sparsity, and the 
case counts sharpen it. Going finer would likely work, with 90.8% of the malaria 
data still included. But visualising the 37 districts that clear the ">12 
months" rule vs the 18 provinces is more difficult. It would also spread almost 
the same case burden across roughly twice as many units (37 districts versus 18 
provinces), so each unit's monthly counts are smaller and noisier, and 303 of 
the 340 districts are too sparse to use at all. The province scale captures 
nearly all cases (97.5%) in the fewest robust units, and so it is a pragmatic 
choice. Going finer would suit a focused analysis of the high-burden districts; 
going coarser (national) would give complete series but average over very d
ifferent climates and malaria regimes,likely washing out the signal.

**2.3** Most provinces show a seasonal rise and fall, but not all peak in the same
month; malaria seasonality depends on local climate, elevation, land use, and
the timing of the rains, which differ across the country. Border provinces and
forested provinces often behave differently from central ones.

**2.4** *P. vivax* and *P. falciparum* differ in incubation period, in relapse
behaviour (vivax can relapse from liver-stage hypnozoites months later), and
likely in their climate sensitivity. But they have similar mosquito vectors. So
whether or not pooling them means blurring the climate-malaria signal depends on
how climate-related those factors are, and whether the mosquito relationship 
with climate would dominate. A potential future direction to explore. 
:::

---

# **Part 3:** Environmental variables (reanalysis)

::: speaker-note
The concept to focus on here is the *spatial mismatch*: climate comes on a 
regular grid, cases come per province, and the two steps below bridge it. 

One detail that is not obvious from the code: temperature stays in Kelvin 
through standardisation because a constant offset (Kelvin vs Celsius) does not 
change a standardised (z-scored) series, so it only matters for the plots, where
we convert to °C.
:::

Malaria is reported per province, but ERA5 climate comes on a regular grid. To
link them, we summarise the grid to provinces in two steps: **strict cell-centre
assignment**, where each grid cell is assigned to the province whose polygon 
contains the cell's centre (a point-in-polygon test with `sf`, matching the
[`regionmask`](https://regionmask.readthedocs.io){target="_blank"} package), and 
then an **area-weighted average** of those cells, weighting each by cos of its 
latitude so that cells count in proportion to their area.

First load the shape files and check that the surveillance province names match 
the shapefile's ADM1_EN names. Keep only the selected provinces that are present 
in the boundaries. Report any that are dropped so name mismatches are visible.
```{r part3-shape, echo=TRUE, eval=TRUE}
library(sf)
library(terra)

# Province (ADM1) boundaries ----------------------------------------------
shape_file <- st_read(
  file.path(data_dir, "shape_files",
            "tha_admbnda_adm1_rtsd_20190221",
            "tha_admbnda_adm1_rtsd_20190221.shp"),
  quiet = TRUE
)
region_names <- shape_file$ADM1_EN

# Preview what we have loaded ---------------------------------------------
knitr::kable(head(shape_file, 5))      # the first 5 rows as a clean table in quarto

cat("Number of provinces in shapefile:", length(region_names), "\n")

missing <- setdiff(selected_provinces, region_names)
if (length(missing))
  cat("Provinces not matched in shapefile and dropped: ",
          paste(missing, collapse = ", "), "\n")

selected_provinces <- selected_provinces[selected_provinces %in% region_names]
cat("Provinces carried forward (matched to boundaries):",
    length(selected_provinces), "\n")
```

Next, read the ERA5 monthly reanalysis: one netCDF file per month for 2 m
temperature (`2t`) and for total precipitation (`tp`), cropped to the Thailand 
region.

```{r part3-reanalysis, echo=TRUE, eval=TRUE}
# One netCDF file per month; sorting the file paths puts them in time order.
tas_files <- sort(list.files(
  file.path(data_dir, "reanalysis", "monthly-averaged", "2t"),
  pattern = "\\.nc$", recursive = TRUE, full.names = TRUE))
pr_files <- sort(list.files(
  file.path(data_dir, "reanalysis", "monthly-averaged", "tp"),
  pattern = "\\.nc$", recursive = TRUE, full.names = TRUE))

tas_r <- rast(tas_files)   # temperature, one layer per month (Kelvin)
pr_r  <- rast(pr_files)    # total precipitation, one layer per month (metres)

# Restrict to the Thailand region 
region_ext <- ext(97, 108, 5, 22)
tas_r <- crop(tas_r, region_ext)
pr_r  <- crop(pr_r,  region_ext)

# Recover the month for each layer from the filename (YYYYMMDD-...)
month_from_file <- function(f)
  floor_date(as.Date(sub(".*_(\\d{8})-\\d{8}\\.nc$", "\\1", basename(f)),
                     format = "%Y%m%d"), "month")
clim_months <- month_from_file(tas_files)
cat("Reanalysis covers", length(clim_months), "months:",
    format(min(clim_months)), "to", format(max(clim_months)), "\n")
```

As a quick check that the data cover the right region, map the time-mean 
temperature and rainfall with the province boundaries drawn on top.

```{r part3-maps, echo=TRUE, eval=TRUE, fig.width=11, fig.height=5}
# Map of the time-mean temperature and rainfall over the region -----------
tas_mean_c <- mean(tas_r) - 273.15          # Kelvin -> Celsius for display
pr_mean    <- mean(pr_r)

map_df_t <- as.data.frame(tas_mean_c, xy = TRUE); names(map_df_t)[3] <- "value"
map_df_r <- as.data.frame(pr_mean,    xy = TRUE); names(map_df_r)[3] <- "value"

m_t <- ggplot(map_df_t, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey30",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_c(name = "°C") +
  coord_sf(expand = FALSE) +
  labs(title = "Mean temperature", x = "lon", y = "lat") +
  theme_bw()

m_r <- ggplot(map_df_r, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey30",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_c(name = "m", option = "mako") +
  coord_sf(expand = FALSE) +
  labs(title = "Mean rainfall", x = "lon", y = "lat") +
  theme_bw()

print(m_t + m_r)
```

Now assign each grid cell to a province and take the area-weighted provincial mean
for every month, giving one temperature and one rainfall series per province.

```{r part3-extract, echo=TRUE, eval=TRUE, fig.width=6, fig.height=7}
# Assign each grid cell to the province whose polygon contains its *centre*.
grid_xy  <- terra::xyFromCell(tas_r, seq_len(terra::ncell(tas_r)))
grid_pts <- sf::st_as_sf(as.data.frame(grid_xy), coords = c("x", "y"),
                         crs = sf::st_crs(shape_file))
within   <- sf::st_within(grid_pts, shape_file)          # cell centre in polygon?
cell_region <- vapply(within,
                      function(h) if (length(h)) h[1] else NA_integer_,
                      integer(1))

# Check the assignment: colour each grid cell by the province it landed in.
assign_df <- data.frame(x = grid_xy[, 1], y = grid_xy[, 2], province_id = cell_region)
print(ggplot(subset(assign_df, !is.na(province_id)), aes(x, y, fill = province_id)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey20",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_c(guide = "none") +
  coord_sf(expand = FALSE) +
  labs(title = "Grid cells coloured by assigned province", x = "lon", y = "lat") +
  theme_bw())

# Provincial means, area-weighted by cos(latitude); see technical note.
# The assignment and weights are reused for the CMIP data in Part 4.
cell_w <- cos(grid_xy[, 2] * pi / 180)                    # per-cell area weight (cos latitude)
region_means <- function(r) {
  vals <- terra::values(r)                                # ncell x nlayer
  out  <- matrix(NA_real_, length(region_names), ncol(vals),
                 dimnames = list(region_names, NULL))
  for (k in seq_along(region_names)) {
    sel <- which(cell_region == k)
    if (!length(sel)) next
    w <- cell_w[sel]
    out[k, ] <- colSums(vals[sel, , drop = FALSE] * w, na.rm = TRUE) / sum(w)
  }
  out
}

tas_prov <- region_means(tas_r)
pr_prov  <- region_means(pr_r)

# Tidy into long tables, keeping only the selected provinces
to_long <- function(mat, value_name) {
  m <- as.matrix(mat)
  rownames(m) <- region_names
  colnames(m) <- as.character(clim_months)
  as.data.frame(m) %>%
    tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = value_name) %>%
    mutate(month = as.Date(month))
}

clim_long <- to_long(tas_prov, "temp_k") %>%
  left_join(to_long(pr_prov, "rain_m"), by = c("province_en", "month")) %>%
  mutate(temp_c = temp_k - 273.15) %>%
  filter(province_en %in% selected_provinces)

knitr::kable(head(clim_long))
```

::: {.callout-note collapse="true"}
## Technical note: Why cos of latitude is the exact area weight

A grid cell's area is its north-south height times its east-west width. On a regular
grid the height is the same for every cell (a fixed step in latitude is a fixed
distance $R\,\Delta\theta$ along the surface), so all the variation is in the width.
The width shrinks towards the poles, because the meridians (lines of longitude)
converge: at latitude $\theta$ the circle of latitude has radius $R\cos\theta$, not
the full $R$.

```{=html}
<svg viewBox="0 0 420 300" width="100%" style="max-width:420px;height:auto;display:block;margin:0.6rem auto" xmlns="http://www.w3.org/2000/svg" role="img" aria-label="Cross-section of a globe: at latitude theta the circle of latitude has radius R times cosine theta, smaller than the equator's radius R.">
  <circle cx="200" cy="160" r="105" fill="#eaf2fb" stroke="#274b6d" stroke-width="1.5"/>
  <line x1="200" y1="47" x2="200" y2="273" stroke="#274b6d" stroke-width="1" stroke-dasharray="4 3"/>
  <text x="200" y="40" text-anchor="middle" font-size="11" fill="#274b6d">pole</text>
  <ellipse cx="200" cy="160" rx="105" ry="24" fill="none" stroke="#274b6d" stroke-width="1" stroke-dasharray="3 3"/>
  <text x="310" y="160" font-size="11" fill="#274b6d">equator (radius R)</text>
  <ellipse cx="200" cy="86" rx="74" ry="13" fill="none" stroke="#b2182b" stroke-width="1.7"/>
  <line x1="200" y1="160" x2="274" y2="86" stroke="#274b6d" stroke-width="1.6"/>
  <text x="248" y="120" font-size="13" fill="#274b6d" font-style="italic">R</text>
  <line x1="200" y1="86" x2="274" y2="86" stroke="#b2182b" stroke-width="2.4"/>
  <circle cx="274" cy="86" r="3.4" fill="#b2182b"/>
  <text x="282" y="84" font-size="12.5" fill="#b2182b" font-style="italic">P</text>
  <text x="237" y="79" font-size="12.5" fill="#b2182b" text-anchor="middle">R cos &#952;</text>
  <text x="118" y="89" font-size="11" fill="#b2182b" text-anchor="end">circle of latitude &#952;</text>
  <line x1="120" y1="86" x2="128" y2="86" stroke="#b2182b" stroke-width="1"/>
  <line x1="200" y1="160" x2="305" y2="160" stroke="#274b6d" stroke-width="1" stroke-dasharray="2 2"/>
  <text x="226" y="153" font-size="13" fill="#274b6d" font-style="italic">&#952;</text>
  <path d="M200,94 L208,94 L208,86" fill="none" stroke="#274b6d" stroke-width="1"/>
  <path d="M258,86 A16 16 0 0 0 262.6,97.3" fill="none" stroke="#274b6d" stroke-width="1"/>
  <text x="252" y="101" font-size="12" fill="#274b6d" font-style="italic">&#952;</text>
</svg>
```

Why $R\cos\theta$? In the cross-section above, the line from the centre out to $P$ is
a radius of the sphere, so it has length $R$, exactly like the equator's radius.
Because the equator and the $R\cos\theta$ segment are parallel, the angle at $P$
equals the latitude $\theta$ (right angle marked where $R\cos\theta$ meets the axis),
so in that right triangle $R\cos\theta$ is the side adjacent to $\theta$. A fixed
longitude span $\Delta\lambda$ therefore covers an east-west distance
$R\cos\theta\,\Delta\lambda$.

Now take a whole cell, bounded by latitudes $\theta_1$ and $\theta_2$ and a longitude
span $\Delta\lambda$:

```{=html}
<!-- Second diagram. To reproduce: run `python3 make_area_weight_diagram.py` (in this folder); it writes area_weight_diagram.svg, then paste that file's <svg>...</svg> in here. Do not hand-edit the path data. -->
<svg viewBox="0 0 430 320" width="100%" style="max-width:430px;height:auto;display:block;margin:0.6rem auto" xmlns="http://www.w3.org/2000/svg" role="img" aria-label="A globe with two meridians running pole to pole that bound a highlighted grid cell between latitudes theta-one and theta-two spanning delta-lambda in longitude.">
<circle cx="205" cy="158" r="105" fill="#eaf2fb" stroke="#274b6d" stroke-width="1.5"/>
<path d="M205.0,202.4 L209.6,202.3 L214.2,202.2 L218.7,202.0 L223.2,201.7 L227.7,201.3 L232.2,200.9 L236.6,200.3 L240.9,199.7 L245.2,199.0 L249.4,198.2 L253.5,197.4 L257.5,196.4 L261.4,195.4 L265.2,194.3 L268.9,193.2 L272.5,192.0 L275.9,190.7 L279.2,189.4 L282.4,188.0 L285.4,186.5 L288.3,185.0 L291.0,183.5 L293.6,181.8 L295.9,180.2 L298.1,178.5 L300.2,176.8 L302.0,175.0 L303.7,173.2 L305.1,171.3 L306.4,169.5 L307.5,167.6 L308.4,165.7 L309.1,163.8 L309.6,161.9 L309.9,159.9 L310.0,158.0" fill="none" stroke="#9bb0c4" stroke-width="1"/><path d="M309.9,156.1 L309.6,154.1 L309.1,152.2 L308.4,150.3 L307.5,148.4 L306.4,146.5 L305.1,144.7 L303.7,142.8 L302.0,141.0 L300.2,139.2 L298.1,137.5 L295.9,135.8 L293.6,134.2 L291.0,132.5 L288.3,131.0 L285.4,129.5 L282.4,128.0 L279.2,126.6 L275.9,125.3 L272.5,124.0 L268.9,122.8 L265.2,121.7 L261.4,120.6 L257.5,119.6 L253.5,118.6 L249.4,117.8 L245.2,117.0 L240.9,116.3 L236.6,115.7 L232.2,115.1 L227.7,114.7 L223.2,114.3 L218.7,114.0 L214.2,113.8 L209.6,113.7 L205.0,113.6 L200.4,113.7 L195.8,113.8 L191.3,114.0 L186.8,114.3 L182.3,114.7 L177.8,115.1 L173.4,115.7 L169.1,116.3 L164.8,117.0 L160.6,117.8 L156.5,118.6 L152.5,119.6 L148.6,120.6 L144.8,121.7 L141.1,122.8 L137.5,124.0 L134.1,125.3 L130.8,126.6 L127.6,128.0 L124.6,129.5 L121.7,131.0 L119.0,132.5 L116.4,134.2 L114.1,135.8 L111.9,137.5 L109.8,139.2 L108.0,141.0 L106.3,142.8 L104.9,144.7 L103.6,146.5 L102.5,148.4 L101.6,150.3 L100.9,152.2 L100.4,154.1 L100.1,156.1 L100.0,158.0" fill="none" stroke="#9bb0c4" stroke-width="1" stroke-dasharray="2 3" opacity="0.5"/><path d="M100.1,159.9 L100.4,161.9 L100.9,163.8 L101.6,165.7 L102.5,167.6 L103.6,169.5 L104.9,171.3 L106.3,173.2 L108.0,175.0 L109.8,176.8 L111.9,178.5 L114.1,180.2 L116.4,181.8 L119.0,183.5 L121.7,185.0 L124.6,186.5 L127.6,188.0 L130.8,189.4 L134.1,190.7 L137.5,192.0 L141.1,193.2 L144.8,194.3 L148.6,195.4 L152.5,196.4 L156.5,197.4 L160.6,198.2 L164.8,199.0 L169.1,199.7 L173.4,200.3 L177.8,200.9 L182.3,201.3 L186.8,201.7 L191.3,202.0 L195.8,202.2 L200.4,202.3 L205.0,202.4" fill="none" stroke="#9bb0c4" stroke-width="1"/>
<path d="M205.0,176.2 L209.4,176.2 L213.8,176.1 L218.2,175.9 L222.6,175.6 L227.0,175.2 L231.2,174.8 L235.5,174.2 L239.7,173.6 L243.8,173.0 L247.9,172.2 L251.8,171.4 L255.7,170.5 L259.5,169.5 L263.2,168.5 L266.7,167.4 L270.2,166.2 L273.5,165.0 L276.7,163.7 L279.8,162.3 L282.7,160.9 L285.5,159.5 L288.1,158.0 L290.5,156.4 L292.8,154.8 L295.0,153.2 L296.9,151.5 L298.7,149.8 L300.3,148.0 L301.7,146.3 L303.0,144.5 L304.0,142.6 L304.9,140.8 L305.6,139.0 L306.0,137.1 L306.3,135.2 L306.4,133.4 L306.3,131.5 L306.0,129.6" fill="none" stroke="#6b8cae" stroke-width="1.2"/><path d="M305.6,127.8 L304.9,125.9 L304.0,124.1 L303.0,122.3 L301.7,120.5 L300.3,118.7 L298.7,117.0 L296.9,115.3 L295.0,113.6 L292.8,111.9 L290.5,110.3 L288.1,108.8 L285.5,107.3 L282.7,105.8 L279.8,104.4 L276.7,103.1 L273.5,101.8 L270.2,100.5 L266.7,99.4 L263.2,98.3 L259.5,97.2 L255.7,96.2 L251.8,95.4 L247.9,94.5 L243.8,93.8 L239.7,93.1 L235.5,92.5 L231.3,92.0 L227.0,91.5 L222.6,91.2 L218.2,90.9 L213.8,90.7 L209.4,90.5 L205.0,90.5 L200.6,90.5 L196.2,90.7 L191.8,90.9 L187.4,91.2 L183.0,91.5 L178.8,92.0 L174.5,92.5 L170.3,93.1 L166.2,93.8 L162.1,94.5 L158.2,95.4 L154.3,96.2 L150.5,97.2 L146.8,98.3 L143.3,99.4 L139.8,100.5 L136.5,101.8 L133.3,103.1 L130.2,104.4 L127.3,105.8 L124.5,107.3 L121.9,108.8 L119.5,110.3 L117.2,111.9 L115.0,113.6 L113.1,115.3 L111.3,117.0 L109.7,118.7 L108.3,120.5 L107.0,122.3 L106.0,124.1 L105.1,125.9 L104.4,127.8" fill="none" stroke="#6b8cae" stroke-width="1.2" stroke-dasharray="2 3" opacity="0.5"/><path d="M104.0,129.6 L103.7,131.5 L103.6,133.4 L103.7,135.2 L104.0,137.1 L104.4,139.0 L105.1,140.8 L106.0,142.6 L107.0,144.5 L108.3,146.3 L109.7,148.0 L111.3,149.8 L113.1,151.5 L115.0,153.2 L117.2,154.8 L119.5,156.4 L121.9,158.0 L124.5,159.5 L127.3,160.9 L130.2,162.3 L133.3,163.7 L136.5,165.0 L139.8,166.2 L143.3,167.4 L146.8,168.5 L150.5,169.5 L154.3,170.5 L158.2,171.4 L162.1,172.2 L166.2,173.0 L170.3,173.6 L174.5,174.2 L178.8,174.8 L183.0,175.2 L187.4,175.6 L191.8,175.9 L196.2,176.1 L200.6,176.2 L205.0,176.2" fill="none" stroke="#6b8cae" stroke-width="1.2"/>
<path d="M205.0,117.0 L208.1,116.9 L211.1,116.9 L214.2,116.7 L217.2,116.5 L220.2,116.3 L223.2,116.0 L226.1,115.6 L229.0,115.2 L231.9,114.7 L234.7,114.2 L237.4,113.6 L240.1,113.0 L242.7,112.3 L245.3,111.6 L247.8,110.8 L250.2,110.0 L252.5,109.2 L254.7,108.3 L256.8,107.3 L258.8,106.4 L260.7,105.4 L262.6,104.3 L264.3,103.2 L265.8,102.1 L267.3,101.0 L268.7,99.8 L269.9,98.6 L271.0,97.4 L272.0,96.2 L272.9,95.0 L273.6,93.7 L274.2,92.4 L274.7,91.2 L275.0,89.9 L275.2,88.6 L275.3,87.3 L275.2,86.0 L275.0,84.7 L274.7,83.4 L274.2,82.1 L273.6,80.9 L272.9,79.6 L272.0,78.4 L271.0,77.1 L269.9,75.9 L268.7,74.7 L267.3,73.6 L265.8,72.4" fill="none" stroke="#6b8cae" stroke-width="1.2"/><path d="M264.3,71.3 L262.6,70.2 L260.7,69.2 L258.8,68.2 L256.8,67.2 L254.7,66.3 L252.5,65.4 L250.2,64.5 L247.8,63.7 L245.3,63.0 L242.7,62.2 L240.1,61.6 L237.4,60.9 L234.7,60.4 L231.9,59.8 L229.0,59.4 L226.1,59.0 L223.2,58.6 L220.2,58.3 L217.2,58.0 L214.2,57.8 L211.1,57.7 L208.1,57.6 L205.0,57.6 L201.9,57.6 L198.9,57.7 L195.8,57.8 L192.8,58.0 L189.8,58.3 L186.8,58.6 L183.9,59.0 L181.0,59.4 L178.1,59.8 L175.3,60.4 L172.6,60.9 L169.9,61.6 L167.3,62.2 L164.7,63.0 L162.2,63.7 L159.8,64.5 L157.5,65.4 L155.3,66.3 L153.2,67.2 L151.2,68.2 L149.3,69.2 L147.4,70.2 L145.7,71.3" fill="none" stroke="#6b8cae" stroke-width="1.2" stroke-dasharray="2 3" opacity="0.5"/><path d="M144.2,72.4 L142.7,73.6 L141.3,74.7 L140.1,75.9 L139.0,77.1 L138.0,78.4 L137.1,79.6 L136.4,80.9 L135.8,82.1 L135.3,83.4 L135.0,84.7 L134.8,86.0 L134.7,87.3 L134.8,88.6 L135.0,89.9 L135.3,91.2 L135.8,92.4 L136.4,93.7 L137.1,95.0 L138.0,96.2 L139.0,97.4 L140.1,98.6 L141.3,99.8 L142.7,101.0 L144.2,102.1 L145.7,103.2 L147.4,104.3 L149.3,105.4 L151.2,106.4 L153.2,107.3 L155.3,108.3 L157.5,109.2 L159.8,110.0 L162.2,110.8 L164.7,111.6 L167.3,112.3 L169.9,113.0 L172.6,113.6 L175.3,114.2 L178.1,114.7 L181.0,115.2 L183.9,115.6 L186.8,116.0 L189.8,116.3 L192.8,116.5 L195.8,116.7 L198.9,116.9 L201.9,116.9 L205.0,117.0" fill="none" stroke="#6b8cae" stroke-width="1.2"/>
<path d="M205.0,253.2 L203.5,254.1 L202.1,255.0 L200.6,255.8 L199.2,256.6 L197.7,257.3 L196.3,257.9 L194.9,258.4 L193.4,258.9 L192.0,259.3 L190.6,259.7 L189.2,259.9 L187.8,260.1 L186.4,260.3 L185.1,260.3 L183.7,260.3 L182.4,260.2 L181.0,260.1 L179.7,259.9 L178.5,259.6" fill="none" stroke="#274b6d" stroke-width="1.4" stroke-dasharray="2 3" opacity="0.5"/><path d="M177.2,259.2 L175.9,258.8 L174.7,258.3 L173.5,257.7 L172.3,257.1 L171.1,256.4 L170.0,255.6 L168.9,254.8 L167.8,253.9 L166.7,252.9 L165.7,251.9 L164.6,250.8 L163.7,249.6 L162.7,248.4 L161.8,247.1 L160.9,245.8 L160.0,244.4 L159.1,242.9 L158.3,241.4 L157.6,239.8 L156.8,238.2 L156.1,236.5 L155.4,234.7 L154.8,232.9 L154.2,231.1 L153.6,229.2 L153.1,227.2 L152.5,225.2 L152.1,223.2 L151.6,221.1 L151.3,219.0 L150.9,216.8 L150.6,214.6 L150.3,212.3 L150.0,210.1 L149.8,207.7 L149.7,205.4 L149.5,203.0 L149.4,200.6 L149.4,198.1 L149.4,195.6 L149.4,193.1 L149.4,190.6 L149.5,188.0 L149.7,185.5 L149.8,182.9 L150.0,180.3 L150.3,177.7 L150.6,175.0 L150.9,172.4 L151.3,169.7 L151.6,167.1 L152.1,164.4 L152.5,161.7 L153.1,159.0 L153.6,156.4 L154.2,153.7 L154.8,151.0 L155.4,148.3 L156.1,145.7 L156.8,143.0 L157.6,140.4 L158.3,137.7 L159.1,135.1 L160.0,132.5 L160.9,129.9 L161.8,127.4 L162.7,124.8 L163.7,122.3 L164.6,119.8 L165.7,117.3 L166.7,114.9 L167.8,112.5 L168.9,110.1 L170.0,107.7 L171.1,105.4 L172.3,103.1 L173.5,100.9 L174.7,98.7 L175.9,96.5 L177.2,94.4 L178.5,92.3 L179.7,90.3 L181.0,88.3 L182.4,86.4 L183.7,84.5 L185.1,82.6 L186.4,80.9 L187.8,79.1 L189.2,77.4 L190.6,75.8 L192.0,74.3 L193.4,72.7 L194.9,71.3 L196.3,69.9 L197.7,68.6 L199.2,67.3 L200.6,66.1 L202.1,64.9 L203.5,63.9 L205.0,62.8" fill="none" stroke="#274b6d" stroke-width="1.4"/>
<path d="M205.0,253.2 L206.5,254.1 L207.9,255.0 L209.4,255.8 L210.8,256.6 L212.3,257.3 L213.7,257.9 L215.1,258.4 L216.6,258.9 L218.0,259.3 L219.4,259.7 L220.8,259.9 L222.2,260.1 L223.6,260.3 L224.9,260.3 L226.3,260.3 L227.6,260.2 L229.0,260.1 L230.3,259.9 L231.5,259.6" fill="none" stroke="#274b6d" stroke-width="1.4" stroke-dasharray="2 3" opacity="0.5"/><path d="M232.8,259.2 L234.1,258.8 L235.3,258.3 L236.5,257.7 L237.7,257.1 L238.9,256.4 L240.0,255.6 L241.1,254.8 L242.2,253.9 L243.3,252.9 L244.3,251.9 L245.4,250.8 L246.3,249.6 L247.3,248.4 L248.2,247.1 L249.1,245.8 L250.0,244.4 L250.9,242.9 L251.7,241.4 L252.4,239.8 L253.2,238.2 L253.9,236.5 L254.6,234.7 L255.2,232.9 L255.8,231.1 L256.4,229.2 L256.9,227.2 L257.5,225.2 L257.9,223.2 L258.4,221.1 L258.7,219.0 L259.1,216.8 L259.4,214.6 L259.7,212.3 L260.0,210.1 L260.2,207.7 L260.3,205.4 L260.5,203.0 L260.6,200.6 L260.6,198.1 L260.6,195.6 L260.6,193.1 L260.6,190.6 L260.5,188.0 L260.3,185.5 L260.2,182.9 L260.0,180.3 L259.7,177.7 L259.4,175.0 L259.1,172.4 L258.7,169.7 L258.4,167.1 L257.9,164.4 L257.5,161.7 L256.9,159.0 L256.4,156.4 L255.8,153.7 L255.2,151.0 L254.6,148.3 L253.9,145.7 L253.2,143.0 L252.4,140.4 L251.7,137.7 L250.9,135.1 L250.0,132.5 L249.1,129.9 L248.2,127.4 L247.3,124.8 L246.3,122.3 L245.4,119.8 L244.3,117.3 L243.3,114.9 L242.2,112.5 L241.1,110.1 L240.0,107.7 L238.9,105.4 L237.7,103.1 L236.5,100.9 L235.3,98.7 L234.1,96.5 L232.8,94.4 L231.5,92.3 L230.3,90.3 L229.0,88.3 L227.6,86.4 L226.3,84.5 L224.9,82.6 L223.6,80.9 L222.2,79.1 L220.8,77.4 L219.4,75.8 L218.0,74.3 L216.6,72.7 L215.1,71.3 L213.7,69.9 L212.3,68.6 L210.8,67.3 L209.4,66.1 L207.9,64.9 L206.5,63.9 L205.0,62.8" fill="none" stroke="#274b6d" stroke-width="1.4"/>
<path d="M151.3,169.7 L154.6,170.6 L158.0,171.4 L161.5,172.1 L165.1,172.8 L168.7,173.4 L172.4,174.0 L176.1,174.5 L179.9,174.9 L183.7,175.3 L187.5,175.6 L191.4,175.8 L195.2,176.0 L199.1,176.2 L203.0,176.2 L207.0,176.2 L210.9,176.2 L214.8,176.0 L218.6,175.8 L222.5,175.6 L226.3,175.3 L230.1,174.9 L233.9,174.5 L237.6,174.0 L241.3,173.4 L244.9,172.8 L248.5,172.1 L252.0,171.4 L255.4,170.6 L258.7,169.7 L258.7,169.7 L258.3,166.6 L257.8,163.5 L257.2,160.4 L256.6,157.3 L256.0,154.2 L255.3,151.1 L254.5,148.0 L253.7,145.0 L252.9,141.9 L252.0,138.8 L251.1,135.8 L250.1,132.8 L249.1,129.8 L248.0,126.8 L247.0,123.9 L245.8,121.0 L244.7,118.1 L243.5,115.3 L242.2,112.5 L242.2,112.5 L239.9,113.0 L237.5,113.6 L235.1,114.1 L232.6,114.6 L230.1,115.0 L227.6,115.4 L225.0,115.7 L222.4,116.0 L219.8,116.3 L217.1,116.5 L214.4,116.7 L211.8,116.8 L209.1,116.9 L206.4,117.0 L203.6,117.0 L200.9,116.9 L198.2,116.8 L195.6,116.7 L192.9,116.5 L190.2,116.3 L187.6,116.0 L185.0,115.7 L182.4,115.4 L179.9,115.0 L177.4,114.6 L174.9,114.1 L172.5,113.6 L170.1,113.0 L167.8,112.5 L167.8,112.5 L166.5,115.3 L165.3,118.1 L164.2,121.0 L163.0,123.9 L162.0,126.8 L160.9,129.8 L159.9,132.8 L158.9,135.8 L158.0,138.8 L157.1,141.9 L156.3,145.0 L155.5,148.0 L154.7,151.1 L154.0,154.2 L153.4,157.3 L152.8,160.4 L152.2,163.5 L151.7,166.6 L151.3,169.7 Z" fill="#f6cccc" fill-opacity="0.8" stroke="#b2182b" stroke-width="1.8"/>
<circle cx="205.0" cy="62.8" r="2.4" fill="#274b6d"/>
<text x="205.0" y="52.8" text-anchor="middle" font-size="11" fill="#274b6d">pole</text>
<text x="97.6" y="137.4" text-anchor="end" font-size="12" fill="#274b6d" font-style="italic">&#952;&#8321;</text>
<text x="128.7" y="91.3" text-anchor="end" font-size="12" fill="#274b6d" font-style="italic">&#952;&#8322;</text>
<line x1="167.8" y1="104.5" x2="242.2" y2="104.5" stroke="#b2182b" stroke-width="1.2"/>
<text x="205.0" y="100.5" text-anchor="middle" font-size="12" fill="#b2182b">&#916;&#955;</text>
<text x="205.0" y="150.1" text-anchor="middle" font-size="12" fill="#b2182b">area A</text>
</svg>
```

The cell is not one rectangle: its width $R\cos\theta\,\Delta\lambda$ keeps shrinking
from $\theta_1$ up to $\theta_2$. So slice it into thin strips of latitude. A strip at
latitude $\theta$ has width $R\cos\theta\,\Delta\lambda$ and height $R\,d\theta$, so
area $R^2\,\Delta\lambda\,\cos\theta\,d\theta$. Adding the strips from $\theta_1$ to
$\theta_2$ (a sum that calculus writes as $\int\cos\theta\,d\theta$) gives the cell's
exact area

$$A = R^2\,\Delta\lambda\,\bigl(\sin\theta_2 - \sin\theta_1\bigr),$$

For a cell of height $\Delta\theta$ centred at $\theta$, the identity
$\sin\!\left(\theta+\tfrac{\Delta\theta}{2}\right) - \sin\!\left(\theta-\tfrac{\Delta\theta}{2}\right) = 2\cos\theta\,\sin\!\left(\tfrac{\Delta\theta}{2}\right)$
turns this into

$$A = R^2\,\Delta\lambda \,\cdot\, 2\sin\!\left(\tfrac{\Delta\theta}{2}\right) \,\cdot\, \cos\theta.$$

Every factor except $\cos\theta$ is identical for all cells on a regular grid, and
constant factors cancel in a normalised weighted mean. So weighting each cell by
$\cos$ of its centre latitude reproduces the true area weighting exactly: it is not a
small-angle approximation, it is exact for a sphere. (The Earth's slight oblateness
shifts cell areas by a fraction of a percent, far below anything that matters at this
provincial scale.)
:::

::: {.callout-note appearance="minimal"}
We weight each cell by its area (cos of latitude), the standard spatial mean. 
For climate and health work an even better choice is **population-density 
weighting**, which captures the climate where people actually live (noting this 
also requires data on the human population density). That matters most for
countries with very uneven populations (for example Australia, where most people 
are concentrated in a few coastal cities); at Thailand's provincial scale the 
difference is small, so we keep area-weighting here.
:::

Plot the resulting monthly temperature and rainfall series for each selected province.

```{r part3-series, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
print(ggplot(clim_long, aes(month, temp_c)) +
  geom_line() + geom_point(size = 0.5) +
  facet_wrap(~ province_en, scales = "free_y") +
  labs(x = "date", y = "monthly mean temperature (°C)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))

print(ggplot(clim_long, aes(month, rain_m)) +
  geom_line() + geom_point(size = 0.5) +
  facet_wrap(~ province_en, scales = "free_y") +
  labs(x = "date", y = "monthly rainfall (m)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

**3.1 We reduced each province's climate to a single monthly average. What might
that miss, and when might it matter?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**3.2 Look at the per-province temperature series. What kinds of variation do you
see, and which do you think will matter most for explaining malaria seasonality?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**3.3 The reanalysis only covers a couple of years here. What are the risks of
learning a climate–malaria relationship from such a short record?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 3: click to reveal (try the questions first)

**3.1** Averaging to a province discards all within-province spatial variation:
elevation gradients, forest edges, urban vs rural. For a large or topographically
varied province, a single average may not represent the conditions where
transmission actually happens (often specific ecological niches), so the link to
cases is weakened. It further assumes _average_ temperature is what is important, 
whereas minimums or maximums may drive the mosquito population changes.

**3.2** Typically the within-year (seasonal) temperature swing within a province is
modest compared with, say, rainfall seasonality, and provinces differ from each
other too. Both the seasonal cycle and the between-province contrasts carry
information; the panel model in Part 5 uses both by standardising within province
and pooling across provinces. This is where checking maximum and minimum temperature 
variations would be informative, and may help you decide which to carry forward.

**3.3** A short record risks fitting to year-specific anomalies (an unusual season,
a one-off outbreak) rather than a stable climate–malaria relationship, and gives
very little power to separate climate effects from other drivers. Projections built
on such a fit should be treated as illustrative.
:::

---

# **Part 4:** Future climate projections (CMIP6)

::: speaker-note
This part summarises the CMIP6 multi-model ensemble to the same provinces, under
two scenarios. The key ideas for participants: (1) CMIP6 is an *ensemble* of many
climate models; we summarise across models with a mean and an interquartile band,
rather than trusting any single model; and (2) the two scenarios (SSP1-2.6 low
emissions, SSP5-8.5 high emissions) bracket a range of plausible futures. This is
a direct translation of the original Python, including the precipitation unit
conversion. It is the heaviest step in the document; on a slow machine give it a
minute, and note that `freeze: auto` caches it afterwards.
:::

The learned relationship (Part 5) is only useful if we can apply it to the future.
CMIP6 provides projections of temperature (`tas`) and precipitation (`pr`) from
many climate models, under different emissions scenarios. We summarise them to the
same provinces, under two scenarios: **SSP1-2.6** (low emissions) and **SSP5-8.5**
(high emissions).

```{r part4-read, echo=TRUE, eval=TRUE}
library(ncdf4)

cmip_dir <- file.path(data_dir, "cmip6", "monthly-averaged")
cmip_files <- list(
  ssp126_tas = "cmip6_regrided_model_ensemble.Thai_region.ssp126.tas.nc",
  ssp585_tas = "cmip6_regrided_model_ensemble.Thai_region.ssp585.tas.nc",
  ssp126_pr  = "cmip6_regrided_model_ensemble.Thai_region.ssp126.pr.nc",
  ssp585_pr  = "cmip6_regrided_model_ensemble.Thai_region.ssp585.pr.nc"
)

# Read one CMIP variable; dimensions are read by name so storage order does not matter.
read_cmip_var <- function(path, var, from_date = NULL) {
  nc  <- nc_open(path)
  on.exit(nc_close(nc))
  lon <- ncvar_get(nc, "lon"); lat <- ncvar_get(nc, "lat")
  tim <- ncvar_get(nc, "time")
  models <- ncvar_get(nc, "model")
  # time units -> dates
  tunits <- ncatt_get(nc, "time", "units")$value
  origin <- as.Date(sub(".*since ", "", tunits))
  dates  <- floor_date(origin + tim, "month")
  arr <- ncvar_get(nc, var)                       # dims in file order
  dn  <- sapply(nc$var[[var]]$dim, function(d) d$name)
  # reorder to lon, lat, time, model
  perm <- match(c("lon", "lat", "time", "model"), dn)
  arr <- aperm(arr, perm)
  # Trim to the reanalysis start so the fit and projection share a start month.
  if (!is.null(from_date)) {
    keep <- dates >= from_date
    arr  <- arr[, , keep, , drop = FALSE]
    dates <- dates[keep]
  }
  list(arr = arr, lon = lon, lat = lat, dates = dates, models = models)
}

cmip_start <- min(clim_months)   # reanalysis start month
tas126 <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp126_tas), "tas", cmip_start)
cat("CMIP6 grid:", length(tas126$lon), "lon x", length(tas126$lat), "lat;",
    length(tas126$models), "models;", length(tas126$dates), "months\n")
```

The next chunk defines the helpers that place each CMIP model on the reanalysis grid
and average it to provinces, keeping every model (used later for the projection
spread) alongside an ensemble summary.

```{r part4-summarise, echo=TRUE, eval=TRUE}
ref_grid <- tas_r[[1]]        # reanalysis grid (already cropped to region)

# Build a north-up, multi-layer SpatRaster from one model's [lon, lat, time] slice
# (lon/lat treated as cell centres, so the extent is padded half a cell each side).
slice_to_rast <- function(slc, lon, lat, ref) {
  lon_o <- order(lon); lat_o <- order(lat)          # index ascending
  slc <- slc[lon_o, lat_o, , drop = FALSE]          # lon asc, lat asc
  a <- aperm(slc, c(2, 1, 3))                        # -> lat x lon x time
  a <- a[dim(a)[1]:1, , , drop = FALSE]              # flip: north (max lat) = row 1
  r <- rast(a)                                       # 3rd dim -> layers
  lo <- sort(lon); la <- sort(lat)
  dx <- mean(diff(lo)); dy <- mean(diff(la))
  ext(r) <- c(min(lo) - dx/2, max(lo) + dx/2,
              min(la) - dy/2, max(la) + dy/2)
  crs(r) <- crs(ref)                                 # match the reanalysis grid
  r
}

# Every model -> provincial means, keeping the per-model detail (province x month x
# model) that the Part 6 projection needs for the spread.
cmip_provincial_bymodel <- function(cv, var, ref, provinces) {
  nm <- length(cv$models)
  # CMIP_MODEL_CAP env var caps the number of models (a faster/debug run); 0 = all.
  cap <- suppressWarnings(as.integer(Sys.getenv("CMIP_MODEL_CAP", "")))
  if (!is.na(cap) && cap > 0) nm <- min(nm, cap)
  out <- vector("list", nm)
  for (m in seq_len(nm)) {
    slc <- cv$arr[, , , m]                     # lon x lat x time
    r <- slice_to_rast(slc, cv$lon, cv$lat, ref)
    names(r) <- as.character(cv$dates)
    if (var == "pr") r <- r * 86400 / 1000     # flux (kg m-2 s-1) -> m (as in original)
    else             r <- r - 273.15           # Kelvin -> Celsius
    r <- resample(r, ref, method = "bilinear") # interpolate onto reanalysis grid
    pm <- region_means(r)                       # strict cell-centre assignment (as Part 3)
    out[[m]] <- pm[provinces, , drop = FALSE]
  }
  arr3 <- simplify2array(out)                  # province x month x model
  list(arr = arr3, dates = cv$dates, provinces = provinces)
}

# Ensemble summary (mean + interquartile band) from a per-model array.
summarise_cmip <- function(bm, scenario, var) {
  arr3 <- bm$arr; nt <- length(bm$dates)
  qf <- function(p) as.vector(apply(arr3, c(1, 2), quantile, probs = p, na.rm = TRUE))
  data.frame(
    province_en = rep(bm$provinces, times = nt),
    month       = rep(bm$dates, each = length(bm$provinces)),
    scenario = scenario, variable = var,
    mean = as.vector(apply(arr3, c(1, 2), mean, na.rm = TRUE)),
    q25 = qf(0.25), q75 = qf(0.75)
  )
}
```

::: {.todo-note}
The CMIP6 interpolation-and-averaging above is the single heaviest step and the
part most likely to need a check on first render (it reads a 45-model ensemble and
resamples it onto the reanalysis grid, exactly as the original Python did). If it
is too slow for a live session, the course team may wish to pre-compute the
provincial summaries once and cache them. It is left as a direct port here so the
method is transparent.
:::

Now apply those helpers to all four files (two variables and two scenarios), restrict
to the models the files share, and build the ensemble summary for plotting.

```{r part4-run, echo=TRUE, eval=TRUE}
tas585 <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp585_tas), "tas", cmip_start)
pr126  <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp126_pr),  "pr",  cmip_start)
pr585  <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp585_pr),  "pr",  cmip_start)

# Restrict all four files to their common models, in the same order (the projection
# pairs temperature and rainfall model-by-model).
common_models <- sort(Reduce(intersect,
  list(tas126$models, tas585$models, pr126$models, pr585$models)))
align_models <- function(cv) {
  idx <- match(common_models, cv$models)
  cv$arr    <- cv$arr[, , , idx, drop = FALSE]
  cv$models <- common_models
  cv
}
tas126 <- align_models(tas126); tas585 <- align_models(tas585)
pr126  <- align_models(pr126);  pr585  <- align_models(pr585)
cat("CMIP models in common across all four files:", length(common_models), "\n")

# Per-model provincial climate (kept for the Part 6 projection)
cmip_tas <- list(
  "SSP1-2.6" = cmip_provincial_bymodel(tas126, "tas", ref_grid, selected_provinces),
  "SSP5-8.5" = cmip_provincial_bymodel(tas585, "tas", ref_grid, selected_provinces))
cmip_pr <- list(
  "SSP1-2.6" = cmip_provincial_bymodel(pr126, "pr", ref_grid, selected_provinces),
  "SSP5-8.5" = cmip_provincial_bymodel(pr585, "pr", ref_grid, selected_provinces))

# Ensemble summary for plotting
cmip_summary <- bind_rows(
  summarise_cmip(cmip_tas[["SSP1-2.6"]], "SSP1-2.6", "tas"),
  summarise_cmip(cmip_tas[["SSP5-8.5"]], "SSP5-8.5", "tas"),
  summarise_cmip(cmip_pr[["SSP1-2.6"]],  "SSP1-2.6", "pr"),
  summarise_cmip(cmip_pr[["SSP5-8.5"]],  "SSP5-8.5", "pr")
)
```

Aggregate the ensemble summary to annual means and plot the two scenarios for one
example province.

```{r part4-plot, echo=TRUE, eval=TRUE, fig.width=11, fig.height=8}
example_province <- selected_provinces[1]

annual <- cmip_summary %>%
  mutate(year = year(month)) %>%
  filter(province_en == example_province) %>%
  group_by(province_en, scenario, variable, year) %>%
  summarise(mean = mean(mean), q25 = mean(q25), q75 = mean(q75), .groups = "drop")

lab <- c(tas = "temperature (°C)", pr = "total precipitation (m)")

print(ggplot(annual, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_y",
             labeller = labeller(variable = lab)) +
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = NULL,
       title = paste0("CMIP6 projections for ", example_province),
       subtitle = "Model mean (line) and interquartile range (band)") +
  theme_bw())
```

**4.1 The projections are shown as a model mean with an interquartile band. Why do
we show a *range* across models rather than picking the single "best" model?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**4.2 Compare the two scenarios (SSP1-2.6 and SSP5-8.5). How far apart are the
temperature projections by the end of the century, and how does the model agreement
on rainfall compare?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**4.3 We are about to feed these projections into a malaria model that was fitted
on only a couple of years of reanalysis. What assumption are we making about the
climate–malaria relationship when we extrapolate it decades into the future?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 4: click to reveal (try the questions first)

**4.1** Individual climate models disagree, especially on rainfall. The
ensemble mean is more reliable than any single model, and the spread (here the
interquartile band) is itself the honest expression of projection *uncertainty*.
Picking one "best" model hides that uncertainty.

**4.2** Temperature projections usually diverge clearly between the two scenarios
by late century, with SSP5-8.5 substantially warmer, and models tend to agree on
the direction of warming. Rainfall is far less certain: the inter-model band is
usually wide and the scenarios may overlap, reflecting genuine disagreement about
how precipitation will change.

**4.3** We are assuming the climate–malaria relationship learned over a couple of
recent years holds, unchanged, decades into the future, i.e. that no adaptation,
intervention, land-use change, drug resistance, or shift in the mosquito/parasite
system alters it. That is a strong assumption and the main caveat on any such
projection.
:::

---

# **Part 5:** Fitting climate to incidence

::: speaker-note
This is the core of the session: a statistical model relating climate to malaria
cases across all provinces at once (a "panel" model). The design choices to draw
out: (1) we model the *first difference* (month-to-month change) in cases rather
than the level, and we *standardise* per province, so that provinces of very
different sizes contribute comparably; (2) climate enters as temperature, rainfall,
and their squares, allowing a hump-shaped (optimum) response; and (3) the fitted
coefficients are *shared* across provinces, one climate–malaria relationship
applied everywhere. This is a direct translation of the original "panel method"
ordinary-least-squares fit. Frame it explicitly as the *simplest* model, one set
of coefficients for the whole country. Part 6 then pushes it forward under CMIP6 to
project future cases; the optional SEIRS section at the end of this session is the
more scientifically sound alternative (per-province transmission) for those who want
a mechanistic model.
:::

We now link climate to cases directly with a simple statistical model, pooling all
provinces. We use a **panel method**, the same approach used to learn the impact
of temperature and rainfall on the gross domestic product of the world's economies
([Burke, Hsiang & Miguel, *Nature*, 2015](https://www.nature.com/articles/nature15725){target="_blank"})
, but here applied to learn the impact of temperature and rainfall on the
month-to-month *change* in infection count. Because these simple panel methods
cannot account for trends, we first remove the trend by taking the first difference,
then standardise each province by its own mean and standard deviation over time.
Temperature and rainfall enter as themselves and squared, allowing a non-linear
(hump-shaped) response, with one shared relationship across provinces. Because we
model the first difference (the "integrated" step) and reconstruct the level by a
cumulative sum, this is an **ARIMA-style** model, more precisely an autoregressive
model with exogenous inputs (ARX), with temperature and rainfall as the exogenous
climate drivers (hence `arx_fit` below).

This is perhaps the most basic model one can build, and is designed to illustrate
the logic of integrating the climate data. It uses a single set of coefficients for
all provinces. In Part 6 we push this fitted model forward under the CMIP6 climate
to project future cases. The optional SEIRS section at the end of this session is
more scientifically sound (it has per-province transmission) and gives you a code
base for fitting a transmission model directly.

```{r part5-matrices, echo=TRUE, eval=TRUE}
provinces <- selected_provinces
months    <- sort(unique(clim_long$month))

# Case matrix: provinces x months, zero-filled where no cases were reported
cases_mat <- counts_prov_month %>%
  filter(province_en %in% provinces) %>%
  complete(province_en = provinces, month = months, fill = list(Count = 0)) %>%
  filter(month %in% months) %>%
  arrange(province_en, month) %>%
  pivot_wider(names_from = month, values_from = Count) %>%
  tibble::column_to_rownames("province_en") %>%
  as.matrix()
cases_mat <- cases_mat[provinces, , drop = FALSE]

# Climate matrices on the same grid
mat_from_long <- function(var) {
  clim_long %>%
    select(province_en, month, all_of(var)) %>%
    pivot_wider(names_from = month, values_from = all_of(var)) %>%
    tibble::column_to_rownames("province_en") %>%
    as.matrix() %>% `[`(provinces, , drop = FALSE)
}
clim_temp <- mat_from_long("temp_k")   # Kelvin (offset irrelevant once standardised)
clim_rain <- mat_from_long("rain_m")

cat("cases:", nrow(cases_mat), "provinces x", ncol(cases_mat), "months;",
    "total cases =", sum(cases_mat), "\n")
```

Standardise the response and the predictors within each province before pooling.

```{r part5-standardise, echo=TRUE, eval=TRUE}
# Population SD (divide by n, like NumPy's np.std), not R's sd() (n-1), so the fit
# reproduces the original.
popsd <- function(x) sqrt(mean((x - mean(x))^2))

# Response: standardised first difference of cases, per province
cases_fd <- cases_mat[, -1] - cases_mat[, -ncol(cases_mat)]
fd_mean  <- rowMeans(cases_fd)
fd_sd    <- apply(cases_fd, 1, popsd)
z_cases_fd <- (cases_fd - fd_mean) / fd_sd

# Predictors: standardised temperature and rainfall, per province
z_temp <- (clim_temp - rowMeans(clim_temp)) / apply(clim_temp, 1, popsd)
z_rain <- (clim_rain - rowMeans(clim_rain)) / apply(clim_rain, 1, popsd)
```

Fit the pooled least-squares panel, then rebuild fitted case counts by reversing the
standardisation and the first-differencing (a cumulative sum).

```{r part5-fit, echo=TRUE, eval=TRUE}
# Predictors: temperature, temperature^2, rainfall, rainfall^2 (drop the first month).
Y <- as.vector(z_cases_fd)
zt <- z_temp[, -1]; zr <- z_rain[, -1]
X  <- cbind(as.vector(zt), as.vector(zt)^2, as.vector(zr), as.vector(zr)^2)

# Centre (as in the original), solve, and predict
Y_mean <- mean(Y);            Y_prime <- Y - Y_mean
X_mean <- colMeans(X);        X_prime <- sweep(X, 2, X_mean)
beta   <- solve(t(X_prime) %*% X_prime, t(X_prime) %*% Y_prime)
Y_hat  <- as.vector(X_prime %*% beta) + Y_mean

colnames(X) <- c("temp", "temp^2", "rain", "rain^2")
print(round(setNames(as.vector(beta), colnames(X)), 3))

# Rebuild case counts: undo the standardisation and the first-differencing (cumsum).
z_hat <- matrix(Y_hat, nrow = nrow(z_cases_fd))
fitted_cases <- cases_mat * 0
for (i in seq_len(nrow(cases_mat))) {
  fitted_cases[i, 1]  <- cases_mat[i, 1]
  fitted_cases[i, -1] <- fitted_cases[i, 1] +
    cumsum(z_hat[i, ] * fd_sd[i] + fd_mean[i])
}
```

```{r verify-export, echo=FALSE, message=FALSE}
# Save this fit's actual outputs to the shared outputs/ folder, so they can be
# compared directly against the Python notebook's saved outputs (see
# python_workflow/compare.R). Instrumentation only: it exports the arrays already
# computed above, it does not recompute anything.
.odir <- file.path("..", "outputs")
dir.create(.odir, showWarnings = FALSE, recursive = TRUE)
jsonlite::write_json(list(
  selected   = provinces,
  months     = as.character(months),
  cases_mat  = cases_mat,
  clim_temp  = clim_temp,
  clim_rain  = clim_rain,
  panel_beta = as.vector(beta),
  arx_fit    = fitted_cases
), file.path(.odir, "r_panel.json"), matrix = "rowmajor", digits = NA)
```

Compare the observed and fitted monthly cases, province by province.

```{r part5-diag, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
fit_long <- bind_rows(
  as.data.frame(cases_mat) %>% tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>%
    mutate(series = "observed"),
  as.data.frame(fitted_cases) %>% tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>%
    mutate(series = "fitted")
) %>% mutate(month = as.Date(month))

print(ggplot(fit_long, aes(month, cases, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c(observed = "black", fitted = "red")) +
  labs(x = "date", y = "monthly cases",
       title = "Observed vs panel-fitted monthly cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

The fitted `beta` coefficients define the climate response. In **Part 6** we apply
them to the future CMIP6 climate to project cases forward.

**5.1 The model uses temperature, temperature², rainfall, and rainfall². What
shape of climate–malaria relationship do the squared terms allow that a
straight-line model could not? Why is that biologically sensible for a
mosquito-borne disease?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**5.2 We standardised cases and climate *per province* before pooling. What would
go wrong if we pooled the raw counts across provinces without standardising?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**5.3 Look at the observed-vs-fitted panels. Where does the model do well, and where
does it do poorly? What might it be missing?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**5.4 To project future cases, we would apply these fitted coefficients to the
CMIP6 climate from Part 4. List one thing you would want to check about the future
climate values before trusting such a projection.**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 5: click to reveal (try the questions first)

**5.1** The squared terms allow a *hump-shaped* (unimodal) response: cases rising
with temperature or rainfall up to an optimum and then falling. This is
biologically sensible because mosquito survival, biting, and parasite development
each have thermal optima, and both drought and flushing rains can reduce breeding,
so the response to climate is rarely monotonic.

**5.2** Provinces differ enormously in population and case numbers. Pooling raw
counts would let a few large-caseload provinces dominate the fit, and would confound
"more cases because bigger" with "more cases because of climate". Standardising per
province puts every province on a comparable footing so the shared coefficients
reflect the climate relationship, not province size.

**5.3** The model tends to capture the broad seasonal rise and fall but misses
sharp peaks, sudden outbreaks, and province-specific idiosyncrasies. Malaria is
also driven by non-climate factors: interventions (bed nets, spraying), human
movement and importation, drug resistance, health-system reporting changes, none
of which are in a climate-only model.

**5.4** Sensible checks include: are the future climate values within the range the
model was trained on (extrapolation beyond the training range is dangerous)? Are the
units and standardisation applied identically to the projections? Is the seasonal
structure of the projections realistic? Does the province averaging match?
:::

---

# **Part 6:** Projecting future cases (statistical model)

::: speaker-note
This is the pay-off: we take the model fitted on the historical period (Part 5) and
push it forward under the CMIP6 climate (Part 4) to project future cases. The
mechanics: standardise each future climate value using the *historical*
per-province mean and standard deviation (never re-standardise on the future data,
or you throw away the trend you are trying to project), apply the fitted `beta`, and
rebuild case counts. We run every climate model separately and show the ensemble
mean and interquartile band, exactly as we did for the raw climate in Part 4.

Be sure to make the key caveat explicit (see the warning box): reconstructing counts
by cumulatively summing predicted month-to-month changes over ~75 years behaves like
a random walk and will drift. This is a genuine limitation of projecting a
differenced/ARX model far ahead, and a large part of *why* one might prefer a bounded
mechanistic model (the optional SEIRS section at the end of this session) for long
horizons. Treat the projection as illustrative of method, not as a forecast.
:::

We now push the fitted model forward. For each climate model and scenario, we
standardise the future CMIP6 temperature and rainfall using the **historical**
per-province mean and standard deviation (the same scaling the model was fitted
with), apply the fitted coefficients to get the predicted month-to-month change in
cases, and rebuild counts by cumulative sum, seeding from the last observed month.

::: {.callout-warning}
## A projection, not a forecast
Rebuilding case counts by cumulatively summing predicted monthly *changes* over
decades behaves like a random walk and will drift, so the absolute numbers far into
the future should not be read literally. This is an inherent limitation of pushing a
first-difference (ARX) model a long way forward, and is one reason a bounded
mechanistic model (the optional SEIRS section at the end of this session) is often
preferred for long-range projection. Use this to see *how* climate projections feed
a fitted model, not as a prediction of future case counts.
:::

```{r part6-project, echo=TRUE, eval=TRUE}
# Historical per-province scaling (in the CMIP arrays' units). Standardising is
# offset-invariant, so this matches the scaling the model was fitted with in Part 5.
temp_mean_p <- rowMeans(clim_temp) - 273.15   # clim_temp is Kelvin; CMIP tas is °C
temp_sd_p   <- apply(clim_temp, 1, popsd)
rain_mean_p <- rowMeans(clim_rain)
rain_sd_p   <- apply(clim_rain, 1, popsd)

last_obs <- cases_mat[, ncol(cases_mat)]      # seed the projection at the last observed month
hist_end <- max(months)

# Apply the fitted ARX model to one climate model's future climate, rebuild counts.
project_one <- function(tas_pm, pr_pm, dates) {
  zt <- (tas_pm - temp_mean_p) / temp_sd_p    # province x month (vectors recycle per row)
  zr <- (pr_pm  - rain_mean_p) / rain_sd_p
  yhat <- (zt   - X_mean[1]) * beta[1] +      # predicted standardised first-difference
          (zt^2 - X_mean[2]) * beta[2] +
          (zr   - X_mean[3]) * beta[3] +
          (zr^2 - X_mean[4]) * beta[4] + Y_mean
  dcases <- yhat * fd_sd + fd_mean            # back to case-change units, per province
  fut <- which(dates > hist_end)             # project future months only
  proj <- last_obs + t(apply(dcases[, fut, drop = FALSE], 1, cumsum))
  list(proj = proj, dates = dates[fut])
}

# scenario -> array (province x future-month x model), dropping unstable models
project_scenario <- function(scenario) {
  tas <- cmip_tas[[scenario]]; pr <- cmip_pr[[scenario]]
  nm  <- dim(tas$arr)[3]
  lst <- lapply(seq_len(nm), function(m)
    project_one(tas$arr[, , m], pr$arr[, , m], tas$dates)$proj)
  arr  <- simplify2array(lst)                 # province x month x model
  keep <- apply(arr, 3, function(x) all(is.finite(x)))
  arr[, , keep, drop = FALSE]
}

fut_dates <- project_one(cmip_tas[[1]]$arr[, , 1], cmip_pr[[1]]$arr[, , 1],
                         cmip_tas[[1]]$dates)$dates
proj_arr  <- list("SSP1-2.6" = project_scenario("SSP1-2.6"),
                  "SSP5-8.5" = project_scenario("SSP5-8.5"))
```

Aggregate the projected months to annual totals per province, and plot the ensemble
mean and interquartile band for each scenario.

```{r part6-plot, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
annual_proj <- function(arr, dates, scenario) {
  yr <- lubridate::year(dates); years <- sort(unique(yr))
  ap <- sapply(years, function(y)
    apply(arr[, yr == y, , drop = FALSE], c(1, 3), sum))   # -> (province*?)...
  ap <- array(ap, dim = c(length(provinces), dim(arr)[3], length(years)))  # prov x model x year
  ap <- aperm(ap, c(1, 3, 2))                              # prov x year x model
  data.frame(
    province_en = rep(provinces, times = length(years)),
    year = rep(years, each = length(provinces)),
    scenario = scenario,
    mean = as.vector(apply(ap, c(1, 2), mean)),
    q25  = as.vector(apply(ap, c(1, 2), quantile, 0.25)),
    q75  = as.vector(apply(ap, c(1, 2), quantile, 0.75))
  )
}

proj_annual <- bind_rows(
  annual_proj(proj_arr[["SSP1-2.6"]], fut_dates, "SSP1-2.6"),
  annual_proj(proj_arr[["SSP5-8.5"]], fut_dates, "SSP5-8.5")
)

print(ggplot(proj_annual, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = "projected annual cases",
       title = "Projected annual cases under CMIP6 (statistical/ARX model)",
       subtitle = "Ensemble mean (line) and interquartile range (band). See caveat above.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

**6.1 The projection standardises the future climate using the *historical* mean and
standard deviation, not the future data's own. Why is that essential? What would go
wrong if we standardised the future climate on itself?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

**6.2 Look at the projected trajectories and re-read the caveat. Which features of
these plots would you trust, and which would you not? How would you describe the
result honestly to a decision-maker?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## Solutions for Part 6: click to reveal (try the questions first)

**6.1** The model was fitted on climate standardised by its *historical* mean and
standard deviation, so the coefficients only make sense on that same scale. If we
re-standardised the future climate on itself, we would subtract off the very warming
(or drying/wetting) trend we are trying to project; every period would look
"average" relative to its own era, and the projected climate signal would vanish. We
must apply the historical scaling so that a genuinely hotter future shows up as a
large positive standardised value.

**6.2** Trust the *direction and relative* differences, e.g. that the higher-
emissions scenario diverges from the lower one, and the ranking of provinces, more
than the absolute case numbers, which drift because of the cumulative-sum
reconstruction. Honest framing to a decision-maker: "this shows how a climate-driven
statistical model *responds* to two emissions futures; the shape and the gap between
scenarios are informative, but the absolute counts decades out are illustrative, not
a forecast, and a mechanistic model would be needed for credible long-range numbers."
:::

---

# Optional: a mechanistic SEIRS alternative

::: speaker-note
This whole section is optional and more advanced: it lives behind an expandable
box so it does not interrupt the main flow, and so Session 3 can stay a clean group
discussion. Run it only if the group powers through Parts 1–6 with time to spare, or
point interested participants to it afterwards. It reuses the case and climate data
already built above (no re-loading), fits a compartmental transmission model, and
doubles as a **worked code base for fitting a transmission model directly**. Keep the
optimiser budget small for a live demo: the fit is deliberately not run to
convergence, and every number is illustrative.
:::

::: {.callout-note collapse="true"}
## Why an SEIRS model? (optional and more advanced; click to expand)

The model in Parts 5–6 is deliberately the simplest thing that works: one shared
statistical relationship between climate and the month-to-month *change* in cases. A
mechanistic **SEIRS** model is the more biologically structured alternative. Instead
of regressing cases on climate, it represents the disease process itself: people
move through Susceptible → Exposed → Infected → Recovered → Susceptible compartments
, and lets **climate drive the transmission rate**. That buys three things the
statistical model cannot: per-province transmission with *shared* natural-history
parameters (so provinces borrow strength); built-in saturation and waning immunity
(so it cannot produce the negative or runaway counts a differenced model can); and a
template you can adapt to fit a transmission model directly to your own data. The
cost is more assumptions and a harder fit.

Each province runs an SEIRS model sharing the natural-history parameters but with a
province- and time-varying transmission rate:

$$\frac{dS}{dt} = -\beta(t) S I + \omega R \quad
  \frac{dE}{dt} = \beta(t) S I - \sigma E \quad
  \frac{dI}{dt} = \sigma E - \gamma I \quad
  \frac{dR}{dt} = \gamma I - \omega R$$

with $\beta_p(t) = \beta_{0,p}\exp(b_{temp}\, z_{temp}(p,t) + b_{rain}\, z_{rain}(p,t))$,
and reported cases modelled as
$\text{cases}_p(t) \sim \text{Poisson}(\text{scale}_p \cdot \sigma E_p(t))$, fitted by
maximum likelihood.

**Caveats: read before trusting any number.** The climate covariates stand in for
missing mosquito data; there are no population denominators (so `scale` absorbs
population size and reporting rate); the natural-history parameters
($\sigma,\gamma,\omega$) are fitted freely rather than fixed to published
vivax/falciparum biology; the two parasite species are pooled; and the optimiser uses
a small iteration budget so it runs quickly; it is **not** run to convergence. Treat
every output as illustrative of method, not as an estimate.

**The data.** The SEIRS reuses the case matrix and per-province climate already built
in Parts 2–3 and 5. The only difference is that it standardises climate **globally**
(one shared scale) rather than per province, so that `b_temp`/`b_rain` are a single
comparable climate sensitivity.

```{r seirs-setup, echo=TRUE, eval=TRUE}
# reuse cases_mat, clim_temp (Kelvin), clim_rain (m), popsd() from Part 5
ztemp <- (clim_temp - mean(clim_temp)) / popsd(clim_temp)
zrain <- (clim_rain - mean(clim_rain)) / popsd(clim_rain)
cat(sprintf("SEIRS inputs: %d provinces x %d months, %.0f total cases\n",
            nrow(cases_mat), ncol(cases_mat), sum(cases_mat)))
```

The model is built up in five small pieces below, then fitted. Read them in order.

**Step 1: the transmission dynamics (the ODE right-hand side).** This is the SEIRS
model itself: the instantaneous rate of change of each compartment. `state` is a
5 × n matrix, one column per province, rows S, E, I, R, and C (a running count of
new infections), so everything is vectorised over provinces at once.

```{r seirs-rhs, echo=TRUE, eval=TRUE}
SUBSTEPS_PER_MONTH <- 20   # RK4 sub-steps per month

seirs_rhs <- function(state, beta, sigma, gamma, omega) {
  S <- state[1, ]; E <- state[2, ]; I <- state[3, ]; R <- state[4, ]
  rbind(-beta * S * I + omega * R,     #  dS  (lose to infection, gain from waning immunity)
         beta * S * I - sigma * E,     #  dE  (newly infected, then become infectious)
         sigma * E   - gamma * I,      #  dI  (become infectious, then recover)
         gamma * I   - omega * R,      #  dR  (recover, then lose immunity)
         sigma * E)                    #  dC  (cumulative NEW infections)
}
```

**Step 2: simulating it forward (fixed-step RK4).** We integrate the ODE month by
month using a fixed-step Runge–Kutta (RK4) scheme with `SUBSTEPS_PER_MONTH` steps per
month, for all provinces simultaneously. A burn-in lets the model settle before we
record anything, and we reset the accumulator `C` at the start of each month so that
`incidence` holds the new infections *within* that month.

```{r seirs-sim, echo=TRUE, eval=TRUE}
simulate_incidence <- function(beta, sigma, gamma, omega,
                               i0 = 1e-4, e0 = 1e-4, burn_in_months = 12) {
  n_prov <- nrow(beta); n_months <- ncol(beta); dt <- 1 / SUBSTEPS_PER_MONTH
  state <- rbind(rep(1 - i0 - e0, n_prov), rep(e0, n_prov),
                 rep(i0, n_prov), rep(0, n_prov), rep(0, n_prov))
  step <- function(state, beta_m) {                    # one RK4 sub-step
    k1 <- seirs_rhs(state,              beta_m, sigma, gamma, omega)
    k2 <- seirs_rhs(state + 0.5*dt*k1,  beta_m, sigma, gamma, omega)
    k3 <- seirs_rhs(state + 0.5*dt*k2,  beta_m, sigma, gamma, omega)
    k4 <- seirs_rhs(state + dt*k3,      beta_m, sigma, gamma, omega)
    state + (dt/6) * (k1 + 2*k2 + 2*k3 + k4)
  }
  if (burn_in_months > 0) {                            # settle using month-1 transmission
    beta0_m <- beta[, 1]
    for (b in seq_len(burn_in_months)) {
      for (s in seq_len(SUBSTEPS_PER_MONTH)) state <- step(state, beta0_m)
      state[5, ] <- 0
    }
  }
  incidence <- matrix(0, n_prov, n_months)
  for (m in seq_len(n_months)) {
    state[5, ] <- 0; beta_m <- beta[, m]               # reset accumulator; use this month's beta
    for (s in seq_len(SUBSTEPS_PER_MONTH)) state <- step(state, beta_m)
    incidence[, m] <- state[5, ]
    state[1:4, ] <- pmin(pmax(state[1:4, ], 0), 1)     # keep compartments in [0, 1]
  }
  incidence
}
```

**Step 3: unpacking the parameters.** The optimiser works in an unconstrained,
log-transformed space so that rates stay positive. This turns the flat parameter
vector back into named, natural-scale quantities: the shared natural-history rates,
the two climate sensitivities, and the per-province baseline transmission `beta0` and
reporting `scale`.

```{r seirs-unpack, echo=TRUE, eval=TRUE}
unpack_params <- function(params, n) {
  list(sigma = exp(params[1]), gamma = exp(params[2]), omega = exp(params[3]),
       b_temp = params[4], b_rain = params[5],
       beta0 = exp(params[6:(5 + n)]), scale = exp(params[(6 + n):(5 + 2*n)]))
}
```

**Step 4: the likelihood.** Given a parameter vector, build the climate-forced
transmission rate `beta_t`, simulate the incidence, and score it against the observed
counts with a Poisson negative log-likelihood, the quantity the optimiser minimises.

```{r seirs-nll, echo=TRUE, eval=TRUE}
neg_log_lik <- function(params, cases, ztemp, zrain) {
  n <- nrow(cases); p <- unpack_params(params, n)
  beta_t <- p$beta0 * exp(p$b_temp * ztemp + p$b_rain * zrain)   # climate forces transmission
  beta_t <- pmin(pmax(beta_t, 1e-6), 50)
  incidence <- simulate_incidence(beta_t, p$sigma, p$gamma, p$omega)
  lam <- pmax(p$scale * incidence, 1e-8)                         # expected reported cases
  sum(lam - cases * log(lam))   # Poisson NLL (constant omitted)
}
```

**Step 5: sensible starting values.** A reasonable initial guess for the optimiser:
incubation ~0.5 months, infectious ~1 month, immunity ~6 months, no climate effect to
begin with, and per-province baselines scaled to each province's mean caseload.

```{r seirs-x0, echo=TRUE, eval=TRUE}
default_x0 <- function(cases) {
  n <- nrow(cases)
  c(log(1/0.5), log(1/1.0), log(1/6.0), 0, 0,      # sigma, gamma, omega, b_temp, b_rain
    log(rep(2.5, n)),                              # beta0 (per province)
    log(pmax(rowMeans(cases), 1) * 50))            # scale (per province)
}
```

**Step 6: fit.** Now minimise the negative log-likelihood with a bounded optimiser
(`L-BFGS-B`). The `lower`/`upper` vectors keep every parameter in a biologically
plausible range (incubation/infectious periods between ~0.05 and 3 months, immunity
between 0.5 and 36 months, the climate sensitivities within ±2), and `maxit` is kept
deliberately small so it finishes quickly (raise it for a real fit).

```{r seirs-fit, echo=TRUE, eval=TRUE}
n <- nrow(cases_mat)
x0 <- default_x0(cases_mat)
lower <- c(log(1/3.0), log(1/3.0), log(1/36.0), -2, -2,
           rep(log(0.05), n), rep(log(1.0),  n))
upper <- c(log(1/0.05), log(1/0.05), log(1/0.5), 2, 2,
           rep(log(30),   n), rep(log(1e9),  n))

# maxit small so it runs quickly (raise for a real fit). ndeps matches SciPy's default
# gradient step, so this tracks the notebook as closely as the two L-BFGS-B allow.
fit <- optim(x0, neg_log_lik, method = "L-BFGS-B",
             lower = lower, upper = upper,
             cases = cases_mat, ztemp = ztemp, zrain = zrain,
             control = list(maxit = 20, ndeps = rep(1e-8, length(x0)))) 
p <- unpack_params(fit$par, n)

cat(sprintf("convergence code=%d  nll=%.1f\n", fit$convergence, fit$value))
cat(sprintf("incubation period 1/sigma = %.2f months\n", 1/p$sigma))
cat(sprintf("infectious period 1/gamma = %.2f months\n", 1/p$gamma))
cat(sprintf("immunity duration 1/omega = %.2f months\n", 1/p$omega))
cat(sprintf("climate sensitivity b_temp = %+.3f  b_rain = %+.3f\n", p$b_temp, p$b_rain))
```

**Step 7: diagnostics.** Finally, rebuild the fitted incidence from the estimated
parameters and compare it, province by province, against the observed monthly counts.

```{r seirs-diag, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
beta_fit <- p$beta0 * exp(p$b_temp * ztemp + p$b_rain * zrain)
lam_fit  <- p$scale * simulate_incidence(beta_fit, p$sigma, p$gamma, p$omega)

ss_res <- sum((cases_mat - lam_fit)^2)
ss_tot <- sum((cases_mat - mean(cases_mat))^2)
cat(sprintf("pseudo-R2 (vs fitted lambda): %.3f\n", 1 - ss_res / ss_tot))

diag_long <- bind_rows(
  as.data.frame(cases_mat) %>% tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>% mutate(series = "observed"),
  as.data.frame(lam_fit) %>% `rownames<-`(rownames(cases_mat)) %>%
    `colnames<-`(colnames(cases_mat)) %>%
    tibble::rownames_to_column("province_en") %>%
    pivot_longer(-province_en, names_to = "month", values_to = "cases") %>% mutate(series = "fitted")
) %>% mutate(month = as.Date(month))

print(ggplot(diag_long, aes(month, cases, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c(observed = "black", fitted = "red")) +
  labs(x = "date", y = "monthly cases",
       title = "Observed vs SEIRS-fitted monthly cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

```{r seirs-export, echo=FALSE, message=FALSE}
# Save the fitted SEIRS outputs for the direct comparison (python_workflow/compare.R).
.odir <- file.path("..", "outputs"); dir.create(.odir, showWarnings = FALSE, recursive = TRUE)
jsonlite::write_json(list(
  selected = provinces, months = as.character(months),
  nll = fit$value,
  sigma = p$sigma, gamma = p$gamma, omega = p$omega,
  b_temp = p$b_temp, b_rain = p$b_rain,
  beta0 = p$beta0, scale = p$scale,
  lam_fit = lam_fit
), file.path(.odir, "r_seirs.json"), matrix = "rowmajor", digits = NA, auto_unbox = TRUE)
```

**A note on identifiability.** Fitting this exact model in both R and Python is
instructive. `beta0` (baseline transmission) and `scale` (population × reporting rate)
trade off to set the case *level*, so the optimiser can shuffle magnitude between
them, and the natural-history rates ($\sigma,\gamma,\omega$) are only weakly
constrained by the data. As a result the two implementations can reach similar
likelihoods with noticeably different individual parameters, a textbook symptom of
**non-identifiability**. The practical lesson, and the warning above made concrete:
trust the model's *predictions* and the climate sensitivities `b_temp`/`b_rain`, not
the individual parameter values. This is why the verification in
`python_workflow/compare.R` reports the fitted incidence and climate coefficients and
flags the level parameters as weakly identified, rather than expecting `beta0`/`scale`
to match; run it to see the actual R-versus-Python differences for your fit.

**Discussion (SEIRS): the statistical model (Parts 5–6) and this mechanistic model
both link climate to malaria. What can the mechanistic model tell you that the
statistical one cannot, and vice versa? And why are the fitted natural-history
parameters ($\sigma,\gamma,\omega$) a weakness here, and how would you fix it?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>
