---
title: "Obtaining, manipulating, and fitting climate data"
subtitle: "SPARKLE Session 2 on climate data in infectious disease modelling"
format:
  training-html: default
presenter: "Michael & Roslyn"
contributors:
  - "Dr Vassili Kitsios"
  - "Prof. Roslyn Hickson"
  - "Dr Michael Lydeamore"
  - "Claude (Anthropic) did the porting of the existing Python workflow to R Quarto; all content checked by the course team"
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE,
  results   = "hold"   # show the whole code chunk first, then all its output
)
```

```{css, echo=FALSE}
.speaker-note {
    background-color: #fff8e1;
    border-left: 4px solid #f9a825;
    padding: 10px 14px;
    margin: 14px 0;
    border-radius: 3px;
    font-size: 0.92em;
    color: #5d4037;
}
.speaker-note::before {
    content: "🗒 Facilitator note: ";
    font-weight: bold;
}
.todo-note {
    background-color: #fdecea;
    border-left: 4px solid #d32f2f;
    padding: 10px 14px;
    margin: 14px 0;
    border-radius: 3px;
    font-size: 0.92em;
    color: #611a15;
}
.todo-note::before {
    content: "🚧 TODO (course team): ";
    font-weight: bold;
}
/* Wide printed output (e.g. glimpse, data-frame previews) scrolls sideways
   instead of wrapping, similar to a pandas data-frame preview. */
.cell-output-stdout pre {
    white-space: pre;      /* do not wrap long lines */
    overflow-x: auto;      /* scroll right instead */
}
/* Wide rendered tables (e.g. kable data-frame previews) scroll sideways too. */
.cell-output-display {
    overflow-x: auto;
}
```

::: speaker-note
**Environment.** The R practicals assume the following packages are installed:
`jsonlite`, `dplyr`, `tidyr`, `lubridate`, `ggplot2`, `patchwork`, `sf`,
`terra`, and `ncdf4`. Please confirm these are installed before the session. The
geospatial steps (reanalysis and CMIP6) read netCDF and shapefiles directly and
can take a little while on first render. Using `freeze: auto` caches them
after the first successful render.

**Worked example.** Throughout, the running example is malaria incidence in
Thailand, linked to ERA5 reanalysis climate and CMIP6 future projections.
:::

# Summary {.tabset .tabset-pills .tabset-fade}

This session works through the practical workflow of using climate data in
infectious disease modelling, with malaria in Thailand as the case study: from
obtaining raw data through to a fitted statistical model that links climate to
malaria incidence, with future projections.

The session opens by framing the **research question** and then works through 
six practical parts, with an extension section on fitting an SEIRS transmission 
model:

**The research question.** What we are trying to answer, and how it informs what
data and modelling we need.

**Part 1: Obtaining the data.** Where the data come from (malaria surveillance, 
climate reanalysis, future projections, administrative boundaries) and how to 
obtain them.

**Part 2: Malaria case data.** Reading the surveillance line list and 
summarising cases by month, province, and district.

**Part 3: Environmental variables (reanalysis).** Linking gridded ERA5 climate
data to administrative units using a shapefile, and building per-province 
monthly temperature and rainfall series.

**Part 4: Future climate projections.** Summarising CMIP6 projections under two
emissions scenarios (SSP1-2.6 and SSP5-8.5) for the same provinces.

**Part 5: Fitting climate to incidence.** A statistical (ARIMA/ARX panel) model
relating standardised climate to changes in malaria cases across provinces.

**Part 6: Projecting future cases.** Pushing the fitted statistical model 
forward under the CMIP6 scenarios to project future cases, with the important 
caveats of long-range extrapolation.

**Optional extension: An SEIRS model.** A (basic) transmission model approach.

::: {.callout-tip}
We recommend saving your written answers at the end of each part by printing the
page (there is a **Print this page** button at the end of each part).
:::

::: speaker-note
**Running this in 90 minutes.** The full workflow with every question is a 3-4 hour
hands-on session. To fit 90 minutes: pre-render the document beforehand (so the CMIP6
and projection steps are cached and there is no live compute wait), drive the code from
the front while participants watch and discuss, and pause on just the one **core
question** per part flagged below. The other questions stay in the document with their
revealed answers, for self-study afterwards. The optional SEIRS extension is not part of
the 90-minute run.

Suggested time budget:

- Research question + setup: 5 min
- Part 1 (obtaining data): 10 min, core question **1.2**
- Part 2 (case data): 12 min, core question **2.2**
- Part 3 (reanalysis + grid-to-province): 15 min, core question **3.1**
- Part 4 (CMIP6 projections): 12 min, core question **4.1**
- Part 5 (fitting climate to incidence): 20 min, core question **5.2**
- Part 6 (projecting future cases): 12 min, core question is the "does this projection look realistic?" one
- Wrap-up and signpost the optional SEIRS: 4 min
:::

This is a port of the python analysis
(`python_workflow/malaria_workflow.ipynb`), which we have left available for 
those interested.

---

# The research question

::: speaker-note
Spend the first few minutes establishing *what we are trying to answer* and 
*what decision it would inform*. Draw out that the question is what tells us 
(a) whether climate data is even relevant, (b) which data we need, and (c) 
whether it makes sense to project forward. Participants do this same first step 
for their own work in Session 3.
:::

It is important in modelling for the question to determine the methodology, and 
it also informs what data we need. The data available may in turn force us to 
update and refine our question and methodology.

For this case study, our research question is:

> **How do temperature and rainfall relate to malaria incidence across Thailand, 
> and how might that incidence change under future emissions scenarios?**

Reading that question tells us:

- a measure of the **outcome**: malaria cases, over time and by location.
- the **drivers** we hypothesise to matter: temperature and rainfall.
- a **spatial region**: Thailand. The spatial resolution is often driven by 
  data availability. In this case, we will use provinces (the unit cases are 
  reported in).
- a **historical period** where we observe outcome and drivers together, to \
  *learn* the relationship. This is also often driven by data availability.
- **future climate** for the same region, to *project* that relationship 
  forward.

---

# **Part 1:** Obtaining the data

::: speaker-note
This part is about *where the data come from* and *why each source was chosen*. 
:::

The workflow draws on four data sources. Three are used directly in the 
analysis; the fourth (mosquito occurrence) motivates the approach but is not 
currently used. The table below is the quick reference; each source, and why it 
was chosen, is expanded underneath.

| # | Data | What it is / role | Why this source | Source |
|---|--------|-------------------|-----------------|-------------|
| 1 | Malaria cases | national surveillance line list (the outcome) | authoritative, national, open, right resolution | [Thai DDC](https://malaria-api.ddc-malaria.org/study/2year/calendar){target="_blank"} |
| 2 | Spatial region boundaries | province (ADM1) polygons; aggregate grid to province | official ADM1; names align with the case data | [Open Dev Mekong](https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries){target="_blank"} |
| 3 | ERA5 reanalysis | gap-free past climate, used to *learn* | state of the art, ~0.25°, complete coverage, open | [Copernicus CDS](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download){target="_blank"} |
| 4 | CMIP6 projections | multi-model future climate, used to *project* | latest coordinated ensemble; scenarios bracket futures | [Copernicus CDS](https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download){target="_blank"} |

This list is the data we found available, not the initial list assembled. 

**0.1 What data do you think is missing from this list, and why?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

<p><input type="button" value="Print this page" onClick="window.print()"></p>

::: {.callout-tip collapse="true"}
## 0.1 Suggested answer: click to reveal (try to answer yourself first)

There is no single right answer.

We were initially planning on relating mosquito vector abundance to the climatic
variables, and projecting those forward and linking to malaria incidence via a 
transmission model tracking humans and vectors. But we couldn't find abundance 
data in Thailand at the same time period as the malaria incidence data.
:::

::: {.callout-tip}
Reanalysis (ERA5) is used to *learn* the relationship between climate and malaria
over the historical period; the CMIP6 projections are used to *project* that
relationship forward under future emissions scenarios.
:::

**1. Malaria case data (surveillance line list).** Individual records of malaria
blood draws in Thailand, with date, province, district, and result (*P. vivax* `V`,
*P. falciparum* `F`, or other). This is the outcome we are modelling, and it comes
from the national surveillance system (Thailand's DDC), so it is authoritative and
consistent across the whole country. It is openly queryable via the API (no
data-sharing agreement for this aggregated endpoint) and carries date, province,
district and species per record, at the spatial and temporal resolution we need for
monthly climate. Alternatives like WHO country-year totals are too coarse, and
hospital records are not openly accessible.

*Source and files:* [https://malaria-api.ddc-malaria.org/study/2year/calendar](https://malaria-api.ddc-malaria.org/study/2year/calendar){target="_blank"} , saved as `data/malaria/calendar.json`.

::: {.callout-note appearance="minimal"}
**DDC** is Thailand's Department of Disease Control (its vector-borne disease division runs the malaria surveillance).
:::

**2. Administrative boundaries (shape files).** Province boundaries for Thailand. We
need province polygons because cases are reported per province but climate is
gridded; the polygons let us aggregate the grid to the reporting unit. This set was
chosen because it is the official Thai ADM1 (province) boundaries, so its units and
names align with how the surveillance data is reported (name alignment matters: it
is why some provinces drop out when names do not match), and it is freely available
in a standard GIS format.

*Source and files:* [https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries](https://data.opendevelopmentmekong.net/dataset/thailand-provincial-boundaries){target="_blank"} , saved under `data/shape_files/tha_admbnda_adm1_rtsd_20190221/`.

::: {.callout-note appearance="minimal"}
A **shape file** stores polygon boundaries (countries, provinces, districts); **ADM1** is the first sub-national administrative level, here the provinces.
:::

**3. Climate reanalysis (ERA5).** Gridded, monthly-averaged 2 m temperature (`2t`)
and total precipitation (`tp`): a physically consistent, gap-free reconstruction of
past weather, used here to *learn* how climate relates to malaria. We use reanalysis
rather than weather stations because stations are sparse and gappy, whereas
reanalysis gives complete fields everywhere, which is essential for giving every
province a climate value. We use ERA5 specifically because it is the current
state-of-the-art global reanalysis (~0.25° resolution, well validated, openly
available), carrying the variables we want (temperature and precipitation drive
mosquito abundance and survival) at monthly resolution. It is the de facto standard
for climate and health work.

*Source and files:* [https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=download){target="_blank"} , saved under `data/reanalysis/monthly-averaged/`. The required subset is already curated for you; other variables can be downloaded freely from the Copernicus Climate Change Service (C3S) Climate Data Store.

::: {.callout-note appearance="minimal"}
**ERA5** = ECMWF Re-Analysis, 5th generation (ECMWF is the European Centre for Medium-Range Weather Forecasts; the lineage runs ERA-15, ERA-40, ERA-Interim, ERA5).
:::

**4. Future climate projections (CMIP6).** A multi-model ensemble of monthly
temperature (`tas`) and precipitation (`pr`) for Thailand under two emissions
scenarios (SSP1-2.6 and SSP5-8.5), used to *project* the learned relationship into
the future. We use CMIP6 specifically because it is the latest coordinated
multi-model generation (the models underpinning IPCC AR6), delivered in a standard,
comparable format, so we can use a multi-model ensemble to represent model
uncertainty and the SSP scenarios to bracket futures. SSP1-2.6 and SSP5-8.5 were
chosen as a low- and high-emissions pair to show sensitivity to the pathway. It is
also regriddable onto the ERA5 grid, so the historical and future steps stay
consistent.

*Source and files:* [https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download](https://cds.climate.copernicus.eu/datasets/projections-cmip6?tab=download){target="_blank"} , saved under `data/cmip6/monthly-averaged/`. The data is curated and trimmed to the region for this case study; other variables, scenarios, or the full globe are available from the same Climate Data Store.

::: {.callout-note appearance="minimal"}
**CMIP6** = Coupled Model Intercomparison Project, 6th phase ("coupled" refers to coupled atmosphere and ocean models). In an **SSP** (Shared Socioeconomic Pathway) label such as SSP1-2.6, the first digit is the socioeconomic story (1 = sustainability, 5 = fossil-fuel-intensive) and the second is the resulting radiative forcing (2.6 vs 8.5 W/m² by 2100, i.e. low vs high warming).
:::

**Mosquito occurrence (GBIF), considered but not used.** Mosquito occurrence records
for Thailand. We explored GBIF because it is the largest aggregator of open occurrence
records, so it is the obvious first place to look for vector data. It is not used here
because it gives *occurrence*, not *abundance*, and has no samples after 2022, so it
does not overlap the 2025 to 2026 malaria data. For now we relate malaria incidence in
humans to environmental variables directly (the approach taken in Part 5); regardless,
this is a good, straightforward way to test whether environmental variables such as
these are affecting the health outcomes of interest, and which ones. A natural
extension is to obtain overlapping mosquito data, learn the influence of temperature on
the mosquito population, then the impact of that population on malaria infections via a
transmission model.

*Source and files:* [https://www.gbif.org/dataset/db4211d6-b480-4bc5-9202-87f2072c4b12](https://www.gbif.org/dataset/db4211d6-b480-4bc5-9202-87f2072c4b12){target="_blank"} (not downloaded; not part of the analysis).

::: {.callout-note appearance="minimal"}
**GBIF** is the Global Biodiversity Information Facility, the largest aggregator 
of open species-occurrence records.
:::

::: {.callout-note}
## Obtaining the data yourself
The Copernicus Climate Data Store (ERA5 and CMIP6) requires a free account and,
for bulk downloads, the CDS API (a `~/.cdsapirc` key file). Because these
downloads need accounts/API keys and can be large, they are not run as code 
here. Instead, the curated subset needed for this tutorial is provided 
separately in the `data/` folder at the repository root.
:::

## Ideal vs achievable workflows

::: speaker-note
The key discussion point here: what you *want* to do is often constrained by the
data you can actually get. 
:::

The right workflow is rarely the most sophisticated one, but one that fits 
both your question and that your data can actually support. For malaria, the 
biologically known climate-sensitive link is the *mosquito* population: 
temperature and breeding site available, typically via rainfall, drive mosquito 
abundance, which in turn drives transmission. 

| Approach | Pipeline | Data needed | Biological fidelity |
|---|---|---|---|
| **Ideal** (vector-explicit) | climate → mosquito abundance → transmission model → incidence | overlapping spatio-temporal local mosquito **abundance** data *and* climate | highest |
| **Simplest** (direct statistical) | climate → incidence | overlapping spatio-temporal case data + climate | lowest: no explicit mechanism |
| **Achievable here** (climate-forced SEIRS) | climate → transmission model (transmission forced by climate) → incidence | overlapping spatio-temporal case data + climate | medium: climate link through implicit vector dynamics |

For this example we wanted the **ideal** path, but we couldn't find mosquito 
abundance data that spatio-temporally overlapped the malaria period (see the 
GBIF note above), so we use the lower two tiers: a direct statistical model 
(Part 5) and a climate-forced SEIRS model (the optional section at the end of 
this session). Both skip the mosquito layer. This is a data-driven compromise.

**1.1 For each of the four data sources above, what alternative dataset or
approach could you use in its place? Think about substitutes for ERA5, for
CMIP6, and for the provincial case and boundary data, and what you would trade
off by switching. If you identified other data earlier, what sources are there?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 1.1 Suggested answer: click to reveal (try it first)
There are alternatives for every source, each with trade-offs. For the
reanalysis, other products such as MERRA-2 or JRA-55, or gridded observational
datasets (for example CHIRPS for rainfall, CRU for temperature), could replace
ERA5, trading spatial and temporal resolution and which variables are available.
For the projections, CMIP5 or a downscaled or emulated ensemble could replace
CMIP6, changing resolution, the scenario definitions, and the model spread. For
the boundaries, a different administrative product or a gridded population layer
could be used, changing which spatial unit cases are aggregated to. For the case
data, a different health outcome or a modelled incidence surface could stand in
where line-list surveillance is not available. The point is to notice that each
swap changes resolution, coverage, or assumptions rather than to find one right
answer.
:::

::: speaker-note
Core question for the 90-minute run; pause here. The other Part 1 questions are for self-study.
:::

**1.2 Reanalysis is used to *learn* the climate–disease relationship and CMIP6
projections are used to *project* it into the future. Why can we not simply use
the future projections to learn the relationship in the first place?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 1.2 Suggested answer: click to reveal (try it first)
Future projections are, by construction, *projections*: they describe
plausible futures, not observed weather, and there is no malaria data from the
future to pair them with. We can only learn a relationship where we observe
*both* climate and cases together, which means the historical/reanalysis period.
The projections are then used to *project* that learned relationship forward.
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>

---

# **Part 2:** Malaria case data

::: speaker-note
This part reads the surveillance line list and aggregates it. The key teaching
points are: 

(1) surveillance data are messy and unevenly reported across provinces and time; 
this is why we later restrict to provinces with enough data; and 

(2) aggregating to monthly counts per province gives us the response variable
that we will link to climate. 

The code is a direct translation of the Python `pandas` workflow. Have 
participants run it and *consider* the province panels (question 2.2) before 
moving on.
:::

We start from the surveillance line list: one row per blood draw, with a date,
province, district, and result. We read it, parse the dates, and aggregate to
monthly counts.

First read in the data from the json file.
```{r part2-read, echo=TRUE, eval=TRUE}
library(jsonlite)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(here)

# Locate the data
data_dir <- here::here("data")

# Read the JSON line list and parse dates ---------------------------------
df <- fromJSON(file.path(data_dir, "malaria", "calendar.json")) |>
  mutate(
    date_daily = as.Date(as.character(blood_draw_date), format = "%Y%m%d"),
    month      = floor_date(date_daily, "month")
  )

# Preview what we have loaded ---------------------------------------------
df |>
  slice_head(n = 5) |>
  knitr::kable()
```

There are `r n_distinct(df$province_en)` unique provinces, `r n_distinct(df$district_en)` unique districts, and `r n_distinct(df$blood_draw_date)` unique blood draw dates in the data.

**2.4 The results column pools *P. vivax* and *P. falciparum* together. Why might
it matter, for a climate-driven analysis, that these two species are combined?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 2.4 Suggested answer: click to reveal (try it first)
*P. vivax* and *P. falciparum* differ in incubation period, in relapse
behaviour (vivax can relapse from liver-stage hypnozoites months later), and
likely in their climate sensitivity. But they have similar mosquito vectors. So
whether or not pooling them means blurring the climate-malaria signal depends on
how climate-related those factors are, and whether the mosquito relationship
with climate would dominate. A potential future direction to explore.
:::

Counts for Thailand per month and per day
```{r part2-counts, echo=TRUE, eval=TRUE, fig.width=10, fig.height=6}
# Total counts per day and per month --------------------------------------
counts_per_day <- df |>
  count(date_daily, name = "count") |>
  arrange(date_daily)
counts_per_month <- df |>
  count(month, name = "count") |>
  arrange(month)

p_day <- ggplot(counts_per_day, aes(date_daily, count)) +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(x = "date", y = "total daily counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_month <- ggplot(counts_per_month, aes(month, count)) +
  geom_line() + geom_point(size = 1) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %Y") +
  labs(x = "date (monthly total, aligned to first of month)",
       y = "total monthly counts") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_day / p_month
```

Counts per province per month: since we will be building models estimating the infection counts per province per month, including the impact of the climate.
```{r part2-province, echo=TRUE, eval=TRUE, fig.width=11, fig.height=11}
# Counts per province per month -------------------------------------------
counts_prov_month <- df |>
  count(month, province_en, name = "Count") |>
  arrange(month, province_en)

# Keep provinces with more than 12 months of reported cases 
selected_provinces <- counts_prov_month |>
  add_count(province_en, name = "n_months") |>
  filter(n_months > 12) |>
  distinct(province_en) |>
  pull(province_en)
cat("Selected provinces:", length(selected_provinces), "\n")

# Facet titles: how many of the possible months each province reports ---------
n_possible <- n_distinct(counts_prov_month$month)     # total months in the data (18)
prov_counts <- counts_prov_month |>
  filter(province_en %in% selected_provinces) |>
  group_by(province_en) |>
  summarise(n_months = n_distinct(month), .groups = "drop")
prov_lab <- prov_counts |>
  transmute(
    province_en,
    label = paste0(province_en, " (", n_months, "/", n_possible, ")")
  ) |>
  tibble::deframe()

print(counts_prov_month |>
  filter(province_en %in% selected_provinces) |>
  ggplot(aes(month, Count)) +
  geom_line() + geom_point(size = 0.6) +
  facet_wrap(~ province_en, scales = "free_y",
             labeller = labeller(province_en = prov_lab)) +
  labs(x = "date", y = "monthly count") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

**2.3 Look at the per-province panels. Describe the seasonal pattern you see. Do
all provinces peak at the same time of year? What might explain differences
between provinces?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 2.3 Suggested answer: click to reveal (try it first)
Most provinces show a seasonal rise and fall, but not all peak in the same
month; malaria seasonality depends on local climate, elevation, land use, and
the timing of the rains, which differ across the country. Border provinces and
forested provinces often behave differently from central ones.
:::

The natural next question is whether we could work at a finer spatial scale.
```{r part2-district, echo=TRUE, eval=TRUE, fig.width=8, fig.height=7}
# Counts per district per month (finer spatial scale) ---------------------
counts_dist_month <- df |>
  count(month, district_en, name = "Count")

# Would a finer (district) scale work? 
n_months_total <- n_distinct(df$month)                  # 18 possible months
dist_months <- counts_dist_month |>
  group_by(district_en) |>
  summarise(n_months = n_distinct(month), .groups = "drop")

cat("Districts in total:", nrow(dist_months), "\n")
cat("Districts with >12 months of reported cases:",
    sum(dist_months$n_months > 12), "\n")
cat("Provinces with >12 months (from above):", length(selected_provinces), "\n")

# Share of all reported cases falling in the qualifying districts vs provinces
ok_dist <- dist_months |>
  filter(n_months > 12) |>
  pull(district_en)
cat(sprintf("The %d qualifying districts hold %.1f%% of all reported cases\n",
            length(ok_dist), 100 * mean(df$district_en %in% ok_dist)))
cat(sprintf("The %d qualifying provinces hold %.1f%% of all reported cases\n",
            length(selected_provinces), 100 * mean(df$province_en %in% selected_provinces)))

# Months of reported cases per province and per district
prov_months <- counts_prov_month |>
  group_by(province_en) |>
  summarise(n_months = n_distinct(month), .groups = "drop")

months_bar <- function(d, unit_label) {
  ggplot(d, aes(n_months)) +
    geom_bar(fill = "steelblue") +
    geom_vline(xintercept = 12.5, linetype = "dashed", colour = "red") +
    scale_x_continuous(
      breaks = seq_len(n_months_total),
      limits = c(0.5, n_months_total + 0.5)
    ) +
    labs(y = paste0("number of ", unit_label)) +
    theme_bw()
}

p_prov_freq <- months_bar(prov_months, "provinces") +
  labs(x = NULL, title = "Provinces by number of months with reported cases")
p_dist_freq <- months_bar(dist_months, "districts") +
  labs(x = paste0("months with reported cases (out of ", n_months_total, ")"),
       title = "Districts by number of months with reported cases")

print(p_prov_freq / p_dist_freq)
```

**2.1 How many provinces are reported in the surveillance data, and how many are
kept after the "more than 12 months" filter? Why might we want to restrict to
provinces with a longer record before trying to relate cases to climate?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 2.1 Suggested answer: click to reveal (try it first)
There are many provinces in the raw data, but only those reported in more
than 12 months are kept (18 provinces kept out of 66), and of those only the ones whose names
match the boundary shapefile are carried forward. Restricting to provinces with a
longer record means each province has enough months to show a seasonal signal and to
standardise sensibly; provinces with only a handful of reported months cannot support
a climate–seasonality analysis.
:::

::: speaker-note
Core question for the 90-minute run; pause here. The other Part 2 questions are for self-study.
:::

**2.2 Why analyse at the province scale? What do we gain and lose by treating
the province as the spatial unit, and what would push you to a finer or coarser
scale?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 2.2 Suggested answer: click to reveal (try it first)
The spatial unit is a trade-off between signal and sparsity, and the
case counts sharpen it. Going finer would likely work, with 90.8% of the malaria
data still included. But visualising the 37 districts that clear the ">12
months" rule vs the 18 provinces is more difficult. It would also spread almost
the same case burden across roughly twice as many units (37 districts versus 18
provinces), so each unit's monthly counts are smaller and noisier, and 303 of
the 340 districts are too sparse to use at all. The province scale captures
nearly all cases (97.5%) in the fewest robust units, and so it is a pragmatic
choice. Going finer would suit a focused analysis of the high-burden districts;
going coarser (national) would give complete series but average over very
different climates and malaria regimes, likely washing out the signal.
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>

---

# **Part 3:** Environmental variables (reanalysis)

::: speaker-note
The concept to focus on here is the *spatial mismatch*: climate comes on a 
regular grid, cases come per province, and the two steps below bridge it. 

One detail that is not obvious from the code: temperature stays in Kelvin 
through standardisation because a constant offset (Kelvin vs Celsius) does not 
change a standardised (z-scored) series, so it only matters for the plots, where
we convert to °C.
:::

Malaria is reported per province, but ERA5 climate comes on a regular grid. To
link them, we summarise the grid to provinces in two steps: **strict cell-centre
assignment**, where each grid cell is assigned to the province whose polygon 
contains the cell's centre (a point-in-polygon test with `sf`, matching the
[`regionmask`](https://regionmask.readthedocs.io){target="_blank"} package), and 
then an **area-weighted average** of those cells, weighting each by cos of its 
latitude so that cells count in proportion to their area.

First load the shape files and check that the surveillance province names match 
the shapefile's ADM1_EN names. Keep only the selected provinces that are present 
in the boundaries. Report any that are dropped so name mismatches are visible.
```{r part3-shape, echo=TRUE, eval=TRUE}
library(sf)
library(terra)

# Province (ADM1) boundaries ----------------------------------------------
shape_file <- st_read(
  file.path(data_dir, "shape_files",
            "tha_admbnda_adm1_rtsd_20190221",
            "tha_admbnda_adm1_rtsd_20190221.shp"),
  quiet = TRUE
)
region_names <- shape_file$ADM1_EN

# Preview what we have loaded ---------------------------------------------
shape_file |>
  slice_head(n = 5) |>
  knitr::kable()      # the first 5 rows as a clean table in quarto

cat("Number of provinces in shapefile:", length(region_names), "\n")

missing <- setdiff(selected_provinces, region_names)
if (length(missing))
  cat("Provinces not matched in shapefile and dropped: ",
          paste(missing, collapse = ", "), "\n")

selected_provinces <- intersect(selected_provinces, region_names)
cat("Provinces carried forward (matched to boundaries):",
    length(selected_provinces), "\n")
```

Next, read the ERA5 monthly reanalysis: one netCDF file per month for 2 m
temperature (`2t`) and for total precipitation (`tp`), cropped to the Thailand 
region.

```{r part3-reanalysis, echo=TRUE, eval=TRUE}
# One netCDF file per month; sorting the file paths puts them in time order.
tas_files <- sort(list.files(
  file.path(data_dir, "reanalysis", "monthly-averaged", "2t"),
  pattern = "\\.nc$", recursive = TRUE, full.names = TRUE))
pr_files <- sort(list.files(
  file.path(data_dir, "reanalysis", "monthly-averaged", "tp"),
  pattern = "\\.nc$", recursive = TRUE, full.names = TRUE))

tas_r <- rast(tas_files)   # temperature, one layer per month (Kelvin)
pr_r  <- rast(pr_files)    # total precipitation, one layer per month (metres)

# Restrict to the Thailand region 
region_ext <- ext(97, 108, 5, 22)
tas_r <- crop(tas_r, region_ext)
pr_r  <- crop(pr_r,  region_ext)

# Rainfall time base. ERA5 tp and CMIP flux are per-day rates; multiplying by the days
# in a month reports monthly totals. This ONE constant is the whole monthly-vs-daily
# switch: set it to 1 for daily rates. It is reused by the CMIP conversion in Part 4.
DAYS_PER_MONTH <- 30
pr_r  <- pr_r * DAYS_PER_MONTH   # tp: mean daily rate (m/day) -> monthly total (m/month)

# Recover the month for each layer from the filename (YYYYMMDD-...)
month_from_file <- function(f)
  floor_date(as.Date(sub(".*_(\\d{8})-\\d{8}\\.nc$", "\\1", basename(f)),
                     format = "%Y%m%d"), "month")
clim_months <- month_from_file(tas_files)
cat("Reanalysis covers", length(clim_months), "months:",
    format(min(clim_months)), "to", format(max(clim_months)), "\n")
```

As a quick check that the data cover the right region, map the time-mean 
temperature and rainfall with the province boundaries drawn on top.

```{r part3-maps, echo=TRUE, eval=TRUE, fig.width=11, fig.height=5}
# Map of the time-mean temperature and rainfall over the region -----------
tas_mean_c <- mean(tas_r) - 273.15          # Kelvin -> Celsius for display
pr_mean    <- mean(pr_r)

map_df_t <- as.data.frame(tas_mean_c, xy = TRUE) |>
  rename(value = last_col())
map_df_r <- as.data.frame(pr_mean, xy = TRUE) |>
  rename(value = last_col())

m_t <- ggplot(map_df_t, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey30",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_steps2(low = "#2166ac", mid = "white", high = "#b2182b",
                    midpoint = 25, limits = c(20, 30), breaks = seq(20, 30, 1),
                    name = "°C", oob = scales::squish) +
  coord_sf(expand = FALSE) +
  labs(title = "Mean temperature", x = "lon", y = "lat") +
  theme_bw()

m_r <- ggplot(map_df_r, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey30",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_b(name = "m/month", option = "mako", n.breaks = 8) +
  coord_sf(expand = FALSE) +
  labs(title = "Mean rainfall", x = "lon", y = "lat") +
  theme_bw()

print(m_t + m_r)
```

Now assign each grid cell to a province and take the area-weighted provincial mean
for every month, giving one temperature and one rainfall series per province. Two
checks sit side by side below: the left map confirms each grid cell landed in the
right province, and the right map shows which provinces we actually carry into the
model (the subset chosen in Part 2, the rest greyed out).

```{r part3-extract, echo=TRUE, eval=TRUE, fig.width=11, fig.height=7}
# Assign each grid cell to the province whose polygon contains its *centre*.
grid_xy  <- terra::xyFromCell(tas_r, seq_len(terra::ncell(tas_r)))
grid_pts <- sf::st_as_sf(as.data.frame(grid_xy), coords = c("x", "y"),
                         crs = sf::st_crs(shape_file))
within   <- sf::st_within(grid_pts, shape_file)          # cell centre in polygon?
cell_region <- vapply(within,
                      function(h) if (length(h)) h[1] else NA_integer_,
                      integer(1))

# Check 1: colour each grid cell by the province it landed in.
assign_df <- tibble::tibble(
  x = grid_xy[, 1],
  y = grid_xy[, 2],
  province_id = cell_region
)
p_assign <- assign_df |>
  filter(!is.na(province_id)) |>
  ggplot(aes(x, y, fill = province_id)) +
  geom_raster() +
  geom_sf(data = shape_file, fill = NA, colour = "grey20",
          linewidth = 0.2, inherit.aes = FALSE) +
  scale_fill_viridis_c(guide = "none") +
  coord_sf(expand = FALSE) +
  labs(title = "Grid cells coloured by assigned province", x = "lon", y = "lat") +
  theme_bw()

# Check 2: which provinces are carried into the model (selected in Part 2)?
modelled_sf <- shape_file |>
  mutate(
    modelled = if_else(
      ADM1_EN %in% selected_provinces,
      "modelled",
      "not modelled"
    )
  )
p_modelled <- ggplot(modelled_sf) +
  geom_sf(aes(fill = modelled), colour = "grey40", linewidth = 0.2) +
  scale_fill_manual(values = c("modelled" = "#08519c", "not modelled" = "grey88"),
                    name = NULL) +
  coord_sf(expand = FALSE) +
  labs(title = "Provinces carried into the model", x = "lon", y = "lat") +
  theme_bw()

print(p_assign | p_modelled)   # side by side (patchwork)

# Provincial means, area-weighted by cos(latitude); see technical note.
# The assignment and weights are reused for the CMIP data in Part 4.
cell_w <- cos(grid_xy[, 2] * pi / 180)                    # per-cell area weight (cos latitude)
region_means <- function(r) {
  vals <- terra::values(r)                                # ncell x nlayer
  out  <- matrix(NA_real_, length(region_names), ncol(vals),
                 dimnames = list(region_names, NULL))
  for (k in seq_along(region_names)) {
    sel <- which(cell_region == k)
    if (!length(sel)) next
    w    <- cell_w[sel]
    vsel <- vals[sel, , drop = FALSE]
    # Weighted mean over the NON-NA cells: normalise by the weights actually used, so a
    # layer that is all NA gives NA (not 0). Matches xarray's weighted().mean(skipna);
    # it matters for CMIP, where some models are missing (NA) in the final months.
    out[k, ] <- colSums(vsel * w, na.rm = TRUE) / colSums((!is.na(vsel)) * w)
  }
  out
}

tas_prov <- region_means(tas_r)
pr_prov  <- region_means(pr_r)

# Tidy into long tables, keeping only the selected provinces
to_long <- function(mat, value_name) {
  m <- as.matrix(mat)
  rownames(m) <- region_names
  colnames(m) <- as.character(clim_months)
  as.data.frame(m) |>
    tibble::rownames_to_column("province_en") |>
    pivot_longer(-province_en, names_to = "month", values_to = value_name) |>
    mutate(month = as.Date(month))
}

clim_long <- to_long(tas_prov, "temp_k") |>
  left_join(to_long(pr_prov, "rain_m"), by = c("province_en", "month")) |>
  mutate(temp_c = temp_k - 273.15) |>
  filter(province_en %in% selected_provinces)

clim_long |>
  slice_head(n = 6) |>
  knitr::kable()
```

::: {.callout-note collapse="true"}
## Technical note: Why cos of latitude is the exact area weight

A grid cell's area is its north-south height times its east-west width. On a regular
grid the height is the same for every cell (a fixed step in latitude is a fixed
distance $R\,\Delta\theta$ along the surface), so all the variation is in the width.
The width shrinks towards the poles, because the meridians (lines of longitude)
converge: at latitude $\theta$ the circle of latitude has radius $R\cos\theta$, not
the full $R$.

```{=html}
<svg viewBox="0 0 420 300" width="100%" style="max-width:420px;height:auto;display:block;margin:0.6rem auto" xmlns="http://www.w3.org/2000/svg" role="img" aria-label="Cross-section of a globe: at latitude theta the circle of latitude has radius R times cosine theta, smaller than the equator's radius R.">
  <circle cx="200" cy="160" r="105" fill="#eaf2fb" stroke="#274b6d" stroke-width="1.5"/>
  <line x1="200" y1="47" x2="200" y2="273" stroke="#274b6d" stroke-width="1" stroke-dasharray="4 3"/>
  <text x="200" y="40" text-anchor="middle" font-size="11" fill="#274b6d">pole</text>
  <ellipse cx="200" cy="160" rx="105" ry="24" fill="none" stroke="#274b6d" stroke-width="1" stroke-dasharray="3 3"/>
  <text x="310" y="160" font-size="11" fill="#274b6d">equator (radius R)</text>
  <ellipse cx="200" cy="86" rx="74" ry="13" fill="none" stroke="#b2182b" stroke-width="1.7"/>
  <line x1="200" y1="160" x2="274" y2="86" stroke="#274b6d" stroke-width="1.6"/>
  <text x="248" y="120" font-size="13" fill="#274b6d" font-style="italic">R</text>
  <line x1="200" y1="86" x2="274" y2="86" stroke="#b2182b" stroke-width="2.4"/>
  <circle cx="274" cy="86" r="3.4" fill="#b2182b"/>
  <text x="282" y="84" font-size="12.5" fill="#b2182b" font-style="italic">P</text>
  <text x="237" y="79" font-size="12.5" fill="#b2182b" text-anchor="middle">R cos &#952;</text>
  <text x="118" y="89" font-size="11" fill="#b2182b" text-anchor="end">circle of latitude &#952;</text>
  <line x1="120" y1="86" x2="128" y2="86" stroke="#b2182b" stroke-width="1"/>
  <line x1="200" y1="160" x2="305" y2="160" stroke="#274b6d" stroke-width="1" stroke-dasharray="2 2"/>
  <text x="226" y="153" font-size="13" fill="#274b6d" font-style="italic">&#952;</text>
  <path d="M200,94 L208,94 L208,86" fill="none" stroke="#274b6d" stroke-width="1"/>
  <path d="M258,86 A16 16 0 0 0 262.6,97.3" fill="none" stroke="#274b6d" stroke-width="1"/>
  <text x="252" y="101" font-size="12" fill="#274b6d" font-style="italic">&#952;</text>
</svg>
```

Why $R\cos\theta$? In the cross-section above, the line from the centre out to $P$ is
a radius of the sphere, so it has length $R$, exactly like the equator's radius.
Because the equator and the $R\cos\theta$ segment are parallel, the angle at $P$
equals the latitude $\theta$ (right angle marked where $R\cos\theta$ meets the axis),
so in that right triangle $R\cos\theta$ is the side adjacent to $\theta$. A fixed
longitude span $\Delta\lambda$ therefore covers an east-west distance
$R\cos\theta\,\Delta\lambda$.

Now take a whole cell, bounded by latitudes $\theta_1$ and $\theta_2$ and a longitude
span $\Delta\lambda$:

```{=html}
<!-- Second diagram. To reproduce: run `python3 make_area_weight_diagram.py` (in this folder); it writes area_weight_diagram.svg, then paste that file's <svg>...</svg> in here. Do not hand-edit the path data. -->
<svg viewBox="0 0 430 320" width="100%" style="max-width:430px;height:auto;display:block;margin:0.6rem auto" xmlns="http://www.w3.org/2000/svg" role="img" aria-label="A globe with two meridians running pole to pole that bound a highlighted grid cell between latitudes theta-one and theta-two spanning delta-lambda in longitude.">
<circle cx="205" cy="158" r="105" fill="#eaf2fb" stroke="#274b6d" stroke-width="1.5"/>
<path d="M205.0,202.4 L209.6,202.3 L214.2,202.2 L218.7,202.0 L223.2,201.7 L227.7,201.3 L232.2,200.9 L236.6,200.3 L240.9,199.7 L245.2,199.0 L249.4,198.2 L253.5,197.4 L257.5,196.4 L261.4,195.4 L265.2,194.3 L268.9,193.2 L272.5,192.0 L275.9,190.7 L279.2,189.4 L282.4,188.0 L285.4,186.5 L288.3,185.0 L291.0,183.5 L293.6,181.8 L295.9,180.2 L298.1,178.5 L300.2,176.8 L302.0,175.0 L303.7,173.2 L305.1,171.3 L306.4,169.5 L307.5,167.6 L308.4,165.7 L309.1,163.8 L309.6,161.9 L309.9,159.9 L310.0,158.0" fill="none" stroke="#9bb0c4" stroke-width="1"/><path d="M309.9,156.1 L309.6,154.1 L309.1,152.2 L308.4,150.3 L307.5,148.4 L306.4,146.5 L305.1,144.7 L303.7,142.8 L302.0,141.0 L300.2,139.2 L298.1,137.5 L295.9,135.8 L293.6,134.2 L291.0,132.5 L288.3,131.0 L285.4,129.5 L282.4,128.0 L279.2,126.6 L275.9,125.3 L272.5,124.0 L268.9,122.8 L265.2,121.7 L261.4,120.6 L257.5,119.6 L253.5,118.6 L249.4,117.8 L245.2,117.0 L240.9,116.3 L236.6,115.7 L232.2,115.1 L227.7,114.7 L223.2,114.3 L218.7,114.0 L214.2,113.8 L209.6,113.7 L205.0,113.6 L200.4,113.7 L195.8,113.8 L191.3,114.0 L186.8,114.3 L182.3,114.7 L177.8,115.1 L173.4,115.7 L169.1,116.3 L164.8,117.0 L160.6,117.8 L156.5,118.6 L152.5,119.6 L148.6,120.6 L144.8,121.7 L141.1,122.8 L137.5,124.0 L134.1,125.3 L130.8,126.6 L127.6,128.0 L124.6,129.5 L121.7,131.0 L119.0,132.5 L116.4,134.2 L114.1,135.8 L111.9,137.5 L109.8,139.2 L108.0,141.0 L106.3,142.8 L104.9,144.7 L103.6,146.5 L102.5,148.4 L101.6,150.3 L100.9,152.2 L100.4,154.1 L100.1,156.1 L100.0,158.0" fill="none" stroke="#9bb0c4" stroke-width="1" stroke-dasharray="2 3" opacity="0.5"/><path d="M100.1,159.9 L100.4,161.9 L100.9,163.8 L101.6,165.7 L102.5,167.6 L103.6,169.5 L104.9,171.3 L106.3,173.2 L108.0,175.0 L109.8,176.8 L111.9,178.5 L114.1,180.2 L116.4,181.8 L119.0,183.5 L121.7,185.0 L124.6,186.5 L127.6,188.0 L130.8,189.4 L134.1,190.7 L137.5,192.0 L141.1,193.2 L144.8,194.3 L148.6,195.4 L152.5,196.4 L156.5,197.4 L160.6,198.2 L164.8,199.0 L169.1,199.7 L173.4,200.3 L177.8,200.9 L182.3,201.3 L186.8,201.7 L191.3,202.0 L195.8,202.2 L200.4,202.3 L205.0,202.4" fill="none" stroke="#9bb0c4" stroke-width="1"/>
<path d="M205.0,176.2 L209.4,176.2 L213.8,176.1 L218.2,175.9 L222.6,175.6 L227.0,175.2 L231.2,174.8 L235.5,174.2 L239.7,173.6 L243.8,173.0 L247.9,172.2 L251.8,171.4 L255.7,170.5 L259.5,169.5 L263.2,168.5 L266.7,167.4 L270.2,166.2 L273.5,165.0 L276.7,163.7 L279.8,162.3 L282.7,160.9 L285.5,159.5 L288.1,158.0 L290.5,156.4 L292.8,154.8 L295.0,153.2 L296.9,151.5 L298.7,149.8 L300.3,148.0 L301.7,146.3 L303.0,144.5 L304.0,142.6 L304.9,140.8 L305.6,139.0 L306.0,137.1 L306.3,135.2 L306.4,133.4 L306.3,131.5 L306.0,129.6" fill="none" stroke="#6b8cae" stroke-width="1.2"/><path d="M305.6,127.8 L304.9,125.9 L304.0,124.1 L303.0,122.3 L301.7,120.5 L300.3,118.7 L298.7,117.0 L296.9,115.3 L295.0,113.6 L292.8,111.9 L290.5,110.3 L288.1,108.8 L285.5,107.3 L282.7,105.8 L279.8,104.4 L276.7,103.1 L273.5,101.8 L270.2,100.5 L266.7,99.4 L263.2,98.3 L259.5,97.2 L255.7,96.2 L251.8,95.4 L247.9,94.5 L243.8,93.8 L239.7,93.1 L235.5,92.5 L231.3,92.0 L227.0,91.5 L222.6,91.2 L218.2,90.9 L213.8,90.7 L209.4,90.5 L205.0,90.5 L200.6,90.5 L196.2,90.7 L191.8,90.9 L187.4,91.2 L183.0,91.5 L178.8,92.0 L174.5,92.5 L170.3,93.1 L166.2,93.8 L162.1,94.5 L158.2,95.4 L154.3,96.2 L150.5,97.2 L146.8,98.3 L143.3,99.4 L139.8,100.5 L136.5,101.8 L133.3,103.1 L130.2,104.4 L127.3,105.8 L124.5,107.3 L121.9,108.8 L119.5,110.3 L117.2,111.9 L115.0,113.6 L113.1,115.3 L111.3,117.0 L109.7,118.7 L108.3,120.5 L107.0,122.3 L106.0,124.1 L105.1,125.9 L104.4,127.8" fill="none" stroke="#6b8cae" stroke-width="1.2" stroke-dasharray="2 3" opacity="0.5"/><path d="M104.0,129.6 L103.7,131.5 L103.6,133.4 L103.7,135.2 L104.0,137.1 L104.4,139.0 L105.1,140.8 L106.0,142.6 L107.0,144.5 L108.3,146.3 L109.7,148.0 L111.3,149.8 L113.1,151.5 L115.0,153.2 L117.2,154.8 L119.5,156.4 L121.9,158.0 L124.5,159.5 L127.3,160.9 L130.2,162.3 L133.3,163.7 L136.5,165.0 L139.8,166.2 L143.3,167.4 L146.8,168.5 L150.5,169.5 L154.3,170.5 L158.2,171.4 L162.1,172.2 L166.2,173.0 L170.3,173.6 L174.5,174.2 L178.8,174.8 L183.0,175.2 L187.4,175.6 L191.8,175.9 L196.2,176.1 L200.6,176.2 L205.0,176.2" fill="none" stroke="#6b8cae" stroke-width="1.2"/>
<path d="M205.0,117.0 L208.1,116.9 L211.1,116.9 L214.2,116.7 L217.2,116.5 L220.2,116.3 L223.2,116.0 L226.1,115.6 L229.0,115.2 L231.9,114.7 L234.7,114.2 L237.4,113.6 L240.1,113.0 L242.7,112.3 L245.3,111.6 L247.8,110.8 L250.2,110.0 L252.5,109.2 L254.7,108.3 L256.8,107.3 L258.8,106.4 L260.7,105.4 L262.6,104.3 L264.3,103.2 L265.8,102.1 L267.3,101.0 L268.7,99.8 L269.9,98.6 L271.0,97.4 L272.0,96.2 L272.9,95.0 L273.6,93.7 L274.2,92.4 L274.7,91.2 L275.0,89.9 L275.2,88.6 L275.3,87.3 L275.2,86.0 L275.0,84.7 L274.7,83.4 L274.2,82.1 L273.6,80.9 L272.9,79.6 L272.0,78.4 L271.0,77.1 L269.9,75.9 L268.7,74.7 L267.3,73.6 L265.8,72.4" fill="none" stroke="#6b8cae" stroke-width="1.2"/><path d="M264.3,71.3 L262.6,70.2 L260.7,69.2 L258.8,68.2 L256.8,67.2 L254.7,66.3 L252.5,65.4 L250.2,64.5 L247.8,63.7 L245.3,63.0 L242.7,62.2 L240.1,61.6 L237.4,60.9 L234.7,60.4 L231.9,59.8 L229.0,59.4 L226.1,59.0 L223.2,58.6 L220.2,58.3 L217.2,58.0 L214.2,57.8 L211.1,57.7 L208.1,57.6 L205.0,57.6 L201.9,57.6 L198.9,57.7 L195.8,57.8 L192.8,58.0 L189.8,58.3 L186.8,58.6 L183.9,59.0 L181.0,59.4 L178.1,59.8 L175.3,60.4 L172.6,60.9 L169.9,61.6 L167.3,62.2 L164.7,63.0 L162.2,63.7 L159.8,64.5 L157.5,65.4 L155.3,66.3 L153.2,67.2 L151.2,68.2 L149.3,69.2 L147.4,70.2 L145.7,71.3" fill="none" stroke="#6b8cae" stroke-width="1.2" stroke-dasharray="2 3" opacity="0.5"/><path d="M144.2,72.4 L142.7,73.6 L141.3,74.7 L140.1,75.9 L139.0,77.1 L138.0,78.4 L137.1,79.6 L136.4,80.9 L135.8,82.1 L135.3,83.4 L135.0,84.7 L134.8,86.0 L134.7,87.3 L134.8,88.6 L135.0,89.9 L135.3,91.2 L135.8,92.4 L136.4,93.7 L137.1,95.0 L138.0,96.2 L139.0,97.4 L140.1,98.6 L141.3,99.8 L142.7,101.0 L144.2,102.1 L145.7,103.2 L147.4,104.3 L149.3,105.4 L151.2,106.4 L153.2,107.3 L155.3,108.3 L157.5,109.2 L159.8,110.0 L162.2,110.8 L164.7,111.6 L167.3,112.3 L169.9,113.0 L172.6,113.6 L175.3,114.2 L178.1,114.7 L181.0,115.2 L183.9,115.6 L186.8,116.0 L189.8,116.3 L192.8,116.5 L195.8,116.7 L198.9,116.9 L201.9,116.9 L205.0,117.0" fill="none" stroke="#6b8cae" stroke-width="1.2"/>
<path d="M205.0,253.2 L203.5,254.1 L202.1,255.0 L200.6,255.8 L199.2,256.6 L197.7,257.3 L196.3,257.9 L194.9,258.4 L193.4,258.9 L192.0,259.3 L190.6,259.7 L189.2,259.9 L187.8,260.1 L186.4,260.3 L185.1,260.3 L183.7,260.3 L182.4,260.2 L181.0,260.1 L179.7,259.9 L178.5,259.6" fill="none" stroke="#274b6d" stroke-width="1.4" stroke-dasharray="2 3" opacity="0.5"/><path d="M177.2,259.2 L175.9,258.8 L174.7,258.3 L173.5,257.7 L172.3,257.1 L171.1,256.4 L170.0,255.6 L168.9,254.8 L167.8,253.9 L166.7,252.9 L165.7,251.9 L164.6,250.8 L163.7,249.6 L162.7,248.4 L161.8,247.1 L160.9,245.8 L160.0,244.4 L159.1,242.9 L158.3,241.4 L157.6,239.8 L156.8,238.2 L156.1,236.5 L155.4,234.7 L154.8,232.9 L154.2,231.1 L153.6,229.2 L153.1,227.2 L152.5,225.2 L152.1,223.2 L151.6,221.1 L151.3,219.0 L150.9,216.8 L150.6,214.6 L150.3,212.3 L150.0,210.1 L149.8,207.7 L149.7,205.4 L149.5,203.0 L149.4,200.6 L149.4,198.1 L149.4,195.6 L149.4,193.1 L149.4,190.6 L149.5,188.0 L149.7,185.5 L149.8,182.9 L150.0,180.3 L150.3,177.7 L150.6,175.0 L150.9,172.4 L151.3,169.7 L151.6,167.1 L152.1,164.4 L152.5,161.7 L153.1,159.0 L153.6,156.4 L154.2,153.7 L154.8,151.0 L155.4,148.3 L156.1,145.7 L156.8,143.0 L157.6,140.4 L158.3,137.7 L159.1,135.1 L160.0,132.5 L160.9,129.9 L161.8,127.4 L162.7,124.8 L163.7,122.3 L164.6,119.8 L165.7,117.3 L166.7,114.9 L167.8,112.5 L168.9,110.1 L170.0,107.7 L171.1,105.4 L172.3,103.1 L173.5,100.9 L174.7,98.7 L175.9,96.5 L177.2,94.4 L178.5,92.3 L179.7,90.3 L181.0,88.3 L182.4,86.4 L183.7,84.5 L185.1,82.6 L186.4,80.9 L187.8,79.1 L189.2,77.4 L190.6,75.8 L192.0,74.3 L193.4,72.7 L194.9,71.3 L196.3,69.9 L197.7,68.6 L199.2,67.3 L200.6,66.1 L202.1,64.9 L203.5,63.9 L205.0,62.8" fill="none" stroke="#274b6d" stroke-width="1.4"/>
<path d="M205.0,253.2 L206.5,254.1 L207.9,255.0 L209.4,255.8 L210.8,256.6 L212.3,257.3 L213.7,257.9 L215.1,258.4 L216.6,258.9 L218.0,259.3 L219.4,259.7 L220.8,259.9 L222.2,260.1 L223.6,260.3 L224.9,260.3 L226.3,260.3 L227.6,260.2 L229.0,260.1 L230.3,259.9 L231.5,259.6" fill="none" stroke="#274b6d" stroke-width="1.4" stroke-dasharray="2 3" opacity="0.5"/><path d="M232.8,259.2 L234.1,258.8 L235.3,258.3 L236.5,257.7 L237.7,257.1 L238.9,256.4 L240.0,255.6 L241.1,254.8 L242.2,253.9 L243.3,252.9 L244.3,251.9 L245.4,250.8 L246.3,249.6 L247.3,248.4 L248.2,247.1 L249.1,245.8 L250.0,244.4 L250.9,242.9 L251.7,241.4 L252.4,239.8 L253.2,238.2 L253.9,236.5 L254.6,234.7 L255.2,232.9 L255.8,231.1 L256.4,229.2 L256.9,227.2 L257.5,225.2 L257.9,223.2 L258.4,221.1 L258.7,219.0 L259.1,216.8 L259.4,214.6 L259.7,212.3 L260.0,210.1 L260.2,207.7 L260.3,205.4 L260.5,203.0 L260.6,200.6 L260.6,198.1 L260.6,195.6 L260.6,193.1 L260.6,190.6 L260.5,188.0 L260.3,185.5 L260.2,182.9 L260.0,180.3 L259.7,177.7 L259.4,175.0 L259.1,172.4 L258.7,169.7 L258.4,167.1 L257.9,164.4 L257.5,161.7 L256.9,159.0 L256.4,156.4 L255.8,153.7 L255.2,151.0 L254.6,148.3 L253.9,145.7 L253.2,143.0 L252.4,140.4 L251.7,137.7 L250.9,135.1 L250.0,132.5 L249.1,129.9 L248.2,127.4 L247.3,124.8 L246.3,122.3 L245.4,119.8 L244.3,117.3 L243.3,114.9 L242.2,112.5 L241.1,110.1 L240.0,107.7 L238.9,105.4 L237.7,103.1 L236.5,100.9 L235.3,98.7 L234.1,96.5 L232.8,94.4 L231.5,92.3 L230.3,90.3 L229.0,88.3 L227.6,86.4 L226.3,84.5 L224.9,82.6 L223.6,80.9 L222.2,79.1 L220.8,77.4 L219.4,75.8 L218.0,74.3 L216.6,72.7 L215.1,71.3 L213.7,69.9 L212.3,68.6 L210.8,67.3 L209.4,66.1 L207.9,64.9 L206.5,63.9 L205.0,62.8" fill="none" stroke="#274b6d" stroke-width="1.4"/>
<path d="M151.3,169.7 L154.6,170.6 L158.0,171.4 L161.5,172.1 L165.1,172.8 L168.7,173.4 L172.4,174.0 L176.1,174.5 L179.9,174.9 L183.7,175.3 L187.5,175.6 L191.4,175.8 L195.2,176.0 L199.1,176.2 L203.0,176.2 L207.0,176.2 L210.9,176.2 L214.8,176.0 L218.6,175.8 L222.5,175.6 L226.3,175.3 L230.1,174.9 L233.9,174.5 L237.6,174.0 L241.3,173.4 L244.9,172.8 L248.5,172.1 L252.0,171.4 L255.4,170.6 L258.7,169.7 L258.7,169.7 L258.3,166.6 L257.8,163.5 L257.2,160.4 L256.6,157.3 L256.0,154.2 L255.3,151.1 L254.5,148.0 L253.7,145.0 L252.9,141.9 L252.0,138.8 L251.1,135.8 L250.1,132.8 L249.1,129.8 L248.0,126.8 L247.0,123.9 L245.8,121.0 L244.7,118.1 L243.5,115.3 L242.2,112.5 L242.2,112.5 L239.9,113.0 L237.5,113.6 L235.1,114.1 L232.6,114.6 L230.1,115.0 L227.6,115.4 L225.0,115.7 L222.4,116.0 L219.8,116.3 L217.1,116.5 L214.4,116.7 L211.8,116.8 L209.1,116.9 L206.4,117.0 L203.6,117.0 L200.9,116.9 L198.2,116.8 L195.6,116.7 L192.9,116.5 L190.2,116.3 L187.6,116.0 L185.0,115.7 L182.4,115.4 L179.9,115.0 L177.4,114.6 L174.9,114.1 L172.5,113.6 L170.1,113.0 L167.8,112.5 L167.8,112.5 L166.5,115.3 L165.3,118.1 L164.2,121.0 L163.0,123.9 L162.0,126.8 L160.9,129.8 L159.9,132.8 L158.9,135.8 L158.0,138.8 L157.1,141.9 L156.3,145.0 L155.5,148.0 L154.7,151.1 L154.0,154.2 L153.4,157.3 L152.8,160.4 L152.2,163.5 L151.7,166.6 L151.3,169.7 Z" fill="#f6cccc" fill-opacity="0.8" stroke="#b2182b" stroke-width="1.8"/>
<circle cx="205.0" cy="62.8" r="2.4" fill="#274b6d"/>
<text x="205.0" y="52.8" text-anchor="middle" font-size="11" fill="#274b6d">pole</text>
<text x="97.6" y="137.4" text-anchor="end" font-size="12" fill="#274b6d" font-style="italic">&#952;&#8321;</text>
<text x="128.7" y="91.3" text-anchor="end" font-size="12" fill="#274b6d" font-style="italic">&#952;&#8322;</text>
<line x1="167.8" y1="104.5" x2="242.2" y2="104.5" stroke="#b2182b" stroke-width="1.2"/>
<text x="205.0" y="100.5" text-anchor="middle" font-size="12" fill="#b2182b">&#916;&#955;</text>
<text x="205.0" y="150.1" text-anchor="middle" font-size="12" fill="#b2182b">area A</text>
</svg>
```

The cell is not one rectangle: its width $R\cos\theta\,\Delta\lambda$ keeps shrinking
from $\theta_1$ up to $\theta_2$. So slice it into thin strips of latitude. A strip at
latitude $\theta$ has width $R\cos\theta\,\Delta\lambda$ and height $R\,d\theta$, so
area $R^2\,\Delta\lambda\,\cos\theta\,d\theta$. Adding the strips from $\theta_1$ to
$\theta_2$ (a sum that calculus writes as $\int\cos\theta\,d\theta$) gives the cell's
exact area

$$A = R^2\,\Delta\lambda\,\bigl(\sin\theta_2 - \sin\theta_1\bigr),$$

For a cell of height $\Delta\theta$ centred at $\theta$, the identity
$\sin\!\left(\theta+\tfrac{\Delta\theta}{2}\right) - \sin\!\left(\theta-\tfrac{\Delta\theta}{2}\right) = 2\cos\theta\,\sin\!\left(\tfrac{\Delta\theta}{2}\right)$
turns this into

$$A = R^2\,\Delta\lambda \,\cdot\, 2\sin\!\left(\tfrac{\Delta\theta}{2}\right) \,\cdot\, \cos\theta.$$

Every factor except $\cos\theta$ is identical for all cells on a regular grid, and
constant factors cancel in a normalised weighted mean. So weighting each cell by
$\cos$ of its centre latitude reproduces the true area weighting exactly: it is not a
small-angle approximation, it is exact for a sphere. (The Earth's slight oblateness
shifts cell areas by a fraction of a percent, far below anything that matters at this
provincial scale.)
:::

::: {.callout-note appearance="minimal"}
We weight each cell by its area (cos of latitude), the standard spatial mean. 
For climate and health work an even better choice is **population-density 
weighting**, which captures the climate where people actually live (noting this 
also requires data on the human population density). That matters most for
countries with very uneven populations (for example Australia, where most people 
are concentrated in a few coastal cities); at Thailand's provincial scale the 
difference is small, so we keep area-weighting here.
:::

::: speaker-note
Core question for the 90-minute run; pause here. The other Part 3 questions are for self-study.
:::

**3.1 We reduced each province's climate to a single monthly average. What might
that miss, and when might it matter?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 3.1 Suggested answer: click to reveal (try it first)
Averaging to a province discards all within-province spatial variation:
elevation gradients, forest edges, urban vs rural. For a large or topographically
varied province, a single average may not represent the conditions where
transmission actually happens (often specific ecological niches), so the link to
cases is weakened. It further assumes *average* temperature is what is important,
whereas minimums or maximums may drive the mosquito population changes.
:::

Plot the resulting monthly temperature and rainfall series for each selected province.

```{r part3-series, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
print(ggplot(clim_long, aes(month, temp_c)) +
  geom_line() + geom_point(size = 0.5) +
  facet_wrap(~ province_en, scales = "free_y") +
  labs(x = "date", y = "monthly mean temperature (°C)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))

print(ggplot(clim_long, aes(month, rain_m)) +
  geom_line() + geom_point(size = 0.5) +
  facet_wrap(~ province_en, scales = "free_y") +
  labs(x = "date", y = "monthly rainfall (m/month)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

**3.2 Look at the per-province temperature series. What kinds of variation do you
see, and which do you think will matter most for explaining malaria seasonality?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 3.2 Suggested answer: click to reveal (try it first)
Typically the within-year (seasonal) temperature swing within a province is
modest compared with, say, rainfall seasonality, and provinces differ from each
other too. Both the seasonal cycle and the between-province contrasts carry
information; the panel model in Part 5 uses both by standardising within province
and pooling across provinces. This is where checking maximum and minimum temperature
variations would be informative, and may help you decide which to carry forward.
:::

**3.3 The reanalysis only covers a couple of years here. What are the risks of
learning a climate–malaria relationship from such a short record?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 3.3 Suggested answer: click to reveal (try it first)
A short record risks fitting to year-specific anomalies (an unusual season,
a one-off outbreak) rather than a stable climate–malaria relationship, and gives
very little power to separate climate effects from other drivers. Projections built
on such a fit should be treated as illustrative.
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>

---

# **Part 4:** Future climate projections (CMIP6)

::: speaker-note
This part summarises the CMIP6 multi-model ensemble to the same provinces, under
two scenarios. The key ideas for participants: 

(1) CMIP6 is an *ensemble* of many climate models; we summarise across models 
with a mean and an interquartile band, rather than trusting any single model.

(2) the two scenarios (SSP1-2.6 low emissions, SSP5-8.5 high emissions) bracket 
a range of plausible futures. It is the heaviest step in  the document; on a 
slow machine give it a minute, and note that `freeze: auto` caches it afterwards.
:::

The learned relationship (Part 5) is only useful if we can apply it to the future.
CMIP6 provides projections of climatic variables from many climate models, under 
different emissions scenarios. 

We have curated the required data for you already, and trimmed the data to only 
over the region required for this tutorial. We once again focus on the same two 
we've now learned the relationship between malaria cases for: temperature 
(`tas`) and precipitation (`pr`). We summarise them to the same provinces, under 
two scenarios: **SSP1-2.6** (low emissions) and **SSP5-8.5** (high emissions).

The `read_cmip_var()` helper below is plumbing for reading netCDF files. You can treat
it as a black box: it says "give me this variable as an array, together with its
longitude, latitude, dates, and model list." You do not need to follow the netCDF
details to use what it returns.

```{r part4-read, echo=TRUE, eval=TRUE}
library(ncdf4)

# get common models across all scenarios and all variables
cmip_dir <- file.path(data_dir, "cmip6", "monthly-averaged")
cmip_files <- list(
  ssp126_tas = "cmip6_regrided_model_ensemble.Thai_region.ssp126.tas.nc",
  ssp585_tas = "cmip6_regrided_model_ensemble.Thai_region.ssp585.tas.nc",
  ssp126_pr  = "cmip6_regrided_model_ensemble.Thai_region.ssp126.pr.nc",
  ssp585_pr  = "cmip6_regrided_model_ensemble.Thai_region.ssp585.pr.nc"
)

# Read one CMIP variable; dimensions are read by name so storage order does not matter.
read_cmip_var <- function(path, var, from_date = NULL) {
  nc  <- nc_open(path)
  on.exit(nc_close(nc))
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  tim <- ncvar_get(nc, "time")
  models <- ncvar_get(nc, "model")
  # time units -> dates  S
  tunits <- ncatt_get(nc, "time", "units")$value
  origin <- as.Date(sub(".*since ", "", tunits))
  dates  <- floor_date(origin + tim, "month")
  arr <- ncvar_get(nc, var)                       # dims in file order
  dn <- vapply(nc$var[[var]]$dim, function(d) d$name, character(1))
  # reorder to lon, lat, time, model
  perm <- match(c("lon", "lat", "time", "model"), dn)
  arr <- aperm(arr, perm)
  # Trim to the reanalysis start so the fit and projection share a start month.
  if (!is.null(from_date)) {
    keep <- dates >= from_date
    arr  <- arr[, , keep, , drop = FALSE]
    dates <- dates[keep]
  }
  list(arr = arr, lon = lon, lat = lat, dates = dates, models = models)
}

cmip_start <- min(clim_months)   # reanalysis start month
tas126 <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp126_tas), "tas", cmip_start)
cat("CMIP6 grid:", length(tas126$lon), "lon x", length(tas126$lat), "lat;",
    length(tas126$models), "models;", length(tas126$dates), "months\n")
```

The next chunk defines the helpers that place each CMIP model on the reanalysis 
grid and average it to provinces, keeping every model (used later for the 
projection spread) alongside an ensemble summary.

```{r part4-summarise, echo=TRUE, eval=TRUE}
# Precipitation flux (kg m^-2 s^-1) -> monthly-total depth (m): seconds per day (86400)
# x days per month (DAYS_PER_MONTH, defined in Part 3) / water density (1000 kg m^-3).
# The monthly-vs-daily switch lives entirely in DAYS_PER_MONTH; this line just uses it.
PR_FLUX_TO_M <- 86400 * DAYS_PER_MONTH / 1000

ref_grid <- tas_r[[1]]        # reanalysis grid (already cropped to region)

# Build a north-up, multi-layer SpatRaster from one model's [lon, lat, time] slice
# (lon/lat treated as cell centres, so the extent is padded half a cell each side).
slice_to_rast <- function(slc, lon, lat, ref) {
  lon_o <- order(lon)
  lat_o <- order(lat)                               # index ascending
  slc <- slc[lon_o, lat_o, , drop = FALSE]          # lon asc, lat asc
  a <- aperm(slc, c(2, 1, 3))                        # -> lat x lon x time
  a <- a[dim(a)[1]:1, , , drop = FALSE]              # flip: north (max lat) = row 1
  r <- rast(a)                                       # 3rd dim -> layers
  lo <- sort(lon)
  la <- sort(lat)
  dx <- mean(diff(lo))
  dy <- mean(diff(la))
  ext(r) <- c(min(lo) - dx/2, max(lo) + dx/2,
              min(la) - dy/2, max(la) + dy/2)
  crs(r) <- crs(ref)                                 # match the reanalysis grid
  r
}

# Every model -> provincial means, keeping the per-model detail (province x month x
# model) that the Part 6 projection needs for the spread.
cmip_provincial_bymodel <- function(cv, var, ref, provinces) {
  nm <- length(cv$models)
  # CMIP_MODEL_CAP env var caps the number of models (a faster/debug run); 0 = all.
  cap <- suppressWarnings(as.integer(Sys.getenv("CMIP_MODEL_CAP", "")))
  if (!is.na(cap) && cap > 0) nm <- min(nm, cap)
  out <- vector("list", nm)
  for (m in seq_len(nm)) {
    slc <- cv$arr[, , , m]                     # lon x lat x time
    r <- slice_to_rast(slc, cv$lon, cv$lat, ref)
    names(r) <- as.character(cv$dates)
    if (var == "pr") r <- r * PR_FLUX_TO_M     # flux -> depth (m); see PR_FLUX_TO_M above
    else             r <- r - 273.15           # Kelvin -> Celsius
    r <- resample(r, ref, method = "bilinear") # interpolate onto reanalysis grid
    pm <- region_means(r)                       # strict cell-centre assignment (as Part 3)
    out[[m]] <- pm[provinces, , drop = FALSE]
  }
  arr3 <- simplify2array(out)                  # province x month x model
  list(arr = arr3, dates = cv$dates, provinces = provinces)
}

# Ensemble summary (mean + interquartile band) from a per-model array.
summarise_cmip <- function(bm, scenario, var) {
  arr3 <- bm$arr
  nt <- length(bm$dates)
  qf <- function(p) as.vector(apply(arr3, c(1, 2), quantile, probs = p, na.rm = TRUE))
  tibble::tibble(
    province_en = rep(bm$provinces, times = nt),
    month       = rep(bm$dates, each = length(bm$provinces)),
    scenario = scenario, variable = var,
    mean = as.vector(apply(arr3, c(1, 2), mean, na.rm = TRUE)),
    q05 = qf(0.05), q25 = qf(0.25), q75 = qf(0.75), q95 = qf(0.95)
  )
}
```

Now apply those helpers to all four files (two variables and two scenarios), restrict
to the models the files share, and build the ensemble summary for plotting.

```{r part4-run, echo=TRUE, eval=TRUE}
tas585 <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp585_tas), "tas", cmip_start)
pr126  <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp126_pr),  "pr",  cmip_start)
pr585  <- read_cmip_var(file.path(cmip_dir, cmip_files$ssp585_pr),  "pr",  cmip_start)

# Restrict all four files to their common models, in the same order (the projection
# pairs temperature and rainfall model-by-model).
common_models <- sort(Reduce(intersect,
  list(tas126$models, tas585$models, pr126$models, pr585$models)))
align_models <- function(cv) {
  idx <- match(common_models, cv$models)
  cv$arr    <- cv$arr[, , , idx, drop = FALSE]
  cv$models <- common_models
  cv
}
tas126 <- align_models(tas126)
tas585 <- align_models(tas585)
pr126 <- align_models(pr126)
pr585 <- align_models(pr585)
cat("CMIP models in common across all four files:", length(common_models), "\n")

# Per-model provincial climate (kept for the Part 6 projection)
cmip_tas <- list(
  "SSP1-2.6" = cmip_provincial_bymodel(tas126, "tas", ref_grid, selected_provinces),
  "SSP5-8.5" = cmip_provincial_bymodel(tas585, "tas", ref_grid, selected_provinces))
cmip_pr <- list(
  "SSP1-2.6" = cmip_provincial_bymodel(pr126, "pr", ref_grid, selected_provinces),
  "SSP5-8.5" = cmip_provincial_bymodel(pr585, "pr", ref_grid, selected_provinces))

# Ensemble summary for plotting
cmip_summary <- bind_rows(
  summarise_cmip(cmip_tas[["SSP1-2.6"]], "SSP1-2.6", "tas"),
  summarise_cmip(cmip_tas[["SSP5-8.5"]], "SSP5-8.5", "tas"),
  summarise_cmip(cmip_pr[["SSP1-2.6"]],  "SSP1-2.6", "pr"),
  summarise_cmip(cmip_pr[["SSP5-8.5"]],  "SSP5-8.5", "pr")
)
```

As a spatial check, map the model-and-time-mean temperature and rainfall over 
the region for each scenario, for 2025-2100. 

```{r part4-maps, echo=TRUE, eval=TRUE, fig.width=10, fig.height=11}
# Model-and-time mean CMIP field on the reanalysis grid, for one scenario/variable.
cmip_mean_grid <- function(cv, var) {
  fld <- apply(cv$arr, c(1, 2), mean, na.rm = TRUE)   # lon x lat: mean over time and model
  r <- slice_to_rast(array(fld, c(dim(fld), 1)), cv$lon, cv$lat, ref_grid)
  if (var == "pr") r <- r * PR_FLUX_TO_M else r <- r - 273.15
  resample(r, ref_grid, method = "bilinear")
}

cmip_map <- function(r, title, fill_scale) {
  d <- as.data.frame(r, xy = TRUE) |>
    rename(value = last_col())
  ggplot(d, aes(x, y, fill = value)) +
    geom_raster() +
    geom_sf(data = shape_file, fill = NA, colour = "grey30",
            linewidth = 0.2, inherit.aes = FALSE) +
    fill_scale +
    coord_sf(expand = FALSE) +
    labs(title = title, x = "lon", y = "lat") +
    theme_bw()
}

# Banded scales: blue-white-red for temperature,mako for rainfall. Rain limits 
# are the Part 4 daily values x DAYS_PER_MONTH (monthly totals).
temp_scale <- function() scale_fill_steps2(low = "#2166ac", mid = "white",
  high = "#b2182b", midpoint = 25, limits = c(20, 30), breaks = seq(20, 30, 1),
  name = "°C", oob = scales::squish)
rain_scale <- function() scale_fill_viridis_b(name = "m/month", option = "mako",
  limits = c(0.0036, 0.0084) * DAYS_PER_MONTH,
  breaks = seq(0.0036, 0.0084, 0.0006) * DAYS_PER_MONTH, oob = scales::squish)

t126 <- cmip_map(cmip_mean_grid(tas126, "tas"), "Mean temperature (SSP1-2.6)", temp_scale())
t585 <- cmip_map(cmip_mean_grid(tas585, "tas"), "Mean temperature (SSP5-8.5)", temp_scale())
r126 <- cmip_map(cmip_mean_grid(pr126, "pr"),  "Mean rainfall (SSP1-2.6)",   rain_scale())
r585 <- cmip_map(cmip_mean_grid(pr585, "pr"),  "Mean rainfall (SSP5-8.5)",   rain_scale())

print((t126 + t585) / (r126 + r585))
```

For one example province, plot the CMIP6 climate two ways: the monthly model-mean
series, and the annual series showing the ensemble mean, a single model, and the
interquartile range across models.

```{r part4-plot, echo=TRUE, eval=TRUE, fig.width=9, fig.height=8}
example_province <- selected_provinces[1]
prov_idx  <- match(example_province, selected_provinces)
one_model <- min(21L, dim(cmip_tas[["SSP1-2.6"]]$arr)[3])   # one member, to contrast with the ensemble

lab  <- c(tas = "temperature (°C)", pr = "total precipitation (m/month)")
cols <- c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")

# Monthly model-mean, from the ensemble summary
monthly <- cmip_summary |>
  filter(province_en == example_province)
print(ggplot(monthly, aes(month, mean, colour = scenario)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ variable, ncol = 1, scales = "free_y", labeller = labeller(variable = lab)) +
  scale_colour_manual(values = cols) +
  labs(x = "date", y = NULL,
       title = paste0("CMIP6 monthly model-mean for ", example_province)) +
  theme_bw())

# Annual: ensemble mean, one member, and interquartile band across models
annual <- monthly |>
  mutate(year = year(month)) |>
  group_by(scenario, variable, year) |>
  summarise(mean = mean(mean), q25 = mean(q25), q75 = mean(q75), .groups = "drop")

# one member's annual series for this province (dips into the per-model arrays)
one_annual <- bind_rows(
  lapply(names(cmip_tas), function(sc) tibble::tibble(
    year = year(cmip_tas[[sc]]$dates), scenario = sc, variable = "tas",
    value = cmip_tas[[sc]]$arr[prov_idx, , one_model])),
  lapply(names(cmip_pr), function(sc) tibble::tibble(
    year = year(cmip_pr[[sc]]$dates), scenario = sc, variable = "pr",
    value = cmip_pr[[sc]]$arr[prov_idx, , one_model]))
) |>
  group_by(scenario, variable, year) |>
  summarise(value = mean(value), .groups = "drop")

annual_lines <- bind_rows(
  transmute(annual, year, scenario, variable, value = mean, series = "model mean"),
  transmute(one_annual, year, scenario, variable, value, series = "one model")
)

print(ggplot() +
  geom_ribbon(data = annual, aes(year, ymin = q25, ymax = q75, fill = scenario), alpha = 0.2) +
  geom_line(data = annual_lines,
            aes(year, value, colour = scenario, linetype = series), linewidth = 0.7) +
  facet_wrap(~ variable, ncol = 1, scales = "free_y", labeller = labeller(variable = lab)) +
  scale_colour_manual(values = cols) +
  scale_fill_manual(values = cols, guide = "none") +
  scale_linetype_manual(name = "series", values = c("model mean" = "solid", "one model" = "dashed")) +
  labs(x = "year", y = NULL,
       title = paste0("CMIP6 annual projections for ", example_province),
       subtitle = "Solid = model mean, dashed = one model, band = interquartile range (only the middle 50% of models)") +
  theme_bw())
```

::: speaker-note
Core question for the 90-minute run; pause here. The other Part 4 questions are for self-study.
:::

**4.1 The projections are shown as a model mean with an interquartile band. Why do
we show a *range* across models rather than picking the single "best" model?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 4.1 Suggested answer: click to reveal (try it first)
Individual climate models disagree, especially on rainfall: note how the
randomly chosen single model often exceeds the interquartile range. The ensemble
mean is more reliable than any single model, and the spread is itself the honest
expression of projection *uncertainty*. Picking one "best" model hides that
uncertainty. Note also that the band in the plot above is *only* the
interquartile range, the middle 50% of models; the full disagreement is wider.
The plot below redraws the example province's temperature with the 5-95% range
as well.

```{r part4-sol-spread, echo=TRUE, eval=TRUE, fig.width=8, fig.height=4}
ci90 <- cmip_summary |>
  filter(province_en == example_province, variable == "tas") |>
  mutate(year = year(month)) |>
  group_by(scenario, year) |>
  summarise(mean = mean(mean), q05 = mean(q05), q25 = mean(q25),
            q75 = mean(q75), q95 = mean(q95), .groups = "drop")

print(ggplot(ci90, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.12, colour = NA) +   # 5-95% range
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.28, colour = NA) +   # interquartile range
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = "temperature (°C)",
       title = paste0("Model spread in projected temperature, ", example_province),
       subtitle = "Line = mean, inner band = interquartile range, outer band = 5-95% of models") +
  theme_bw())
```
:::

**4.2 Compare the two scenarios (SSP1-2.6 and SSP5-8.5). How far apart are the
temperature projections by the end of the century, and how does the model agreement
on rainfall compare?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 4.2 Suggested answer: click to reveal (try it first)
Temperature projections usually diverge clearly between the two scenarios
by late century, with SSP5-8.5 substantially warmer, and models tend to agree on
the direction of warming. Rainfall is far less certain: the inter-model band is
usually wide and the scenarios may overlap, reflecting genuine disagreement about
how precipitation will change.
:::

**4.3 We are about to feed these projections into a malaria model that was fitted
on only a couple of years of reanalysis. What assumption are we making about the
climate–malaria relationship when we extrapolate it decades into the future?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 4.3 Suggested answer: click to reveal (try it first)
We are assuming the climate–malaria relationship learned over a couple of
recent years holds, unchanged, decades into the future, i.e. that no adaptation,
intervention, land-use change, drug resistance, or shift in the mosquito/parasite
system alters it. That is a strong assumption and the main caveat on any such
projection.
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>

---

# **Part 5:** Fitting climate to incidence

::: speaker-note
This is the core of the session: a statistical model relating climate to malaria
cases across all provinces at once (a "panel" model). 

The design choices to draw out: 

(1) we model the *first difference* (month-to-month change) in cases rather
than the level, and we *standardise* per province, so that provinces of very
different sizes contribute comparably.

(2) climate enters as temperature, rainfall, and their squares, allowing a 
hump-shaped (optimum) response. 

(3) the fitted coefficients are *shared* across provinces, one climate–malaria 
relationship applied everywhere. 

Frame it explicitly as the *simplest* model, one set of coefficients for the 
whole country. 
:::

We now link climate to cases directly with a simple statistical model that pools all
provinces: a **panel method**, the same approach used to learn how temperature and
rainfall affect the gross domestic product of the world's economies
([Burke, Hsiang & Miguel, *Nature*, 2015](https://www.nature.com/articles/nature15725){target="_blank"}),
here applied to the month-to-month *change* in infection count. It is deliberately the
simplest model that works, one shared set of coefficients for the whole country, built
up in a few steps below.

First, line up the malaria counts and the climate on the same province-by-month grid.

```{r part5-matrices, echo=TRUE, eval=TRUE}
provinces <- selected_provinces
months <- clim_long |>
  distinct(month) |>
  arrange(month) |>
  pull(month)

# Case matrix: provinces x months, zero-filled where no cases were reported
cases_mat <- counts_prov_month |>
  filter(province_en %in% provinces) |>
  complete(province_en = provinces, month = months, fill = list(Count = 0)) |>
  filter(month %in% months) |>
  arrange(province_en, month) |>
  pivot_wider(names_from = month, values_from = Count) |>
  tibble::column_to_rownames("province_en") |>
  as.matrix()
cases_mat <- cases_mat[provinces, , drop = FALSE]

# Climate matrices on the same grid
mat_from_long <- function(var) {
  mat <- clim_long |>
    select(province_en, month, all_of(var)) |>
    pivot_wider(names_from = month, values_from = all_of(var)) |>
    tibble::column_to_rownames("province_en") |>
    as.matrix()
  mat[provinces, , drop = FALSE]
}
clim_temp <- mat_from_long("temp_k")   # Kelvin (offset irrelevant once standardised)
clim_rain <- mat_from_long("rain_m")

cat("cases:", nrow(cases_mat), "provinces x", ncol(cases_mat), "months;",
    "total cases =", sum(cases_mat), "\n")
```

Panel methods cannot account for trends, , as such we must first remove the 
trend by working with the first *difference*, and then *standardise* by the mean 
and standard deviation of the data in time per province.

```{r part5-standardise, echo=TRUE, eval=TRUE}
# SD (divide by n, like NumPy's np.std), not R's sd() (n-1)
popsd <- function(x) sqrt(mean((x - mean(x))^2))

# Response: standardised first difference of cases, per province
cases_fd <- cases_mat[, -1] - cases_mat[, -ncol(cases_mat)]
fd_mean  <- rowMeans(cases_fd)
fd_sd    <- apply(cases_fd, 1, popsd)
z_cases_fd <- (cases_fd - fd_mean) / fd_sd

# Predictors: standardised temperature and rainfall, per province
z_temp <- (clim_temp - rowMeans(clim_temp)) / apply(clim_temp, 1, popsd)
z_rain <- (clim_rain - rowMeans(clim_rain)) / apply(clim_rain, 1, popsd)
```

Vector `Y` is the standardised first difference of the counts, flattened across 
all provinces and months into one long vector. 

Matrix `X` has one row per province-month, and columns for standardised 
temperature, temperature squared, rainfall, and rainfall squared. The squared 
terms are included to allow for potential nonlinear dependencies of the climate 
upon the infection rates.

```{r part5-build, echo=TRUE, eval=TRUE}
# Flatten province x month into one vector; X columns: temp, temp^2, rain, rain^2.
Y  <- as.vector(z_cases_fd)
zt <- z_temp[, -1]
zr <- z_rain[, -1]
X  <- cbind(as.vector(zt), as.vector(zt)^2, as.vector(zr), as.vector(zr)^2)
colnames(X) <- c("temp", "temp^2", "rain", "rain^2")
```

We then solve `Y =  X B` for the coefficients `B` using standard ordinary least 
squares on the pooled data. Although one could in principle use more 
sophisticated elaborate methods (e.g. ridge or lasso regression).

```{r part5-fit, echo=TRUE, eval=TRUE}
# Centre; subtract away means of matrices
Y_mean <- mean(Y)
Y_prime <- Y - Y_mean
X_mean <- colMeans(X)
X_prime <- sweep(X, 2, X_mean)

# solve ordinary least squares problem
XtX = t(X_prime) %*% X_prime  #  XᵀX
XtY = t(X_prime) %*% Y_prime  # XᵀY
B   <- solve(XtX, XtY) # solves the linear system A B = C for B, with A = XᵀX and C = XᵀY. So it returns B = (XᵀX)⁻¹ XᵀY
# which is the closed-form OLS estimate, the coefficients that minimise the squared error ‖X B − Y‖²
print(round(setNames(as.vector(B), colnames(X)), 3))
```

We then evaluate predictions of the *change* in infection cases using this 
regression model (the fitted `B`), reshape the data to be a matrix per 
province and per month, and then finally apply a cummulative sum to return 
infection counts (as opposed to changes in infection counts).

Because we model the first difference and rebuild the level by that cumulative 
sum, this is an **ARIMA-style** model, an autoregressive model with exogenous 
inputs (ARX), temperature and rainfall being the exogenous drivers (hence 
`arx_fit`).

```{r part5-reconstruct, echo=TRUE, eval=TRUE}
# Predict the change 
Y_hat <- as.vector(X_prime %*% B) + Y_mean

# reshape to province x month
z_hat <- matrix(Y_hat, nrow = nrow(z_cases_fd))

#  Rescale first differences and apply cummulative sum to get back to counts
fitted_cases <- cases_mat * 0
for (i in seq_len(nrow(cases_mat))) {
  fitted_cases[i, 1]  <- cases_mat[i, 1]
  fitted_cases[i, -1] <- fitted_cases[i, 1] +
    cumsum(z_hat[i, ] * fd_sd[i] + fd_mean[i])
}
```

```{r verify-export, echo=FALSE, message=FALSE}
# Save this fit's actual outputs to the shared outputs/ folder, so they can be
# compared directly against the Python notebook's saved outputs (see
# python_workflow/compare.R). Instrumentation only: it exports the arrays already
# computed above, it does not recompute anything.
.odir <- file.path("..", "outputs")
dir.create(.odir, showWarnings = FALSE, recursive = TRUE)
jsonlite::write_json(list(
  selected   = provinces,
  months     = as.character(months),
  cases_mat  = cases_mat,
  clim_temp  = clim_temp,
  clim_rain  = clim_rain,
  panel_beta = as.vector(B),
  arx_fit    = fitted_cases
), file.path(.odir, "r_panel.json"), matrix = "rowmajor", digits = NA)
```

Finally, we visualise the model estimates (red line) as compared to the actual 
data (black line).

This is perhaps the most basic model one can develop, and is design to 
illustrate the logic and integration of the climate data. It uses one set of 
coefficients to represent the behaviour across all provinces.

```{r part5-diag, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
matrix_to_long <- function(x, series_name) {
  dimnames(x) <- dimnames(cases_mat)
  as.data.frame(x) |>
    tibble::rownames_to_column("province_en") |>
    pivot_longer(-province_en, names_to = "month", values_to = "cases") |>
    mutate(month = as.Date(month), series = series_name)
}

fit_long <- bind_rows(
  matrix_to_long(cases_mat, "observed"),
  matrix_to_long(fitted_cases, "fitted")
)

print(ggplot(fit_long, aes(month, cases, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c(observed = "black", fitted = "red")) +
  labs(x = "date", y = "monthly cases",
       title = "Observed vs panel-fitted monthly cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

The fitted `B` coefficients define the climate response. In **Part 6** we
apply them to the future CMIP6 climate to project malaria infection cases
forward. The optional SEIRS section at the end of this session fits a
transmission model to the same climate parameters, an illustration of the
mechanistic approach.

::: {.callout-note}
## Principle: collinear predictors are candidate associations, not causes
Temperature, temperature², rainfall and rainfall² are strongly correlated with one
another, so the individual `B` coefficients cannot be cleanly attributed to a single
driver: a large coefficient on one term can be offset by another. Treat them as a
combined climate response that fits the data, not as separate causal effects. In our
own climate-risk work the same collinearity (high variance-inflation factors) is why
predictors are reported as candidate drivers rather than confirmed causes
([Sexton et al. 2026](https://doi.org/10.1016/j.ecoinf.2026.103899){target="_blank"}).
:::

**5.1 What biologically reasonable relationships between temperature and
rainfall and malaria might you expect? Looking back at how the climate variables
enter the model, how were those relationships included here, and what might you
expand upon to capture them more fully?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 5.1 Suggested answer: click to reveal (try it first)
Biologically, we would expect the relationships to be *non-monotonic*
rather than straight lines: mosquito survival, biting, and parasite development
each have a thermal optimum, so transmission rises with temperature to a point and
then falls; and both drought and flushing rains can reduce breeding, so rainfall
also has an intermediate optimum. There are usually *lags* too, since it takes time
for climate to act through the mosquito population and the parasite's incubation.
In this model those relationships were included by adding *squared* terms
(temperature² and rainfall²) alongside the linear ones, which lets the fitted
response be hump-shaped (unimodal) instead of a straight line. To capture the
biology more fully you might add lagged climate terms, an interaction between
temperature and rainfall, higher-order or spline terms for a more flexible shape,
or move to a mechanistic model that represents the vector dynamics directly (as in
the optional SEIRS section).
:::

::: speaker-note
Core question for the 90-minute run; pause here. The other Part 5 questions are for self-study.
:::

**5.2 We standardised cases and climate *per province* before pooling. What would
go wrong if we pooled the raw counts across provinces without standardising?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 5.2 Suggested answer: click to reveal (try it first)
Provinces differ enormously in population and case numbers. Pooling raw
counts would let a few large-caseload provinces dominate the fit, and would confound
"more cases because bigger" with "more cases because of climate". Standardising per
province puts every province on a comparable footing so the shared coefficients
reflect the climate relationship, not province size.

It is worth separating two things that our per-province standardisation bundles
together. Dividing by population *first* (noting that this would require yet
another data source though!), to get a per-capita incidence, is a good
idea in its own right: it removes the "more cases because bigger" effect while
keeping each province's real *level* of incidence, which is a meaningful quantity to
compare across provinces. Our z-scoring goes further and also removes each province's
mean and scale, so the fit uses only the within-province variation over time. That is
what makes the coefficients comparable, but it also discards the between-province
signal, so it can obscure relationships that live in the differences *between*
provinces, for example a sparsely populated but high-incidence province whose burden
is driven by climate. A reasonable alternative is to do both: convert to incidence
first, then standardise, so the level information is at least available to inspect
before it is removed.
:::

**5.3 Look at the observed-vs-fitted panels. Where does the model do well, and where
does it do poorly? What might it be missing?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 5.3 Suggested answer: click to reveal (try it first)
The model tends to capture the broad seasonal rise and fall but misses
sharp peaks, sudden outbreaks, and province-specific idiosyncrasies. Malaria is
also driven by non-climate factors: interventions (bed nets, spraying), human
movement and importation, drug resistance, health-system reporting changes, none
of which are in a climate-only model. By construction it also uses only
*within-province* variation over time: standardising each province to its own mean
and SD removes each province's baseline, so the model learns how a province's
cases move with its *own* climate, not why one province carries a higher caseload
than another, which may itself be climate-driven. That between-province spatial signal
is deliberately traded away to avoid confounding climate with everything else that
differs between provinces.
:::

**5.4 To project future cases, we would apply these fitted coefficients to the
CMIP6 climate from Part 4. List one thing you would want to check about the future
climate values before trusting such a projection.**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 5.4 Suggested answer: click to reveal (try it first)
Sensible checks include: are the future climate values within the range the
model was trained on (extrapolation beyond the training range is dangerous)? Are the
units and standardisation applied identically to the projections? Is the seasonal
structure of the projections realistic? Does the province averaging match?
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>

---

# **Part 6:** Projecting future cases (statistical model)

::: speaker-note
Here we take the model fitted on the historical period (Part 5) and
push it forward under the CMIP6 climate (Part 4) to project future cases. 

Be sure to make the key caveat explicit (see the warning box). Treat the 
projection as illustrative, not as a forecast.
:::

We now project the fitted model forward. For each climate model and scenario, we
standardise the future CMIP6 temperature and rainfall using the **historical**
per-province mean and standard deviation (the same scaling the model was fitted
with), apply the fitted coefficients to get the predicted month-to-month change 
in cases, and rebuild counts by cumulative sum, seeding from the last observed 
month.

::: {.callout-warning}
## Limitations to keep in mind

**An illustrative case study, not a definitive projection**\
Rebuilding case counts by cumulatively summing predicted monthly *changes* over
decades behaves like a random walk and will drift, so the absolute numbers far 
into the future should not be read literally. This is an inherent limitation of 
pushing a first-difference (ARX) model a long way forward, and is one reason a 
bounded mechanistic model (the optional SEIRS section at the end of this 
session) may be preferred for long-range projection. Use this to see *how*
climate projections feed a fitted model, not as a prediction of future case
counts.

**Extrapolation beyond the training range**\
The model learned its climate response over the historical period, but future 
CMIP6 temperatures will exceed anything it was fitted on, so the quadratic 
response is being *extrapolated* into conditions never observed. That is a 
distinct source of uncertainty on top of the drift above, and one that is easy 
to miss because the code runs happily regardless. Formal diagnostics (for 
example mobility-oriented parity, or simply flagging where inputs leave the 
calibration range) can show which regions are being extrapolated into and 
treated with more caution. See, for example,
([Sexton et al. 2026](https://doi.org/10.1016/j.ecoinf.2026.103899){target="_blank"}).

**Climate is only one driver**\
Even a perfect climate-to-cases relationship captures just one component of 
malaria risk. Vector control, land use, human movement, immunity and 
health-system changes all matter and are held fixed here, so a projection that 
varies only the climate answers "how does the climate signal alone propagate",
not "how many cases will there be".
:::

```{r part6-project, echo=TRUE, eval=TRUE}
# Historical per-province scaling (in the CMIP arrays' units). Standardising is
# offset-invariant, so this matches the scaling the model was fitted with in Part 5.
temp_mean_p <- rowMeans(clim_temp) - 273.15   # clim_temp is Kelvin; CMIP tas is °C
temp_sd_p   <- apply(clim_temp, 1, popsd)
rain_mean_p <- rowMeans(clim_rain)
rain_sd_p   <- apply(clim_rain, 1, popsd)

last_obs <- cases_mat[, ncol(cases_mat)]      # seed the projection at the last observed month
hist_end <- max(months)

# Apply the fitted ARX model to one climate model's future climate, rebuild counts.
project_one <- function(tas_pm, pr_pm, dates) {
  zt <- (tas_pm - temp_mean_p) / temp_sd_p    # province x month (vectors recycle per row)
  zr <- (pr_pm  - rain_mean_p) / rain_sd_p
  yhat <- (zt   - X_mean[1]) * B[1] +      # predicted standardised first-difference
          (zt^2 - X_mean[2]) * B[2] +
          (zr   - X_mean[3]) * B[3] +
          (zr^2 - X_mean[4]) * B[4] + Y_mean
  dcases <- yhat * fd_sd + fd_mean            # back to case-change units, per province
  fut <- which(dates > hist_end)             # project future months only
  proj <- last_obs + t(apply(dcases[, fut, drop = FALSE], 1, cumsum))
  list(proj = proj, dates = dates[fut])
}

# scenario -> array (province x future-month x model), dropping unstable models
project_scenario <- function(scenario) {
  tas <- cmip_tas[[scenario]]
  pr <- cmip_pr[[scenario]]
  nm  <- dim(tas$arr)[3]
  lst <- lapply(seq_len(nm), function(m)
    project_one(tas$arr[, , m], pr$arr[, , m], tas$dates)$proj)
  arr  <- simplify2array(lst)                 # province x month x model
  keep <- apply(arr, 3, function(x) all(is.finite(x)))
  arr[, , keep, drop = FALSE]
}

fut_dates <- project_one(cmip_tas[[1]]$arr[, , 1], cmip_pr[[1]]$arr[, , 1],
                         cmip_tas[[1]]$dates)$dates
proj_arr  <- list("SSP1-2.6" = project_scenario("SSP1-2.6"),
                  "SSP5-8.5" = project_scenario("SSP5-8.5"))
```

**6.1 The projection standardises the future climate using the *historical* mean and
standard deviation, not the future data's own. Why is that essential? What would go
wrong if we standardised the future climate on itself?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 6.1 Suggested answer: click to reveal (try it first)
The model was fitted on climate standardised by its *historical* mean and
standard deviation, so the coefficients only make sense on that same scale. If we
re-standardised the future climate on itself, we would subtract off the very warming
(or drying/wetting) trend we are trying to project; every period would look
"average" relative to its own era, and the projected climate signal would vanish. We
must apply the historical scaling so that a genuinely hotter future shows up as a
large positive standardised value.
:::

Aggregate the projected months to annual totals per province, and plot the ensemble
mean and interquartile band for each scenario.

```{r part6-plot, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
# Turn one scenario's projection (province x month x model) into a per-province,
# per-year summary across models (mean and interquartile band), ready for ggplot.
annual_proj <- function(arr, dates, scenario) {
  yr <- lubridate::year(dates)                             # the calendar year of each month
  years <- sort(unique(yr))
  years <- years[vapply(years, function(y) sum(yr == y) == 12, logical(1))]  # complete (12-month) years only
  # For each year, sum that year's months -> a province x model matrix (annual totals).
  # sapply stacks those matrices as columns, giving a (province*model) x year matrix.
  ap <- sapply(years, function(y)
    apply(arr[, yr == y, , drop = FALSE], c(1, 3), sum))
  # Fold that flat matrix back into a 3-D array. R fills column-major, so the fastest-
  # varying index (province) comes first, then model, then year.
  ap <- array(ap, dim = c(length(provinces), dim(arr)[3], length(years)))  # prov x model x year
  ap <- aperm(ap, c(1, 3, 2))                              # reorder to prov x year x model
  # Collapse across models: mean and the 25th/75th percentiles for each province-year.
  # apply(ap, c(1, 2), .) keeps province & year, summarising over the model dimension;
  # as.vector unrolls it province-fastest, matching the rep() pattern below.
  tibble::tibble(
    province_en = rep(provinces, times = length(years)),  # province cycles fastest
    year = rep(years, each = length(provinces)),          # year repeats per province block
    scenario = scenario,
    mean = as.vector(apply(ap, c(1, 2), mean)),           # ensemble mean line
    q25  = as.vector(apply(ap, c(1, 2), quantile, 0.25)), # lower edge of the band
    q75  = as.vector(apply(ap, c(1, 2), quantile, 0.75))  # upper edge of the band
  )
}

# One tidy data frame with both scenarios stacked.
proj_annual <- bind_rows(
  annual_proj(proj_arr[["SSP1-2.6"]], fut_dates, "SSP1-2.6"),
  annual_proj(proj_arr[["SSP5-8.5"]], fut_dates, "SSP5-8.5")
)

# Historical average annual reported cases per province (mean monthly count x 12),
# a fixed reference line to compare the projection against.
obs_annual <- tibble::tibble(
  province_en = provinces,
  avg = rowMeans(cases_mat) * 12
)

# One small-multiple per province: the shaded band is the inter-model IQR, the line the mean.
print(ggplot(proj_annual, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +  # interquartile band
  geom_line(linewidth = 0.6) +                                          # ensemble mean
  geom_hline(data = obs_annual, aes(yintercept = avg), inherit.aes = FALSE,
             linetype = "dashed", colour = "grey40", linewidth = 0.4) + # historical average
  facet_wrap(~ province_en, scales = "free_y") +   # free y so small-caseload provinces are visible
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +  # blue vs red
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = "projected annual cases",
       title = "Projected annual cases under CMIP6 (statistical/ARX model)",
       subtitle = "Ensemble mean (line) and IQR (band); dashed line is the historical average annual cases. See caveat above.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

::: speaker-note
Core question for the 90-minute run; pause here. The other Part 6 questions are for self-study.
:::

**6.2 Does this projection look realistic to you?** Look at the trajectories against the
historical average (the dashed line, currently squashed near the bottom of each panel).
Is there anything about the numbers you would want to check before believing them?

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 6.2 Suggested answer: click to reveal (look at the plot first)
The projected counts climb far above anything in the historical record, and the free
y-axis hides how far. Comparing the largest projected monthly count to the largest one
ever observed makes it concrete:

```{r part6-sanity, echo=TRUE, eval=TRUE}
vapply(proj_arr, function(a) max(a, na.rm = TRUE), numeric(1)) # largest projected monthly count, per scenario
max(cases_mat)                                       # largest observed monthly count
```

The projected numbers are orders of magnitude larger, and far into the century they can
run past any plausible case count (for scale, Thailand's entire population is about 71.7
million, [World Bank](https://data.worldbank.org/indicator/SP.POP.TOTL?locations=TH){target="_blank"}).
This is not a coding error: the statistical model rebuilds counts by cumulatively summing
predicted monthly *changes*, an unbounded random walk that drifts, and it is being
extrapolated into future temperatures well outside its training range. Read the *shape*
and the *gap* between scenarios, not the absolute counts. The bounded SEIRS in the
optional section is one way to avoid this.
:::

**6.3 Look at the projected trajectories and re-read the caveat. Which features of
these plots would you trust, and which would you not? How would you describe the
result honestly to a decision-maker?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 6.3 Suggested answer: click to reveal (try it first)
Trust the *direction and relative* differences, e.g. that the higher-
emissions scenario diverges from the lower one, and the ranking of provinces, more
than the absolute case numbers, which drift because of the cumulative-sum
reconstruction. Honest framing to a decision-maker: "this shows how a climate-driven
statistical model *responds* to two emissions futures; the shape and the gap between
scenarios are informative, but the absolute counts decades out are illustrative, not
a forecast, and a mechanistic model would be needed for credible long-range numbers."
:::

The band above is the *spread* of model magnitudes (the interquartile range across
the ensemble), which is not the same as the models *agreeing*. Another approach
to considering the impact of the climate model variation is to determine a 
**consensus**: where do the models agree on the sign of the change? Or perhaps 
roughly magnitude as well. This mirrors the model-agreement maps in
[Sexton et al. (2026)](https://doi.org/10.1016/j.ecoinf.2026.103899){target="_blank"},
which colour only the regions where a majority of models agree on the level and
direction of change. Here we compare each province's mean annual cases in the 
first decade of the projection against the last, per climate model, and map the 
fraction of models projecting an increase.

```{r part6-consensus, echo=TRUE, eval=TRUE, fig.width=11, fig.height=6}
# Per-model change in annual cases: first projection decade vs last, per province.
model_change <- function(arr, dates, n_win = 10) {
  yr <- lubridate::year(dates)
  years <- sort(unique(yr))
  years <- years[vapply(years, function(y) sum(yr == y) == 12, logical(1))]  # complete years only
  ap <- sapply(years, function(y)
    apply(arr[, yr == y, , drop = FALSE], c(1, 3), sum))    # province*model collapsed per year
  ap <- array(ap, dim = c(length(provinces), dim(arr)[3], length(years)))  # prov x model x year
  early <- rowMeans(ap[, , seq_len(n_win), drop = FALSE], dims = 2)             # prov x model
  late  <- rowMeans(ap[, , tail(seq_along(years), n_win), drop = FALSE], dims = 2)
  late - early
}

# Fraction of models projecting an increase, per province (0 = all decrease, 1 = all increase).
consensus_df <- function(arr, dates, scenario) {
  chg <- model_change(arr, dates)
  tibble::tibble(
    ADM1_EN = provinces,
    scenario = scenario,
    frac_up = rowMeans(chg > 0)
  )
}

consensus <- bind_rows(
  consensus_df(proj_arr[["SSP1-2.6"]], fut_dates, "SSP1-2.6"),
  consensus_df(proj_arr[["SSP5-8.5"]], fut_dates, "SSP5-8.5"))

# Join to the province polygons and map (grey backdrop = provinces not in the panel).
consensus_sf <- shape_file |>
  select(ADM1_EN) |>
  left_join(consensus, by = "ADM1_EN")

print(ggplot(consensus_sf) +
  geom_sf(data = shape_file, fill = "grey95", colour = "grey80", linewidth = 0.2) +
  geom_sf(aes(fill = frac_up), colour = "grey40", linewidth = 0.2) +
  facet_wrap(~ scenario) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0.5, limits = c(0, 1),
                       breaks = c(0, 0.5, 1),
                       labels = c("all decrease", "split", "all increase"),
                       name = "models projecting\nan increase") +
  labs(title = "Model consensus on the direction of change in annual cases",
       subtitle = paste("Fraction of CMIP6 models projecting more cases (last decade vs first). White = no consensus.",
                        "Read as the agreed direction, not a robust magnitude: the change still inherits the ARX drift.",
                        sep = "\n")) +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()))
```

**6.4 The trajectory band is the interquartile range across models; the consensus
map instead shows the fraction of models agreeing on the *direction* of change. What
does the consensus map tell you that the band does not, and why might agreement on
the sign be more useful to a decision-maker than the ensemble mean?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

::: {.callout-tip collapse="true"}
## 6.4 Suggested answer: click to reveal (try it first)
The band mixes together models that disagree: a wide band could mean the
models point in opposite directions, or that they agree on the direction but differ
in magnitude. The consensus map separates those, showing *where* the models agree on
the sign of the change. Agreement on the direction is usually more decision-relevant
than the ensemble mean, because a robust "cases rise here under both scenarios" is
something you can act on even when the absolute numbers are uncertain, whereas an
ensemble mean can sit in the middle of a bimodal, no-consensus spread and imply a
false precision. The map also shows that agreement is not spatially uniform: some
provinces are far more consistent than others.
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>

---

# Optional: a mechanistic SEIRS alternative

::: speaker-note
This whole section is optional. 

It reuses the case and climate data already built above (no re-loading), fits a 
compartmental transmission model, and doubles as a **code base for fitting a 
transmission model directly**. 

The fit runs for `maxit = 200` iterations with a relaxed convergence tolerance
(`factr`), which settles this weakly-identified model; reduce `maxit` for a quicker,
rougher live-demo fit.
:::

::: {.callout-note collapse="true"}
## Why an SEIRS model? (optional; click to expand)

The model in Parts 5–6 is deliberately the simplest thing that works: one shared
statistical relationship between climate and the month-to-month *change* in 
malaria cases. 

This takes a different, mechanistic approach: rather than regressing malaria 
cases on climate, it fits a compartmental **SEIRS** model (Susceptible → Exposed
→ Infected → Recovered → Susceptible) per province, *sharing* the disease's 
natural-history parameters across all provinces but letting the transmission 
rate vary by province and over time:
$$\begin{aligned}
\frac{dS}{dt} &= -\beta(t) S I + \omega R \\
\frac{dE}{dt} &= \beta(t) S I - \sigma E \\
\frac{dI}{dt} &= \sigma E - \gamma I \\
\frac{dR}{dt} &= \gamma I - \omega R
\end{aligned}$$
where\
* $\sigma$ (1/incubation period), $\gamma$ (1/infectious period) and $\omega$ 
(1/duration of immunity) are **shared across every province** -- these are 
properties of the parasite/host biology, not the location.\
* $\beta(t)$, the transmission rate, is **province-specific and time-varying**: 
$\beta_p(t) = \beta_{0,p} \exp(b_{temp} \cdot z_{temp}(p,t) + b_{rain} \cdot z_{rain}(p,t))$, 
where $z_{temp}$/$z_{rain}$ are standardised monthly temperature/rainfall. This 
is a deliberately simple proxy: malaria is mosquito-borne and mosquito abundance
changs with temperature and breeding site availability ($\approx$rainfall), but 
we didn't find mosquito data for the same time period, so climate stands in 
directly for the missing vector population dynamics. $b_{temp}$ and $b_{rain}$ 
are shared across provinces (one climate-sensitivityrelationship, applied 
everywhere); $\beta_{0,p}$ is a free per-province baseline.\
* Reported monthly cases are modelled as 
$\text{cases}_p(t) \sim \text{Poisson}(\text{scale}_p \cdot y_p(t))$, where
$y_p(t) = \int_{\text{month } t} \sigma E_p(\tau)\,\mathrm{d}\tau$ is the new 
infections *accumulated over* month $t$ (the simulation integrates $\sigma E$ 
within each month and resets the accumulator at the start of each month), fit by 
maximum likelihood (L-BFGS-B on the negative log-likelihood, parameters in 
log-space to keep them positive).

**7.1 Why does the transmission rate use an *exponential* of the climate variables,
$\beta_0\exp(b_{temp}z_{temp} + b_{rain}z_{rain})$, rather than a quadratic like the
ARX model in Part 5?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

:::: {.callout-tip collapse="true"}
## 7.1 Suggested answer: click to reveal (try it first)
Two reasons. First, $\beta$ must stay positive; $\exp(\cdot)$ is
positive for any coefficients, whereas a bare quadratic can go negative (the ARX
quadratic is fine because it acts on the standardised *change in cases*, which 
can be either sign). Second, in the ARX model the quadratic is the *only* source 
of nonlinearity, so it has to bend the response itself; the SEIRS already 
produces non-monotonic case dynamics from its structure (susceptible depletion, 
waning immunity, the incubation delay), so a simple positive, multiplicative 
modulation of the rate is probably enough. You could still add a temperature 
optimum by putting a quadratic *inside* the exponent (a Gaussian tolerance 
curve), but two coefficients is about all this short record can identify.
::::


::: {.callout-warning}
## Important limitations, to be read before trusting any number below

Treat every output as illustrative of method, not as an estimate, because:\
* The climate covariates stand in for mosquito data.\
* There are no population denominators used, so `scale` absorbs both the known
human population size and the unknown case ascertainment rate.\
* The data for two parasite species, _Plasmodium vivax_ and _falciparum_, are 
pooled, even though they have different incubation periods, relapse behaviour, etc.\
* Partly due to the pooling, the natural-history parameters ($\sigma,\gamma,\omega$)
are estimated from the data rather than fixed to published vivax/falciparum biology.
They are still held to a biologically plausible range by the optimiser bounds (see
Step 6), but the data are what pick the value within that range, rather than the
literature.\
* The optimiser uses a small iteration budget so it runs quickly; it is **not** 
run to convergence. 

:::

**The data.** The SEIRS reuses the case matrix and per-province climate already built
in Parts 2–3 and 5. The only difference is that it standardises climate **globally**
(one shared scale) rather than per province, so that `b_temp`/`b_rain` are a single
comparable climate sensitivity.

```{r seirs-setup, echo=TRUE, eval=TRUE}
# reuse cases_mat, clim_temp (Kelvin), clim_rain (m), popsd() from Part 5
ztemp <- (clim_temp - mean(clim_temp)) / popsd(clim_temp)
zrain <- (clim_rain - mean(clim_rain)) / popsd(clim_rain)
cat(sprintf("SEIRS inputs: %d provinces x %d months, %.0f total cases\n",
            nrow(cases_mat), ncol(cases_mat), sum(cases_mat)))
```

The model is built up in five chunks below, then fitted. 

**Step 1: the transmission dynamics (the ODE right-hand side).** This is the SEIRS
model itself: the instantaneous rate of change of each compartment. `state` is a
5 × n matrix, one column per province, rows S, E, I, R, and C (a running count of
new infections), so everything is vectorised over provinces at once.

```{r seirs-rhs, echo=TRUE, eval=TRUE}
SUBSTEPS_PER_MONTH <- 20   # RK4 sub-steps per month

seirs_rhs <- function(state, beta, sigma, gamma, omega) {
  S <- state[1, ]
  E <- state[2, ]
  I <- state[3, ]
  R <- state[4, ]
  rbind(-beta * S * I + omega * R,     #  dS  (lose to infection, gain from waning immunity)
         beta * S * I - sigma * E,     #  dE  (newly infected, then become infectious)
         sigma * E   - gamma * I,      #  dI  (become infectious, then recover)
         gamma * I   - omega * R,      #  dR  (recover, then lose immunity)
         sigma * E)                    #  dC  (cumulative NEW infectious -- those more likely to be detectable)
}
```

**Step 2: simulating it forward (fixed-step RK4).** We integrate the ODE month by
month using a fixed-step Runge–Kutta (RK4) scheme with `SUBSTEPS_PER_MONTH` steps per
month, for all provinces simultaneously. A burn-in lets the model settle before we
record anything, and we reset the accumulator `C` at the start of each month so that
`incidence` holds the new infections *within* that month.

A proper ODE solver would be more accurate but slower. 

```{r seirs-sim, echo=TRUE, eval=TRUE}
simulate_incidence <- function(beta, sigma, gamma, omega,
                               i0 = 1e-4, e0 = 1e-4, burn_in_months = 12,
                               init_state = NULL, return_state = FALSE) {
  n_prov <- nrow(beta); n_months <- ncol(beta)
  dt <- 1 / SUBSTEPS_PER_MONTH
  # Initial conditions. If init_state is supplied (used by the projection to continue from
  # the disease state at the end of the historical fit) we start from it and skip the
  # burn-in. Otherwise we start from a near-fully-susceptible population and burn in to
  # settle; this default is what the fit uses and is smooth in the parameters, so the
  # optimiser behaves. return_state = TRUE also returns the final state, so the fit can hand
  # its end state to the projection.
  if (!is.null(init_state)) {
    state <- init_state
    burn_in_months <- 0
  } else {
    state <- rbind(rep(1 - i0 - e0, n_prov), rep(e0, n_prov),
                   rep(i0, n_prov), rep(0, n_prov), rep(0, n_prov))
  }
  step <- function(state, beta_m) {                    # one RK4 sub-step
    k1 <- seirs_rhs(state,              beta_m, sigma, gamma, omega)
    k2 <- seirs_rhs(state + 0.5*dt*k1,  beta_m, sigma, gamma, omega)
    k3 <- seirs_rhs(state + 0.5*dt*k2,  beta_m, sigma, gamma, omega)
    k4 <- seirs_rhs(state + dt*k3,      beta_m, sigma, gamma, omega)
    state + (dt/6) * (k1 + 2*k2 + 2*k3 + k4)
  }
  if (burn_in_months > 0) {                            # settle using month-1 transmission
    beta0_m <- beta[, 1]
    for (b in seq_len(burn_in_months)) {
      for (s in seq_len(SUBSTEPS_PER_MONTH)) state <- step(state, beta0_m)
      state[5, ] <- 0
    }
  }
  incidence <- matrix(0, n_prov, n_months)
  for (m in seq_len(n_months)) {
    state[5, ] <- 0
    beta_m <- beta[, m]                                # use this month's beta
    for (s in seq_len(SUBSTEPS_PER_MONTH)) state <- step(state, beta_m)
    incidence[, m] <- state[5, ]
    state[1:4, ] <- pmin(pmax(state[1:4, ], 0), 1)     # keep compartments in [0, 1]; shouldn't need this but since manually RK4, good to check
  }
  if (return_state) list(incidence = incidence, state = state) else incidence
}
```

**Step 3: unpacking the parameters.** The optimiser works in an unconstrained,
log-transformed space so that rates stay positive. This turns the flat parameter
vector back into named, natural-scale quantities: the shared natural-history rates,
the two climate sensitivities, and the per-province baseline transmission `beta0` and
reporting `scale`.

```{r seirs-unpack, echo=TRUE, eval=TRUE}
unpack_params <- function(params, n) {
  list(sigma = exp(params[1]), gamma = exp(params[2]), omega = exp(params[3]),
       b_temp = params[4], b_rain = params[5],
       beta0 = exp(params[6:(5 + n)]), scale = exp(params[(6 + n):(5 + 2*n)]))
}
```

**Step 4: the likelihood.** Given a parameter vector, build the climate-forced
transmission rate `beta_t`, simulate the incidence, and score it against the observed
counts with a Poisson negative log-likelihood, the quantity the optimiser minimises.

```{r seirs-nll, echo=TRUE, eval=TRUE}
neg_log_lik <- function(params, cases, ztemp, zrain) {
  n <- nrow(cases)
  p <- unpack_params(params, n)
  beta_t <- p$beta0 * exp(p$b_temp * ztemp + p$b_rain * zrain)   # climate forces transmission
  beta_t <- pmin(pmax(beta_t, 1e-6), 50)
  incidence <- simulate_incidence(beta_t, p$sigma, p$gamma, p$omega)
  lam <- pmax(p$scale * incidence, 1e-8)                         # expected reported cases
  sum(lam - cases * log(lam))   # Poisson NLL, additive constant log(cases!) omitted
}
```

**Step 5: starting values.** An initial guess for the optimiser: incubation 
~0.5 months (includes the mosquito transmission cycle), infectious ~1 month, 
immunity ~6 months, no climate effect to begin with, and per-province baselines 
scaled to each province's mean caseload.

```{r seirs-x0, echo=TRUE, eval=TRUE}
default_x0 <- function(cases) {
  n <- nrow(cases)
  c(log(1/0.5), log(1/1.0), log(1/6.0), 0, 0,      # sigma, gamma, omega, b_temp, b_rain
    log(rep(2.5, n)),                              # beta0 (per province)
    log(pmax(rowMeans(cases), 1) * 50))            # scale (per province)
}
```

**Step 6: fit.** Now minimise the negative log-likelihood with a bounded optimiser
(`L-BFGS-B`). The `lower`/`upper` vectors keep every parameter in a biologically
plausible range (incubation/infectious periods between ~0.05 and 3 months, immunity
between 0.5 and 36 months, the climate sensitivities within ±2). It runs for `maxit = 200`
iterations with a relaxed convergence tolerance (`factr`), which is enough for this
weakly-identified model to settle; reduce `maxit` for a quicker, rougher live-demo fit.

```{r seirs-fit, echo=TRUE, eval=TRUE}
n <- nrow(cases_mat)
x0 <- default_x0(cases_mat)
# Parameter order (matches unpack_params): sigma, gamma, omega, b_temp, b_rain,
# then beta0 (one per province), then scale (one per province). Rates are in log-space.
lower <- c(log(1/3.0),  log(1/3.0),  log(1/36.0),  -2, -2,   # sigma, gamma, omega, b_temp, b_rain
           rep(log(0.05), n), rep(log(1.0),  n))             # beta0 (per province), scale (per province)
upper <- c(log(1/0.05), log(1/0.05), log(1/0.5),   2,  2,    # sigma, gamma, omega, b_temp, b_rain
           rep(log(30),   n), rep(log(1e9),  n))             # beta0 (per province), scale (per province)

# 200 iterations with a relaxed convergence tolerance (factr): the beta0/scale ridge is
# flat, so the default tight tolerance never triggers and the fit would otherwise just run
# to the iteration cap (reporting "code 1") despite having effectively settled. ndeps
# matches SciPy's default gradient step. factr = 1e9 matches SciPy's ftol = 1e9 * eps.
fit <- optim(x0, neg_log_lik, method = "L-BFGS-B",
             lower = lower, upper = upper,
             cases = cases_mat, ztemp = ztemp, zrain = zrain,
             control = list(maxit = 200, factr = 1e9, ndeps = rep(1e-8, length(x0))))
p <- unpack_params(fit$par, n)

cat(sprintf("convergence code=%d  nll=%.1f\n", fit$convergence, fit$value))  # 0 converged; 1 did not
cat(sprintf("incubation period 1/sigma = %.2f months\n", 1/p$sigma))
cat(sprintf("infectious period 1/gamma = %.2f months\n", 1/p$gamma))
cat(sprintf("immunity duration 1/omega = %.2f months\n", 1/p$omega))
cat(sprintf("climate sensitivity b_temp = %+.3f  b_rain = %+.3f\n", p$b_temp, p$b_rain))
```

**Step 7: diagnostics.** A snapshot of the fitted per-province parameters, with the
observed (reported) and model-fitted total cases side by side, sorted by observed
caseload. `R0_baseline` is the baseline reproduction number $\beta_{0}/\gamma$
(transmission rate over recovery rate, at full susceptibility).

```{r seirs-params, echo=TRUE, eval=TRUE}
# Fitted expected monthly cases (the Poisson mean), reused by the plot below. We also keep
# the disease state at the end of the historical period, to seed the projection (below).
beta_fit      <- p$beta0 * exp(p$b_temp * ztemp + p$b_rain * zrain)
beta_fit      <- pmin(pmax(beta_fit, 1e-6), 50)   # same clipping as the likelihood, so the diagnostic matches the fitted model
sim_hist      <- simulate_incidence(beta_fit, p$sigma, p$gamma, p$omega, return_state = TRUE)
lam_fit       <- p$scale * sim_hist$incidence
fit_end_state <- sim_hist$state   # S/E/I/R/C at the last historical month; the projection continues from here

# Per-province fit snapshot (mirrors the Python param_summary table).
param_summary <- tibble::tibble(
  province_en    = rownames(cases_mat),
  beta0          = p$beta0,
  R0_baseline    = p$beta0 / p$gamma,   # gamma is shared across provinces
  scale          = p$scale,
  reported_cases = rowSums(cases_mat),  # observed total
  fitted_cases   = rowSums(lam_fit)     # model total (expected)
) %>% arrange(desc(reported_cases))

knitr::kable(param_summary, digits = c(NA, 3, 2, 1, 0, 0))
```

Continuing the diagnostics, compare the fitted incidence (`lam_fit`, computed above)
against the observed monthly counts, province by province.

```{r seirs-diag, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
ss_res <- sum((cases_mat - lam_fit)^2)
ss_tot <- sum((cases_mat - mean(cases_mat))^2)
cat(sprintf("pseudo-R2 (vs fitted lambda): %.3f\n", 1 - ss_res / ss_tot))

diag_long <- bind_rows(
  matrix_to_long(cases_mat, "observed"),
  matrix_to_long(lam_fit, "fitted")
)

print(ggplot(diag_long, aes(month, cases, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c(observed = "black", fitted = "red")) +
  labs(x = "date", y = "monthly cases",
       title = "Observed vs SEIRS-fitted monthly cases") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

```{r seirs-export, echo=FALSE, message=FALSE}
# Save the fitted SEIRS outputs for the direct comparison (python_workflow/compare.R).
.odir <- file.path("..", "outputs")
dir.create(.odir, showWarnings = FALSE, recursive = TRUE)
jsonlite::write_json(list(
  selected = provinces, months = as.character(months),
  nll = fit$value,
  sigma = p$sigma, gamma = p$gamma, omega = p$omega,
  b_temp = p$b_temp, b_rain = p$b_rain,
  beta0 = p$beta0, scale = p$scale,
  lam_fit = lam_fit
), file.path(.odir, "r_seirs.json"), matrix = "rowmajor", digits = NA, auto_unbox = TRUE)
```

::: {.callout-note}
## A note on identifiability
`beta0` (baseline transmission) and `scale` (population × reporting rate) trade 
off to set the case *level*, so the optimiser can shuffle magnitude between 
them, and the natural-history rates ($\sigma,\gamma,\omega$) are only weakly 
constrained by the data. 

This is very clearly shown when comparing the python and R outputs.
:::

**Step 8: Projecting forward, and model consensus.** The mechanistic model can
also be projected forward under CMIP6, exactly as the statistical model was in
Part 6. For each climate model we *continue* the simulation from the disease state
at the end of the historical fit and run the future months only, seeded that way so
there is no artificial restart or initial-epidemic transient (this is the same idea
as the Part 6 ARX projection, which continues from the last observed month). The
future climate is standardised with the *same global historical mean and standard
deviation* the model was fitted with. We plot the projected annual cases (ensemble
mean and interquartile band, as in Part 6).

```{r seirs-project, echo=TRUE, eval=TRUE, fig.width=11, fig.height=10}
# Global (shared) historical scale the SEIRS was fitted with; see seirs-setup.
tmean <- mean(clim_temp)
tsd <- popsd(clim_temp)   # Kelvin
rmean <- mean(clim_rain)
rsd <- popsd(clim_rain)   # m/month

# Project the fitted SEIRS under one climate model's future climate, continuing from the
# disease state at the end of the historical fit and projecting future months only (as the
# Part 6 ARX projection continues from the last observed month).
seirs_project_one <- function(tas_pm, pr_pm, dates) {
  zt <- (tas_pm - tmean) / tsd                       # both in °C (see tmean above), as the Python does
  zr <- (pr_pm - rmean) / rsd
  beta_t <- p$beta0 * exp(p$b_temp * zt + p$b_rain * zr)
  beta_t <- pmin(pmax(beta_t, 1e-6), 50)             # same clipping as the likelihood
  fut <- which(dates > hist_end)                     # future months only
  p$scale * simulate_incidence(beta_t[, fut, drop = FALSE], p$sigma, p$gamma, p$omega,
                               init_state = fit_end_state)   # seed from the fit's end state; no burn-in
}

# scenario -> province x future-month x model of projected reported cases
# (exclude climate models that produced unstable future estimates)
seirs_project_scenario <- function(scenario) {
  tas <- cmip_tas[[scenario]]
  pr <- cmip_pr[[scenario]]
  lst <- lapply(seq_len(dim(tas$arr)[3]),
                function(m) seirs_project_one(tas$arr[, , m], pr$arr[, , m], tas$dates))
  arr  <- simplify2array(lst)
  keep <- apply(arr, 3, function(x) all(is.finite(x)))
  arr[, , keep, drop = FALSE]
}

seirs_arr   <- list("SSP1-2.6" = seirs_project_scenario("SSP1-2.6"),
                    "SSP5-8.5" = seirs_project_scenario("SSP5-8.5"))
seirs_dates <- { d <- cmip_tas[[1]]$dates; d[d > hist_end] }   # future months only

# Projected annual cases per province: ensemble mean and interquartile band (reuses annual_proj).
seirs_annual <- bind_rows(
  annual_proj(seirs_arr[["SSP1-2.6"]], seirs_dates, "SSP1-2.6"),
  annual_proj(seirs_arr[["SSP5-8.5"]], seirs_dates, "SSP5-8.5"))

print(ggplot(seirs_annual, aes(year, mean, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.6) +
  geom_hline(data = obs_annual, aes(yintercept = avg), inherit.aes = FALSE,
             linetype = "dashed", colour = "grey40", linewidth = 0.4) + # historical average (from Part 6)
  facet_wrap(~ province_en, scales = "free_y") +
  scale_colour_manual(values = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  scale_fill_manual(values   = c("SSP1-2.6" = "#2166ac", "SSP5-8.5" = "#b2182b")) +
  labs(x = "year", y = "projected annual cases",
       title = "Projected annual cases under CMIP6 (SEIRS model)",
       subtitle = "Ensemble mean (line) and IQR (band); dashed line is the historical average annual cases.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)))
```

**7.2 Compare with the statistical projection.** Put these SEIRS trajectories next to the
ARX ones in Part 6. How do the projected *levels* differ over the century, and what in
each model's construction explains it?

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

:::: {.callout-tip collapse="true"}
## 7.2 Suggested answer: click to reveal (try it first)
The ARX trajectories drift upward without limit (some far past any plausible case count),
while the SEIRS ones stay in a realistic range and flatten or oscillate. The reason is
structural. The ARX model rebuilds counts by cumulatively summing predicted monthly
*changes*, so nothing stops the level wandering, and extrapolating its quadratic climate
response into unseen future temperatures gives it a persistent drift. The SEIRS instead
tracks population *fractions* (S + E + I + R = 1) and converts the infected fraction into
cases through a fitted per-province `scale`, so its output is bounded by that scale
(calibrated to the observed caseload), while saturation and waning immunity keep the
dynamics finite. That boundedness comes from the fitted `scale`, not from plugging in a
real population, so the SEIRS gives sensible-looking levels but not a true attack rate.
For long-range projection the bounded model is the safer choice; the ARX is really
illustrative only, but shows a consistent direction of change and gap between
SSP scenarios.
::::

Now we map the **consensus**: the fraction of models in which a province's 
annual cases change in similar direction and quantity from the first to the last 
decade.The map reuses the helpers from Part 6.
```{r seirs-consensus, echo=TRUE, eval=TRUE, fig.width=11, fig.height=6}
# Reuse the Part 6 consensus helpers (model_change, consensus_df) on the SEIRS projection.
seirs_consensus <- bind_rows(
  consensus_df(seirs_arr[["SSP1-2.6"]], seirs_dates, "SSP1-2.6"),
  consensus_df(seirs_arr[["SSP5-8.5"]], seirs_dates, "SSP5-8.5"))
seirs_consensus_sf <- shape_file |>
  select(ADM1_EN) |>
  left_join(seirs_consensus, by = "ADM1_EN")

print(ggplot(seirs_consensus_sf) +
  geom_sf(data = shape_file, fill = "grey95", colour = "grey80", linewidth = 0.2) +
  geom_sf(aes(fill = frac_up), colour = "grey40", linewidth = 0.2) +
  facet_wrap(~ scenario) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0.5, limits = c(0, 1),
                       breaks = c(0, 0.5, 1),
                       labels = c("all decrease", "split", "all increase"),
                       name = "models projecting\nan increase") +
  labs(title = "SEIRS model consensus on the direction of change in annual cases",
       subtitle = "Fraction of CMIP6 models projecting more cases (last decade vs first). White = no consensus.") +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()))
```

**7.3 Discussion (model structure): compare this SEIRS consensus map with the
statistical one in Part 6. Both are now seeded from the end of the fit and cover the
future only, so they are on the same footing. Where do the two models agree on the
direction of change, and where do they differ? What does any disagreement tell you
about where the uncertainty comes from?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

:::: {.callout-tip collapse="true"}
## 7.3 Suggested answer: click to reveal (try it first)
Because both models now start from the same fitted end state and run over the same
future window, differences between the two maps reflect the models' *structure*, not an
artefact of how each projection was set up. Where the climate signal is strong it tends
to dominate both models, so they agree on the direction of change and that agreement can
be reported with more confidence. Where the signal is weak, the direction in each
province is only lightly set by climate and is instead decided by each model's structure,
so any disagreement there is *structural* (model-form) uncertainty.

The two structures still differ in ways that matter. The statistical model is quadratic
in temperature and rainfall, so a province sitting near the fitted optimum can fall
either side of the hump with a tiny climate nudge, and it rebuilds counts by cumulatively
summing predicted changes, an unbounded random walk that drifts (this is why its
trajectories can grow implausibly large, even past the population). The SEIRS forces
transmission monotonically and is bounded by saturation and waning immunity, so its levels
stay finite.

The lesson is that there are two different uncertainties in play. The consensus map shows
disagreement *across climate models*; the gap between the two maps shows disagreement
*across disease-model structure* (structural or model-form uncertainty). Reporting the
direction of change where the models agree, and preferring a bounded model for long-range
work, is the honest way to handle both.
::::

**7.4 Discussion (SEIRS): the statistical model (Parts 5–6) and this mechanistic model
both link climate to malaria. What can the mechanistic model tell you that the
statistical one cannot, and vice versa? **

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

:::: {.callout-tip collapse="true"}
## 7.4 Suggested answer: click to reveal (try it first)
The mechanistic SEIRS gives interpretable epidemiological quantities: a transmission
rate, a baseline reproduction number, and incubation/infectious/immunity durations,
which map onto biology, can be compared with the literature, and correspond to things
an intervention could change. Because it represents the process (susceptible depletion,
waning immunity), it captures saturation and threshold behaviour, stays bounded rather
than drifting, and can answer counterfactual "what if transmission were reduced"
questions the statistical model cannot. That makes it the more defensible tool for
long-range projection and for reasoning about interventions.

The statistical (ARX) model, in return, commits to no mechanism, so it cannot be wrong
about one: it estimates the empirical climate-to-cases association directly. It is
simpler, faster, has fewer parameters to identify, and is transparent about being
purely data-driven, which is a strength when the mechanism is uncertain or the goal is
just to detect and quantify a climate signal. Its cost is that it offers no biological
interpretation and, as Part 6 showed, extrapolates poorly. In short: the statistical
model is a good *description*, the mechanistic model a (contestable) *explanation*.
::::

**7.5 Ways to improve it: the SEIRS works in population *fractions* and folds the true
population into the fitted `scale`. What would you gain by bringing in actual
per-province human population numbers, and what new problem appears if you need them
out to 2100?**

```{=html}
<textarea cols="127" rows="4"></textarea>
```
<p><br></p>

:::: {.callout-tip collapse="true"}
## 7.5 Suggested answer: click to reveal (try it first)
Using real per-province populations as the denominator would separate the two things
`scale` currently confounds: population size and reporting rate. You would recover an
interpretable *attack rate* (cases per head), the compartments would be bounded by a
real number of people rather than a fitted constant, and provinces would be comparable
on a genuine per-capita basis. The catch is that a projection to 2100 needs *future*
population, not today's, so you would have to bring in population projections (from a
demographic model or a national statistics office), and those carry their own
uncertainty (fertility, mortality, migration, urbanisation) that compounds with the
climate-model uncertainty already in the projection. Other improvements worth listing:
fixing the natural-history rates to published *vivax*/*falciparum* biology rather than
fitting them, separating the two species, and adding an explicit mosquito/vector layer
so climate acts through abundance rather than directly on transmission.
::::
:::

<p><input type="button" value="Print this page" onClick="window.print()"></p>
